---
title: "01. Tutorial to generate comprehensive heatmaps to show results on all combinations"
output: html_document
date: "2025-09-23"
author: "Edoardo Giuili"
---

## Introduction

This is the first notebook to reproduce the analysis of the bench.resmarking results. More specifically, in this notebook we will go over the results obtained on the in-silico mixtures. In total, 1640 WGBS, 1550 and 440 samples (WGBS, RRBS and MNase-cfRRBS, respectively) have been deconvolved. These samples were generated at different sequencing depths, spanning 10 expected tumor fractions, different tumor types and using 3 DMR selection tools. We will therefore generate plots for all these combinations, step by step.

### Benchmarking metrics

It is important to highlight that to evaluate the performance of the deconvolution tools we considered 3 different key metrics: the root-mean-squared error (RMSE), the area under the curve (AUC-ROC) and the Spearman's rank correlation coefficient (ρ {\displaystyle \rho }).

### Setting up the environment

First, set up the main directory

```{r setup}
# Don't change this if you git cloned the entire GitHub repository and you are working in the notebooks/ folder
knitr::opts_knit$set(root.dir = './')
```

Import the necessary libraries for the analysis

```{r message=FALSE, warning=FALSE}
library(ggpubr)
library(dplyr)
library(tidyverse)
library(ComplexHeatmap)
library(stringr)
library(pROC)
```

Import the useful functions and the metrics

```{r}
source("../R/compute_auc.R")

# RMSE
rmse <- function(actual, predicted) {
  round(sqrt(mean((actual - predicted)^2)),4)
}
# Spearman's rank correlation coefficient (SCC)
scc <- function(actual, predicted) {
  cor(actual, predicted, method = "spearman", use = "pairwise.complete.obs")
}
```

Now, we will import all the results generated and described above. Note that here we expect that you have git cloned the repository and thus have the expected folder structure where in the folder resources/ there are the results files.

```{r}
df_raw <- read.csv("../resources/final_benchmarking_dataset.csv")
bench.res <- subset(df_raw, df_raw$mixture_type == "in_silico" & !(df_raw$deconv_tool %in% c("MeDeCom","RefFreeCellMix")))
```

### Missing values

An important aspect to take into consideration is the number of missing values per tool.

```{r}
missing_percentage <- bench.res %>%
     group_by(deconv_tool) %>%
     summarise(
         total = n(),
         missing = sum(is.na(predicted_tf)),
         percent_missing = 100 * missing / total
     )
missing_percentage
```

Since the presence of missing values is very low (meaning that all the tools almost always produce an output), we will just remove them without penalizing the tools that produced missing values or imputing them.

```{r}
# Remove rows with NA values for further analysis 
bench.res <- bench.res %>%
  filter(!is.na(predicted_tf))
```

## Plot the results

### Heatmap of all combinations

As a first step in the analysis, we think that a comprehensive heatamap of all possible combinations might be useful as a starting point. We will first compute the metrics necessary for the heatmap: being RMSE, AUC-ROC value and SCC. Since the RMSE is not a scale-free metric, it will increase at increasing observed tumoral fraction. This determines that the RMSE at high tumoral fraction (e.g. 0.25 and 0.5) will dominate the overall and advantage those tools that perform well uniquely at high tumoral fractions. To account for this bias, in the final metrics used to rank the tools we have decided to compute the RMSE separately at each observed tumoral fraction within each combination, then min-max scale it across deconvolution tools and aggregate the individual min-max scaled RMSE at each tumoral fraction using the mean to get a unique RMSE per combination. However, here we want to see how the metrics change across the different combinations, so we need to visualize the raw RMSE. For this reason, we will plot 2 separate heatmaps: 1 heatmap summarizing the raw AUC-ROC value and SCC across the different tumor types and sequencing depths; and 1 heatmap summarizing the raw RMSE across the different tumor types, sequencing depths and expected tumor fractions.

```{r}
rmse_raw <- bench.res %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
  summarize("1-RMSE" = 1-rmse(expected_tf, predicted_tf))
```

Then we compute the AUC for each tool. To compute it, we will loop over each combination and calculate the AUC-ROC score at each tumor fraction. Then, to get an estimation of the AUC-ROC score per combination, we will average the individual AUC-ROC scores per fraction.

```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(bench.res$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(bench.res$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(bench.res, bench.res$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)

          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_raw <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(AUC=mean(AUC),.groups = "drop")
```

Finally, we compute the SCC. The SCC is computed for each combination pooling all the observed tumor fractions together to get a value for each combination. 

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_raw <- bench.res %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop")
```

We can now combine the AUC-ROC and SCC metrics into a unique dataframe, while keeping the RMSE separated.

```{r}
df.metrics <- scc_raw %>%
  left_join(auc_raw)
```

### Plot the data of WGBS

#### AUC-ROC and SCC

We are now ready to plot an comprehensive heatmap of the WGBS dataset describing all the different combinations. We will first subset the metrics dataframe and retain only the tumor samples belonging to the WGBS dataset.

```{r}
final_wgbs_df <- df.metrics %>% filter(tumor_type %in% c("LUAD","LUSC","BRCA","BLCA"))
```

Change the format of the dataframe to a long format

```{r}
# Create 2 columns corresponding to the deconvolution tool names collapsed with the sequencing depths and dmr tools
final_wgbs_df_mutated <- final_wgbs_df %>%
  mutate(tool_dmr_combination = paste(deconv_tool, dmr_tool, sep = "_")) %>%
  mutate(tumor_seq_combination = paste(tumor_type, seq_depth, sep = "_"))

# Make a long format of the df
final_wgbs_df_long <- final_wgbs_df_mutated %>% 
  pivot_longer(
    cols = c(AUC,SCC),
    names_to = "metric",
    values_to = "value"
  ) %>%
  group_by(tool_dmr_combination, metric, tumor_seq_combination) %>%
  summarise(value = value, .groups = "drop")
```

Now, per metric, select only the column corresponding to it, go back to the wider format and transform it to a matrix.

```{r}
# Specify the order of the columns for visualization purposes
desired_order <- c("LUAD_62M", "LUAD_183M", "LUAD_310M", "LUAD_620M",
                   "LUSC_62M", "LUSC_183M", "LUSC_310M", "LUSC_620M",
                   "BRCA_62M", "BRCA_183M", "BRCA_310M", "BRCA_620M",
                   "BLCA_62M", "BLCA_183M", "BLCA_310M", "BLCA_620M")

# AUC matrix
wgbs_auc_mat <- final_wgbs_df_long %>%
  filter(metric == "AUC") %>%
  select(tool_dmr_combination, tumor_seq_combination, value) %>%
  pivot_wider(names_from = tumor_seq_combination, values_from = value) %>%
  column_to_rownames("tool_dmr_combination") %>%
  as.matrix()
wgbs_auc_mat <- wgbs_auc_mat[, desired_order]

# SCC matrix
wgbs_scc_mat <- final_wgbs_df_long %>%
  filter(metric == "SCC") %>%
  select(tool_dmr_combination, tumor_seq_combination, value) %>%
  pivot_wider(names_from = tumor_seq_combination, values_from = value) %>%
  column_to_rownames("tool_dmr_combination") %>%
  as.matrix()
wgbs_scc_mat <- wgbs_scc_mat[, desired_order]
```

Order the rows based on the Score values to give kind of a ranking

```{r}
# Compute row means of the Score matrix
score_means <- rowMeans(wgbs_auc_mat) + rowMeans(wgbs_scc_mat)

# Order the row names based on descending Score (you can use decreasing = FALSE for ascending)
ordered_rows <- names(sort(score_means, decreasing = TRUE))

# Reorder all matrices by the ordered row names
wgbs_auc_mat   <- wgbs_auc_mat[ordered_rows, ]
wgbs_scc_mat   <- wgbs_scc_mat[ordered_rows, ]
```

And now we can plot the results in a heatmap using ComplexHeatmap package.

```{r}
# Define tumor types and sequencing depths for the annotations
seq_depth <- rep(c("62M", "183M", "310M", "620M"), 4)
tumors    <- rep(c("LUAD", "LUSC", "BRCA", "BLCA"), each = 4)

# Assign palettes
col_anno <- HeatmapAnnotation(
  SeqDepth = seq_depth,
  Tumor    = tumors,
  col = list(
    SeqDepth = c("62M"="#FAEBDDFF","183M"="#F69C73FF","310M"="#A11A5BFF","620M"="#03051AFF"),
    Tumor = c("LUAD"="#ff7f0e", "BRCA"="#2ca02c",
              "LUSC"="#9467bd", "BLCA"="#8c564b")
  ),
  show_annotation_name = FALSE
)

# Define heatmap settings shared across all plots
common_opts <- list(
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  column_names_rot = 45,
  show_row_dend = TRUE,
  show_column_names = FALSE,
  show_column_dend = FALSE,
  top_annotation = col_anno,
  column_labels = tumors
)

# Put all matrices and their titles/names in a list
heatmap_data <- list(
  AUC         = wgbs_auc_mat,
  SCC         = wgbs_scc_mat
)

# Generate heatmaps with do.call
heatmaps <- lapply(names(heatmap_data), function(name) {
  do.call(Heatmap, c(
    list(
      matrix = heatmap_data[[name]],
      name = name,
      column_title = name
    ),
    common_opts
  ))
})
draw(Reduce(`+`, heatmaps))
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/01_comprehensive_heatmaps"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
pdf(paste0(output_dir,"/wgbs_scc_auc_heatmap.pdf"), width = 12, height = 6)
draw(Reduce(`+`, heatmaps))
dev.off()
```


#### RMSE

We are now ready to plot an comprehensive heatmap of the WGBS dataset describing all the different combinations for the RMSE metric, similarly to what we have done above with the addition of the expected fraction. We will first subset the metrics dataframe and retain only the tumor samples belonging to the WGBS dataset.

```{r}
final_wgbs_rmse_df <- rmse_raw %>% filter(tumor_type %in% c("LUAD","LUSC","BRCA","BLCA"))
```

Change the format of the dataframe to a long format

```{r}
# Create 2 columns corresponding to the deconvolution tool names collapsed with the sequencing depths and dmr tools
final_wgbs_rmse_df_mutated <- final_wgbs_rmse_df %>%
  mutate(tool_dmr_combination = paste(deconv_tool, dmr_tool, sep = "_")) %>%
  mutate(tumor_seq_tf_combination = paste(tumor_type, seq_depth, expected_tf, sep = "_"))

# Make a long format of the df
final_wgbs_rmse_df_long <- final_wgbs_rmse_df_mutated %>% 
  pivot_longer(
    cols = c("1-RMSE"),
    names_to = "metric",
    values_to = "value"
  ) %>%
  group_by(tool_dmr_combination, metric, tumor_seq_tf_combination) %>%
  summarise(value = value, .groups = "drop")
```

Now, per metric, select only the column corresponding to it, go back to the wider format and transform it to a matrix.

```{r}
# RMSE matrix
wgbs_rmse_mat <- final_wgbs_rmse_df_long %>%
  filter(metric == "1-RMSE") %>%
  select(tool_dmr_combination, tumor_seq_tf_combination, value) %>%
  pivot_wider(names_from = tumor_seq_tf_combination, values_from = value) %>%
  column_to_rownames("tool_dmr_combination") %>%
  as.matrix()

# Define col names
cols <- colnames(wgbs_rmse_mat)
# Define cancer type order
cancer_order <- c("LUAD", "LUSC", "BRCA", "BLCA")

# Split names into parts
parts <- do.call(rbind, strsplit(cols, "_"))
cancer <- factor(parts[,1], levels = cancer_order)
middle <- as.numeric(parts[,3])
last   <- as.numeric(sub("M", "", parts[,2]))

# Order by: last → middle → cancer
sorted_cols <- cols[order(middle, cancer, last)]
wgbs_rmse_mat <- wgbs_rmse_mat[, sorted_cols]
```

Order the rows based on the Score values to give kind of a ranking

```{r}
# Compute row means of the Score matrix
score_means <- rowMeans(wgbs_rmse_mat)

# Order the row names based on descending Score (you can use decreasing = FALSE for ascending)
ordered_rows <- names(sort(score_means, decreasing = TRUE))

# Reorder all matrices by the ordered row names
wgbs_rmse_mat   <- wgbs_rmse_mat[ordered_rows, ]
```

And now we can plot the results in a heatmap using ComplexHeatmap package.

```{r}
# Define tumor types, tf and sequencing depths for the annotations
tf        <- rep(sort(unique(middle)), each = 16)
seq_depth <- rep(c("62M", "183M", "310M", "620M"), 40)
tumors    <- rep(c("LUAD", "LUSC", "BRCA", "BLCA"), each = 4, 10)


# Assign palettes
col_anno <- HeatmapAnnotation(
  Tumor    = tumors,
  ExpectedTF = tf,
  SeqDepth = seq_depth,
  col = list(
    SeqDepth = c("62M"="#FAEBDDFF","183M"="#F69C73FF","310M"="#A11A5BFF","620M"="#03051AFF"),
    Tumor = c("LUAD"="#ff7f0e", "BRCA"="#2ca02c", "LUSC"="#9467bd", "BLCA"="#8c564b"),
    ExpectedTF = c(  "1e-04" = "#DEEBF7", "0.001" = "#C6DBEF", "0.003" = "#9ECAE1",
                     "0.007" = "#6BAED6", "0.01"  = "#4292C6", "0.025" = "#2171B5",
                     "0.05"  = "#08519C", "0.1"   = "#08306B", "0.25"  = "#041836",
                     "0.5"   = "#000010")),
  show_annotation_name = FALSE
)

# Define heatmap settings shared across all plots
common_opts <- list(
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  column_names_rot = 45,
  show_row_dend = TRUE,
  show_column_names = FALSE,
  show_column_dend = FALSE,
  top_annotation = col_anno,
  column_split = tf
)

# Put all matrices and their titles/names in a list
heatmap_data <- list(
  "1-RMSE"         = wgbs_rmse_mat
)

# Generate heatmaps with do.call
heatmaps <- lapply(names(heatmap_data), function(name) {
  do.call(Heatmap, c(
    list(
      matrix = heatmap_data[[name]],
      name = name,
      column_title = name
    ),
    common_opts
  ))
})
draw(Reduce(`+`, heatmaps))
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/01_comprehensive_heatmaps"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
pdf(paste0(output_dir,"/wgbs_rmse_heatmap.pdf"), width = 12, height = 6)
draw(Reduce(`+`, heatmaps))
dev.off()
```


### Plot the data of RRBS

#### AUC-ROC and SCC

Here we will follow the same steps as before, but for the RRBS dataset.

```{r}
final_rrbs_df <- df.metrics %>% filter(tumor_type %in% c("CRC","LC","PC"))
```

Change the format of the dataframe to a long format

```{r}
# Create 2 columns corresponding to the deconvolution tool names collapsed with the sequencing depths and dmr tools
final_rrbs_df_mutated <- final_rrbs_df %>%
  mutate(tool_dmr_combination = paste(deconv_tool, dmr_tool, sep = "_")) %>%
  mutate(tumor_seq_combination = paste(tumor_type, seq_depth, sep = "_"))

# Make a long format of the df
final_rrbs_df_long <- final_rrbs_df_mutated %>% 
  pivot_longer(
    cols = c(AUC,SCC),
    names_to = "metric",
    values_to = "value"
  ) %>%
  group_by(tool_dmr_combination, metric, tumor_seq_combination) %>%
  summarise(value = value, .groups = "drop")
```

Now, per metric, select only the column corresponding to it, go back to the wider format and transform it to a matrix.

```{r}
# Specify the order of the columns for visualization purposes
desired_order <- c("LC_2M","LC_5M", "LC_10M", "LC_15M", "LC_20M",
                   "CRC_2M","CRC_5M", "CRC_10M", "CRC_15M", "CRC_20M",
                   "PC_2M","PC_5M", "PC_10M", "PC_15M", "PC_20M")

# AUC matrix
rrbs_auc_mat <- final_rrbs_df_long %>%
  filter(metric == "AUC") %>%
  select(tool_dmr_combination, tumor_seq_combination, value) %>%
  pivot_wider(names_from = tumor_seq_combination, values_from = value) %>%
  column_to_rownames("tool_dmr_combination") %>%
  as.matrix()
rrbs_auc_mat <- rrbs_auc_mat[, desired_order]

# SCC matrix
rrbs_scc_mat <- final_rrbs_df_long %>%
  filter(metric == "SCC") %>%
  select(tool_dmr_combination, tumor_seq_combination, value) %>%
  pivot_wider(names_from = tumor_seq_combination, values_from = value) %>%
  column_to_rownames("tool_dmr_combination") %>%
  as.matrix()
rrbs_scc_mat <- rrbs_scc_mat[, desired_order]
```

Order the rows based on the Score values to give kind of a ranking

```{r}
# Compute row means of the Score matrix
score_means <- rowMeans(rrbs_auc_mat) + rowMeans(rrbs_scc_mat)

# Order the row names based on descending Score (you can use decreasing = FALSE for ascending)
ordered_rows <- names(sort(score_means, decreasing = TRUE))

# Reorder all matrices by the ordered row names
rrbs_auc_mat   <- rrbs_auc_mat[ordered_rows, ]
rrbs_scc_mat   <- rrbs_scc_mat[ordered_rows, ]
```

And now we can plot the results in a heatmap using ComplexHeatmap package.


```{r}
seq_depth <- rep(c("2M", "5M", "10M", "15M", "20M"), 3)
tumors    <- rep(c("LC", "CRC", "PC"), each = 5)

col_anno <- HeatmapAnnotation(
  SeqDepth = seq_depth,
  Tumor    = tumors,
  col = list(
    SeqDepth = c("2M"="#FDE725FF","5M"="#7AD151FF","10M"="#2A788EFF","15M"="#414487FF","20M"="#440154FF"),
    Tumor = c("LC"="#ff7f0e", "CRC"="#2ca02c", "PC"="#9467bd")
  ),
  show_annotation_name = FALSE
)

# Define heatmap settings shared across all plots
common_opts <- list(
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  column_names_rot = 45,
  show_row_dend = TRUE,
  show_column_names = FALSE,
  show_column_dend = FALSE,
  top_annotation = col_anno,
  column_labels = tumors,
  column_split = seq_depth
)

# Put all matrices and their titles/names in a list
heatmap_data <- list(
  AUC         = rrbs_auc_mat,
  SCC         = rrbs_scc_mat
)

# Generate heatmaps with do.call
heatmaps <- lapply(names(heatmap_data), function(name) {
  do.call(Heatmap, c(
    list(
      matrix = heatmap_data[[name]],
      name = name,
      column_title = name
    ),
    common_opts
  ))
})
draw(Reduce(`+`, heatmaps))
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/01_comprehensive_heatmaps"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Save
pdf(paste0(output_dir,"/rrbs_scc_auc_heatmap.pdf"), width = 12, height = 6)
draw(Reduce(`+`, heatmaps))
dev.off()
```

#### RMSE

We are now ready to plot an comprehensive heatmap of the RRBS dataset describing all the different combinations for the RMSE metric, similarly to what we have done in the WGBS dataset.

```{r}
final_rrbs_rmse_df <- rmse_raw %>% filter(tumor_type %in% c("LC", "CRC", "PC"))
```

Change the format of the dataframe to a long format

```{r}
# Create 2 columns corresponding to the deconvolution tool names collapsed with the sequencing depths and dmr tools
final_rrbs_rmse_df_mutated <- final_rrbs_rmse_df %>%
  mutate(tool_dmr_combination = paste(deconv_tool, dmr_tool, sep = "_")) %>%
  mutate(tumor_seq_tf_combination = paste(tumor_type, seq_depth, expected_tf, sep = "_"))

# Make a long format of the df
final_rrbs_rmse_df_long <- final_rrbs_rmse_df_mutated %>% 
  pivot_longer(
    cols = c("1-RMSE"),
    names_to = "metric",
    values_to = "value"
  ) %>%
  group_by(tool_dmr_combination, metric, tumor_seq_tf_combination) %>%
  summarise(value = value, .groups = "drop")
```

Now, per metric, select only the column corresponding to it, go back to the wider format and transform it to a matrix.

```{r}
# RMSE matrix
rrbs_rmse_mat <- final_rrbs_rmse_df_long %>%
  filter(metric == "1-RMSE") %>%
  select(tool_dmr_combination, tumor_seq_tf_combination, value) %>%
  pivot_wider(names_from = tumor_seq_tf_combination, values_from = value) %>%
  column_to_rownames("tool_dmr_combination") %>%
  as.matrix()

# Define col names
cols <- colnames(rrbs_rmse_mat)
# Define cancer type order
cancer_order <- c("LC", "CRC", "PC")

# Split names into parts
parts <- do.call(rbind, strsplit(cols, "_"))
cancer <- factor(parts[,1], levels = cancer_order)
middle <- as.numeric(parts[,3])
last   <- as.numeric(sub("M", "", parts[,2]))

# Order by: last → middle → cancer
sorted_cols <- cols[order(middle, cancer, last)]
rrbs_rmse_mat <- rrbs_rmse_mat[, sorted_cols]
```

Order the rows based on the Score values to give kind of a ranking

```{r}
# Compute row means of the Score matrix
score_means <- rowMeans(rrbs_rmse_mat)

# Order the row names based on descending Score (you can use decreasing = FALSE for ascending)
ordered_rows <- names(sort(score_means, decreasing = TRUE))

# Reorder all matrices by the ordered row names
rrbs_rmse_mat   <- rrbs_rmse_mat[ordered_rows, ]
```

And now we can plot the results in a heatmap using ComplexHeatmap package.

```{r}
# Define tumor types, tf and sequencing depths for the annotations
tf        <- rep(sort(unique(middle)), each = 15)
seq_depth <- rep(c("2M", "5M", "10M", "15M", "20M"), 30)
tumors    <- rep(c("LC", "CRC", "PC"), each = 5, 10)

# Assign palettes
col_anno <- HeatmapAnnotation(
  Tumor    = tumors,
  ExpectedTF = tf,
  SeqDepth = seq_depth,
  col = list(
    SeqDepth = c("2M"="#FDE725FF","5M"="#7AD151FF","10M"="#2A788EFF","15M"="#414487FF","20M"="#440154FF"),
    Tumor = c("LC"="#ff7f0e", "CRC"="#2ca02c", "PC"="#9467bd"),
    ExpectedTF = c(  "1e-04" = "#DEEBF7", "0.001" = "#C6DBEF", "0.003" = "#9ECAE1",
                     "0.007" = "#6BAED6", "0.01"  = "#4292C6", "0.025" = "#2171B5",
                     "0.05"  = "#08519C", "0.1"   = "#08306B", "0.25"  = "#041836",
                     "0.5"   = "#000010")),
  show_annotation_name = FALSE
)

# Define heatmap settings shared across all plots
common_opts <- list(
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  column_names_rot = 45,
  show_row_dend = TRUE,
  show_column_names = FALSE,
  show_column_dend = FALSE,
  top_annotation = col_anno,
  column_split = tf
)

# Put all matrices and their titles/names in a list
heatmap_data <- list(
  "1-RMSE"         = rrbs_rmse_mat
)

# Generate heatmaps with do.call
heatmaps <- lapply(names(heatmap_data), function(name) {
  do.call(Heatmap, c(
    list(
      matrix = heatmap_data[[name]],
      name = name,
      column_title = name
    ),
    common_opts
  ))
})
draw(Reduce(`+`, heatmaps))
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/01_comprehensive_heatmaps"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Save
pdf(paste0(output_dir,"/rrbs_rmse_heatmap.pdf"), width = 12, height = 6)
draw(Reduce(`+`, heatmaps))
dev.off()
```

### Plot the data of cfRRBS

#### AUC-ROC and SCC

Here we will follow the same steps as before, but for the cfRRBS dataset.

```{r}
final_cfrrbs_df <- df.metrics %>% filter(tumor_type %in% c("NBL"))
```

Change the format of the dataframe to a long format

```{r}
# Create 2 columns corresponding to the deconvolution tool names collapsed with the sequencing depths and dmr tools
final_cfrrbs_df_mutated <- final_cfrrbs_df %>%
  mutate(tool_dmr_combination = paste(deconv_tool, dmr_tool, sep = "_")) %>%
  mutate(tumor_seq_combination = paste(tumor_type, seq_depth, sep = "_"))

# Make a long format of the df
final_cfrrbs_df_long <- final_cfrrbs_df_mutated %>% 
  pivot_longer(
    cols = c(AUC,SCC),
    names_to = "metric",
    values_to = "value"
  ) %>%
  group_by(tool_dmr_combination, metric, tumor_seq_combination) %>%
  summarise(value = value, .groups = "drop")
```

Now, per metric, select only the column corresponding to it, go back to the wider format and transform it to a matrix.

```{r}
# Specify the order of the columns for visualization purposes
desired_order <- c("NBL_2M","NBL_5M", "NBL_10M", "NBL_15M", "NBL_20M")

# AUC matrix
cfrrbs_auc_mat <- final_cfrrbs_df_long %>%
  filter(metric == "AUC") %>%
  select(tool_dmr_combination, tumor_seq_combination, value) %>%
  pivot_wider(names_from = tumor_seq_combination, values_from = value) %>%
  column_to_rownames("tool_dmr_combination") %>%
  as.matrix()
cfrrbs_auc_mat <- cfrrbs_auc_mat[, desired_order]

# SCC matrix
cfrrbs_scc_mat <- final_cfrrbs_df_long %>%
  filter(metric == "SCC") %>%
  select(tool_dmr_combination, tumor_seq_combination, value) %>%
  pivot_wider(names_from = tumor_seq_combination, values_from = value) %>%
  column_to_rownames("tool_dmr_combination") %>%
  as.matrix()
cfrrbs_scc_mat <- cfrrbs_scc_mat[, desired_order]
```

Order the rows based on the Score values to give kind of a ranking

```{r}
# Compute row means of the Score matrix
score_means <- rowMeans(cfrrbs_auc_mat) + rowMeans(cfrrbs_scc_mat)

# Order the row names based on descending Score (you can use decreasing = FALSE for ascending)
ordered_rows <- names(sort(score_means, decreasing = TRUE))

# Reorder all matrices by the ordered row names
cfrrbs_auc_mat   <- cfrrbs_auc_mat[ordered_rows, ]
cfrrbs_scc_mat   <- cfrrbs_scc_mat[ordered_rows, ]
```

And now we can plot the results in a heatmap using ComplexHeatmap package.


```{r}
seq_depth <- factor(rep(c("2M", "5M", "10M", "15M", "20M"), 1),
                    levels = c("2M", "5M", "10M", "15M", "20M"))
tumors    <- rep(c("NBL"), each = 5)

col_anno <- HeatmapAnnotation(
  SeqDepth = seq_depth,
  Tumor    = tumors,
  col = list(
    SeqDepth = c("2M"="#FDE725FF","5M"="#7AD151FF","10M"="#2A788EFF","15M"="#414487FF","20M"="#440154FF"),
    Tumor = c("NBL"="#ff7f0e")
  ),
  show_annotation_name = FALSE
)

# Define heatmap settings shared across all plots
common_opts <- list(
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  column_names_rot = 45,
  show_row_dend = TRUE,
  show_column_names = FALSE,
  show_column_dend = FALSE,
  top_annotation = col_anno,
  column_labels = tumors,
  column_split = seq_depth
)

# Put all matrices and their titles/names in a list
heatmap_data <- list(
  AUC         = cfrrbs_auc_mat,
  SCC         = cfrrbs_scc_mat
)

# Generate heatmaps with do.call
heatmaps <- lapply(names(heatmap_data), function(name) {
  do.call(Heatmap, c(
    list(
      matrix = heatmap_data[[name]],
      name = name,
      column_title = name
    ),
    common_opts
  ))
})
draw(Reduce(`+`, heatmaps))
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/01_comprehensive_heatmaps"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Save
pdf(paste0(output_dir,"/cfrrbs_scc_auc_heatmap.pdf"), width = 12, height = 6)
draw(Reduce(`+`, heatmaps))
dev.off()
```

#### RMSE

We are now ready to plot an comprehensive heatmap of the RRBS dataset describing all the different combinations for the RMSE metric, similarly to what we have done in the WGBS dataset.

```{r}
final_cfrrbs_rmse_df <- rmse_raw %>% filter(tumor_type %in% c("NBL"))
```

Change the format of the dataframe to a long format

```{r}
# Create 2 columns corresponding to the deconvolution tool names collapsed with the sequencing depths and dmr tools
final_cfrrbs_rmse_df_mutated <- final_cfrrbs_rmse_df %>%
  mutate(tool_dmr_combination = paste(deconv_tool, dmr_tool, sep = "_")) %>%
  mutate(tumor_seq_tf_combination = paste(tumor_type, seq_depth, expected_tf, sep = "_"))

# Make a long format of the df
final_cfrrbs_rmse_df_long <- final_cfrrbs_rmse_df_mutated %>% 
  pivot_longer(
    cols = c("1-RMSE"),
    names_to = "metric",
    values_to = "value"
  ) %>%
  group_by(tool_dmr_combination, metric, tumor_seq_tf_combination) %>%
  summarise(value = value, .groups = "drop")
```

Now, per metric, select only the column corresponding to it, go back to the wider format and transform it to a matrix.

```{r}
# RMSE matrix
cfrrbs_rmse_mat <- final_cfrrbs_rmse_df_long %>%
  filter(metric == "1-RMSE") %>%
  select(tool_dmr_combination, tumor_seq_tf_combination, value) %>%
  pivot_wider(names_from = tumor_seq_tf_combination, values_from = value) %>%
  column_to_rownames("tool_dmr_combination") %>%
  as.matrix()

# Define col names
cols <- colnames(cfrrbs_rmse_mat)
# Define cancer type order
cancer_order <- c("NBL")

# Split names into parts
parts <- do.call(rbind, strsplit(cols, "_"))
cancer <- factor(parts[,1], levels = cancer_order)
middle <- as.numeric(parts[,3])
last   <- as.numeric(sub("M", "", parts[,2]))

# Order by: last → middle → cancer
sorted_cols <- cols[order(middle, cancer, last)]
cfrrbs_rmse_mat <- cfrrbs_rmse_mat[, sorted_cols]
```

Order the rows based on the Score values to give kind of a ranking

```{r}
# Compute row means of the Score matrix
score_means <- rowMeans(cfrrbs_rmse_mat)

# Order the row names based on descending Score (you can use decreasing = FALSE for ascending)
ordered_rows <- names(sort(score_means, decreasing = TRUE))

# Reorder all matrices by the ordered row names
cfrrbs_rmse_mat   <- cfrrbs_rmse_mat[ordered_rows, ]
```

And now we can plot the results in a heatmap using ComplexHeatmap package.

```{r}
# Define tumor types, tf and sequencing depths for the annotations
tf        <- rep(sort(factor(unique(middle))), each = 5)
seq_depth <- rep(c("2M", "5M", "10M", "15M", "20M"), 10)
tumors    <- rep(c("NBL"), each = 50)

# Assign palettes
col_anno <- HeatmapAnnotation(
  Tumor    = tumors,
  ExpectedTF = tf,
  SeqDepth = seq_depth,
  col = list(
    SeqDepth = c("2M"="#FDE725FF","5M"="#7AD151FF","10M"="#2A788EFF","15M"="#414487FF","20M"="#440154FF"),
    Tumor = c("NBL"="#ff7f0e"),
    ExpectedTF = c(  "1e-04" = "#DEEBF7", "0.001" = "#C6DBEF", "0.003" = "#9ECAE1",
                     "0.007" = "#6BAED6", "0.01"  = "#4292C6", "0.025" = "#2171B5",
                     "0.05"  = "#08519C", "0.1"   = "#08306B", "0.25"  = "#041836",
                     "0.5"   = "#000010")
  ),
  show_annotation_name = FALSE
)

# Define heatmap settings shared across all plots
common_opts <- list(
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  column_names_rot = 45,
  show_row_dend = TRUE,
  show_column_names = FALSE,
  show_column_dend = FALSE,
  top_annotation = col_anno,
  column_split = tf
)

# Put all matrices and their titles/names in a list
heatmap_data <- list(
  "1-RMSE"         = cfrrbs_rmse_mat
)

# Generate heatmaps with do.call
heatmaps <- lapply(names(heatmap_data), function(name) {
  do.call(Heatmap, c(
    list(
      matrix = heatmap_data[[name]],
      name = name,
      column_title = name
    ),
    common_opts
  ))
})
draw(Reduce(`+`, heatmaps))
```



```{r}
# Create the main directory to save the plots
output_dir <- "../plots/01_comprehensive_heatmaps"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
pdf(paste0(output_dir, "/cfrrbs_rmse_heatmap.pdf"), width = 12, height = 6)
draw(Reduce(`+`, heatmaps))
dev.off()
```
