---
title: "04. Analysis of reference free tools"
author: "Edoardo Giuili"
date: "2025-09-24"
output: html_document
---

## Introduction

In this tutorial we want to show how to compare the performance of the reference-free methods with the performance of the reference-based methods. Specifically, we want to see whether the former ones out- or underperform the latter.

### Setting up the environment

First, set up the main directory

```{r setup}
# Don't change this if you git cloned the entire GitHub repository and you are working in the notebooks/ folder
knitr::opts_knit$set(root.dir = './')
```

Import the necessary libraries for the analysis

```{r message=FALSE, warning=FALSE}
library(ggpubr)
library(ggplot2)
library(dplyr)
library(pROC)
library(tidytext)
```

Import the useful functions and the metrics

```{r}
source("../R/compute_auc.R")
source("../R/compute_lod.R")
# RMSE
rmse <- function(actual, predicted) {
  round(sqrt(mean((actual - predicted)^2)),4)
}
# Spearman's rank correlation coefficient (SCC)
scc <- function(actual, predicted) {
  cor(actual, predicted, method = "spearman", use = "pairwise.complete.obs")
}
```

Read the results generated by both reference-free and reference-based tools and concatenate them in a unique dataframe.

```{r}
bench_refree <- read.csv("../resources/refree_final_dataset.csv", row.names = 1)
bench_refree <- bench_refree %>% 
  filter(!is.na(bench_refree$predicted_tf)) %>%
  mutate(fraction_class = ifelse( expected_tf <= 0.025, "low", ifelse( expected_tf >0.025 & expected_tf <= 0.1, "medium", "high")))
bench_refbased <- read.csv("../resources/final_benchmarking_dataset.csv")
bench_refbased <- bench_refbased %>% 
  filter(!is.na(bench_refbased$predicted_tf))  %>%
  mutate(fraction_class = ifelse( expected_tf <= 0.025, "low", ifelse( expected_tf >0.025 & expected_tf <= 0.1, "medium", "high")))
names(bench_refree)[5] <- "dmr_tool"
bench <- rbind(bench_refbased, bench_refree)
```

Select the WGBS and RRBS datasets

```{r}
in_silico_wgbs <- subset(bench, bench$seq_method == "wgbs")
in_silico_rrbs <- subset(bench, bench$seq_method == "rrbs")
in_silico_cfrrbs <- subset(bench, bench$seq_method == "cfrrbs" & bench$mixture_type == "in_silico")
```


## WGBS-TT

As always, we will start our comparison analysing the WGBS-TT dataset, to then proceed with the RRBS-TT and RRBS-CL datasets. 

### AUC-ROC values

We will first start computing and plotting the AUC-ROC values.


```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(in_silico_wgbs$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(in_silico_wgbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_wgbs, in_silico_wgbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)
          
          # If the dataframe is empty, skip
          if (dim(filt_df)[1] == 0) { next }
          
          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_wgbs <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(AUC=mean(AUC),.groups = "drop") %>%
    mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```

```{r}
# Rank the tools by mean RMSE
median_rank <- auc_wgbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(AUC)) %>%
  arrange(desc(Median))
plot_auc <- auc_wgbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_auc, aes(x = deconv_tool, y = AUC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "WGBS-TT dataset",
    x = "",
    y = "AUC-ROC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
     "Reference-based" = "#009E73",
     "Reference-free" = "#E69F00"
   )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0,1)
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("AUC-ROC_WGBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

We can now try to see how the reference-free methods perform at increasing sequencing depth

```{r warning=FALSE}
plot_auc_refree <- subset(auc_wgbs, auc_wgbs$deconv_tool %in% c("RefFreeCellMix","MeDeCom"))
plot_auc_refree$seq_depth <- factor(plot_auc_refree$seq_depth, levels = c("62M","183M", "310M", "620M"))
# Define comparisons
my_comparisons <- list(
  c("62M", "183M"),
  c("183M", "310M"),
  c("183M", "620M"),
  c("310M", "620M")
)
# Plot
p <- ggplot(data = plot_auc_refree, aes(
    x = seq_depth, y = AUC, fill = seq_depth)
  ) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  facet_wrap(~deconv_tool, ncol = 5) +
  stat_compare_means(
      method = "wilcox.test", comparisons = my_comparisons,
      p.adjust.method = "BH", label = "adj.p", hide.ns = F,                    
      method.args = list(alternative="less"),
      label.y = c(1, 1.05, 1.10, 1.15), size = 3
  ) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major.y = element_line(color = "grey90"),
          panel.grid.minor.y = element_line(color = "grey90")) +
  labs(title = "WGBS-TT dataset",
       x = "", y = "AUC") +
  scale_y_continuous(
    breaks = c(0.4, 0.6, 0.8, 1) # only show these breaks
  ) +
  scale_fill_manual(values = c("62M"="#FAEBDDFF","183M"="#F69C73FF","310M"="#A11A5BFF","620M"="#03051AFF")) +
  coord_cartesian(ylim=c(0.6,1.25))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("robustness_AUC_seqdepth_refree_WGBS.pdf"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### SCC

Now, we compute the SCC. The SCC is computed for each combination pooling all the observed tumor fractions together to get a value for each combination. The SCC is then min-max scaled.

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_wgbs <- in_silico_wgbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop") %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- scc_wgbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(SCC)) %>%
  arrange(desc(Median))
plot_scc <- scc_wgbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_scc, aes(x = deconv_tool, y = SCC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "WGBS-TT dataset",
    x = "",
    y = "SCC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0,1)
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("SCC_WGBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

We can now try to see how the reference-free methods perform at increasing sequencing depth

```{r warning=FALSE}
plot_scc_refree <- subset(scc_wgbs, scc_wgbs$deconv_tool %in% c("RefFreeCellMix","MeDeCom"))
plot_scc_refree$seq_depth <- factor(plot_scc_refree$seq_depth, levels = c("62M","183M", "310M", "620M"))
# Define comparisons
my_comparisons <- list(
  c("62M", "183M"),
  c("183M", "310M"),
  c("183M", "620M"),
  c("310M", "620M")
)
# Plot
p <- ggplot(data = plot_scc_refree, aes(
    x = seq_depth, y = SCC, fill = seq_depth)
  ) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  facet_wrap(~deconv_tool, ncol = 5) +
  stat_compare_means(
      method = "wilcox.test", comparisons = my_comparisons,
      p.adjust.method = "BH", label = "adj.p", hide.ns = F,                    
      method.args = list(alternative="less"),
      label.y = c(1, 1.05, 1.10, 1.15), size = 3
  ) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major.y = element_line(color = "grey90"),
          panel.grid.minor.y = element_line(color = "grey90")) +
  labs(title = "WGBS-TT dataset",
       x = "", y = "SCC") +
  scale_y_continuous(
    breaks = c(0.4, 0.6, 0.8, 1) # only show these breaks
  ) +
  scale_fill_manual(values = c("62M"="#FAEBDDFF","183M"="#F69C73FF","310M"="#A11A5BFF","620M"="#03051AFF")) +
  coord_cartesian(ylim=c(0.6,1.25))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("robustness_SCC_seqdepth_refree_WGBS.pdf"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### RMSE 

Compute the RMSE.

```{r}
rmse_wgbs <- in_silico_wgbs %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, fraction_class) %>%
  summarize(RMSE = 1-rmse(expected_tf, predicted_tf)) %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- rmse_wgbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(RMSE)) %>%
  arrange(desc(Median))
plot_rmse <- rmse_wgbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))
plot_rmse$fraction_class <- factor(plot_rmse$fraction_class, levels = c("low","medium","high"))

p <- ggplot(plot_rmse, aes(
  x = reorder_within(deconv_tool,-RMSE, fraction_class, median),
  y = RMSE, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "WGBS-TT dataset",
    x = "",
    y = "1-RMSE"
  ) +
  facet_wrap(~fraction_class, scales = "free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_x_reordered() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0.5,1)
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("RMSE_WGBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

```{r warning=FALSE}
plot_rmse_refree <- subset(rmse_wgbs, rmse_wgbs$deconv_tool %in% c("RefFreeCellMix","MeDeCom"))
plot_rmse_refree$seq_depth <- factor(plot_rmse_refree$seq_depth, levels = c("62M","183M", "310M", "620M"))
# Define comparisons
my_comparisons <- list(
  c("62M", "183M"),
  c("183M", "310M"),
  c("183M", "620M"),
  c("310M", "620M")
)
# Plot
p <- ggplot(data = plot_rmse_refree, aes(
    x = seq_depth, y = RMSE, fill = seq_depth)
  ) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  facet_wrap(~deconv_tool, ncol = 5) +
  stat_compare_means(
      method = "wilcox.test", comparisons = my_comparisons,
      p.adjust.method = "BH", label = "adj.p", hide.ns = F,                    
      method.args = list(alternative="less"),
      label.y = c(1, 1.05, 1.10, 1.15), size = 3
  ) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major.y = element_line(color = "grey90"),
          panel.grid.minor.y = element_line(color = "grey90")) +
  labs(title = "WGBS-TT dataset",
       x = "", y = "1-RMSE") +
  scale_y_continuous(
    breaks = c(0.4, 0.6, 0.8, 1) # only show these breaks
  ) +
  scale_fill_manual(values = c("62M"="#FAEBDDFF","183M"="#F69C73FF","310M"="#A11A5BFF","620M"="#03051AFF")) +
  coord_cartesian(ylim=c(0.5,1.25))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("robustness_RMSE_seqdepth_refree_WGBS.pdf"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### LoD

```{r}
my_comp <- list(
  c("0", "1e-04"), 
  c("0", "0.001"), 
  c("0", "0.003"), 
  c("0", "0.007"),
  c("0", "0.01"),
  c("0", "0.025")
)
filt_df <- subset(in_silico_wgbs, in_silico_wgbs$expected_tf %in% c("0", "1e-04","0.001", "0.003", "0.007","0.01","0.025") & in_silico_wgbs$dmr_tool=="rrbs20-200" & in_silico_wgbs$seq_depth=="310M" & in_silico_wgbs$deconv_tool=="MeDeCom" & in_silico_wgbs$tumor_type=="LUAD")
stats <- compute_lod(filt_df, my_comp, correction.method = "BH")
p <- filt_df %>%
  group_by(deconv_tool) %>%
  ggplot(aes(x = as.factor(expected_tf), y = predicted_tf)) +
  geom_boxplot() +
  geom_point(size = 1) +
  labs(
    title = paste(),
    x = "",
    y = "Estimated tumoral fraction (%)"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5)  # add horizontal grid
  ) + 
  stat_pvalue_manual(stats, label = "star", xmin = "group1", xmax = "group2", y.position = "y.position")
p
```


And finally compute the LoD score

```{r warning=FALSE}
# Define the comparisons and an empty dataframe to store the results
my_comparisons <- list(
  c("0", "1e-04"), 
  c("0", "0.001"), 
  c("0", "0.003"), 
  c("0", "0.007"),
  c("0", "0.01"),
  c("0", "0.025"),
  c("0", "0.05"),
  c("0", "0.1"),
  c("0", "0.25"),
  c("0", "0.5")
)
lod_refree_wgbs <- data.frame()
in_silico_refree_wgbs <- subset(in_silico_wgbs, in_silico_wgbs$deconv_tool %in% c("MeDeCom","RefFreeCellMix"))

# Loop over the type of tumors
for (type in unique(in_silico_refree_wgbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_refree_wgbs, in_silico_refree_wgbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      filt_dmr <- subset(filt_depth, filt_depth$dmr_tool == dmrtool)
      for (deconv in unique(filt_depth$deconv_tool)) {
        
        # Filter over the deconvolution tool and run the statistical test to find the LoD
        filt_df <- subset(filt_dmr, filt_dmr$deconv_tool == deconv)
        stats <- compute_lod(filt_df, my_comparisons, correction.method = "BH")

        lod <- stats %>%
          arrange(as.numeric(group2)) %>%
          filter(as.numeric(p.adj) < 0.01) %>%
          slice(1) %>%
          pull(group2)
            
        # If lod is empty or NA, assign 1
        if (length(lod) == 0 || is.na(lod)) {
          lod <- 1
        }
        
        # Concatenate the results into the dataframe
        tmp <- data.frame(deconv_tool = deconv, dmr_tool = dmrtool, tumor_type = type, seq_depth = seqdepth, lod = lod, stringsAsFactors = F)
        lod_refree_wgbs <- rbind(lod_refree_wgbs, tmp)
      }
    }
  }
}
```


```{r}
lod_refbased_wgbs <- read.csv("../resources/lod_allstats.csv", row.names = 1)[1:5]
lod_refbased_wgbs <- subset(lod_refbased_wgbs, lod_refbased_wgbs$tumor_type %in% c("BRCA","BLCA","LUAD","LUSC"))

all.lods <- rbind(lod_refbased_wgbs, lod_refree_wgbs)  %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- all.lods %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(as.numeric(lod))) %>%
  arrange(Median)
plot_lods <- all.lods %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_lods, aes(
  x = deconv_tool, y = as.numeric(lod), 
  color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20") +
  labs(
    title = "WGBS-TT dataset",
    x = "",
    y = "LoD"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("LoD_WGBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

## RRBS-TT

We can now proceed with the RRBS-TT dataset.

### AUC-ROC values

We will first start computing and plotting the AUC-ROC values.


```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(in_silico_rrbs$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(in_silico_rrbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_rrbs, in_silico_rrbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)
          
          # If the dataframe is empty, skip
          if (dim(filt_df)[1] == 0) { next }
          
          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_rrbs <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(AUC=mean(AUC),.groups = "drop") %>%
    mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```

```{r}
# Rank the tools by mean RMSE
median_rank <- auc_rrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(AUC)) %>%
  arrange(desc(Median))
plot_auc <- auc_rrbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_auc, aes(x = deconv_tool, y = AUC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "RRBS-TT dataset",
    x = "",
    y = "AUC-ROC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
     "Reference-based" = "#009E73",
     "Reference-free" = "#E69F00"
   )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("AUC-ROC_RRBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

And we can now check the influence of sequencing depth on AUC

```{r warning=FALSE}
plot_auc_refree <- subset(auc_rrbs, auc_rrbs$deconv_tool %in% c("RefFreeCellMix","MeDeCom"))
plot_auc_refree$seq_depth <- factor(plot_auc_refree$seq_depth, levels = c("2M", "5M", "10M", "15M", "20M"))

# Define comparisons
my_comparisons <- list(
  c("2M", "5M"),
  c("5M", "10M"),
  c("10M","15M"),
  c("10M", "20M"),
  c("15M", "20M")
)
# Plot
p <- ggplot(data = plot_auc_refree, aes(
    x = seq_depth, y = AUC, fill = seq_depth)
  ) +
  #geom_jitter(size=1.5, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  facet_wrap(~deconv_tool, ncol = 5) +
  stat_compare_means(
      method = "wilcox.test", comparisons = my_comparisons,
      p.adjust.method = "BH", label = "adj.p", hide.ns = F,                    
      method.args = list(alternative="less"),
      label.y = c(0.95, 0.98, 1, 1.05,1.1), size = 3
  ) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major.y = element_line(color = "grey90"),
          panel.grid.minor.y = element_line(color = "grey90")) +
  labs(title = "RRBS-TT dataset",
       x = "", y = "AUC") +
  scale_y_continuous(
    breaks = c(0.8, 0.9, 1) # only show these breaks
  ) +
  scale_fill_manual(values = c("2M"="#FDE725FF","5M"="#7AD151FF","10M"="#2A788EFF","15M"="#414487FF","20M"="#440154FF"))  +
  coord_cartesian(ylim=c(0.8,1.15))
p
```
```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("robustness_AUC_seqdepth_refree_RRBS.pdf"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### SCC

Now, we compute the SCC.

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_rrbs <- in_silico_rrbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop") %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- scc_rrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(SCC)) %>%
  arrange(desc(Median))
plot_scc <- scc_rrbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_scc, aes(x = deconv_tool, y = SCC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "RRBS-TT dataset",
    x = "",
    y = "SCC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0,1)
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("SCC_RRBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

We can now try to see how the reference-free methods perform at increasing sequencing depth

```{r warning=FALSE}
plot_scc_refree <- subset(scc_rrbs, scc_rrbs$deconv_tool %in% c("RefFreeCellMix","MeDeCom"))
plot_scc_refree$seq_depth <- factor(plot_scc_refree$seq_depth, levels = c("2M", "5M", "10M", "15M", "20M"))

# Define comparisons
my_comparisons <- list(
  c("2M", "5M"),
  c("5M", "10M"),
  c("10M","15M"),
  c("10M", "20M"),
  c("15M", "20M")
)
# Plot
p <- ggplot(data = plot_scc_refree, aes(
    x = seq_depth, y = SCC, fill = seq_depth)
  ) +
  #geom_jitter(size=1.5, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  facet_wrap(~deconv_tool, ncol = 5) +
  stat_compare_means(
      method = "wilcox.test", comparisons = my_comparisons,
      p.adjust.method = "BH", label = "adj.p", hide.ns = F,                    
      method.args = list(alternative="less"),
      label.y = c(1, 1.02, 1.04, 1.06, 1.08), size = 3
  ) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major.y = element_line(color = "grey90"),
          panel.grid.minor.y = element_line(color = "grey90")) +
  labs(title = "RRBS-TT dataset",
       x = "", y = "SCC") +
  scale_y_continuous(
    breaks = c(0.9, 1) # only show these breaks
  ) +
  scale_fill_manual(values = c("2M"="#FDE725FF","5M"="#7AD151FF","10M"="#2A788EFF","15M"="#414487FF","20M"="#440154FF"))  +
  coord_cartesian(ylim=c(0.85,1.1))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("robustness_SCC_seqdepth_refree_RRBS.pdf"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### RMSE 

Compute the RMSE.

```{r}
rmse_rrbs <- in_silico_rrbs %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, fraction_class) %>%
  summarize(RMSE = 1-rmse(expected_tf, predicted_tf)) %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- rmse_rrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(RMSE)) %>%
  arrange(desc(Median))
plot_rmse <- rmse_rrbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))
plot_rmse$fraction_class <- factor(plot_rmse$fraction_class, levels = c("low","medium","high"))

p <- ggplot(plot_rmse, aes(
  x = reorder_within(deconv_tool,-RMSE, fraction_class, median),
  y = RMSE, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "RRBS-TT dataset",
    x = "",
    y = "1-RMSE"
  ) +
  facet_wrap(~fraction_class, scales = "free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_x_reordered() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0.5,1)
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("RMSE_RRBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

We can now try to see how the reference-free methods perform at increasing sequencing depth

```{r warning=FALSE}
plot_rmse_refree <- subset(rmse_rrbs, rmse_rrbs$deconv_tool %in% c("RefFreeCellMix","MeDeCom"))
plot_rmse_refree$seq_depth <- factor(plot_rmse_refree$seq_depth, levels = c("2M", "5M", "10M", "15M", "20M"))

# Define comparisons
my_comparisons <- list(
  c("2M", "5M"),
  c("5M", "10M"),
  c("10M","15M"),
  c("10M", "20M"),
  c("15M", "20M")
)
# Plot
p <- ggplot(data = plot_rmse_refree, aes(
    x = seq_depth, y = RMSE, fill = seq_depth)
  ) +
  #geom_jitter(size=1.5, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  facet_wrap(~deconv_tool, ncol = 5) +
  stat_compare_means(
      method = "wilcox.test", comparisons = my_comparisons,
      p.adjust.method = "BH", label = "adj.p", hide.ns = F,                    
      method.args = list(alternative="less"),
      label.y = c(1, 1.02, 1.04, 1.08, 1.13), size = 3
  ) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major.y = element_line(color = "grey90"),
          panel.grid.minor.y = element_line(color = "grey90")) +
  labs(title = "RRBS-TT dataset",
       x = "", y = "1-RMSE") +
  scale_y_continuous(
    breaks = c(0.4, 0.6, 0.8, 1) # only show these breaks
  ) +
  scale_fill_manual(values = c("2M"="#FDE725FF","5M"="#7AD151FF","10M"="#2A788EFF","15M"="#414487FF","20M"="#440154FF"))  +
  coord_cartesian(ylim=c(0.4,1.15))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("robustness_RMSE_seqdepth_refree_RRBS.pdf"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### LoD

And finally compute the LoD score

```{r warning=FALSE}
# Define the comparisons and an empty dataframe to store the results
my_comparisons <- list(
  c("0", "1e-04"), 
  c("0", "0.001"), 
  c("0", "0.003"), 
  c("0", "0.007"),
  c("0", "0.01"),
  c("0", "0.025"),
  c("0", "0.05"),
  c("0", "0.1"),
  c("0", "0.25"),
  c("0", "0.5")
)
lod_refree_rrbs <- data.frame()
in_silico_refree_rrbs <- subset(in_silico_rrbs, in_silico_rrbs$deconv_tool %in% c("MeDeCom","RefFreeCellMix"))

# Loop over the type of tumors
for (type in unique(in_silico_refree_rrbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_refree_rrbs, in_silico_refree_rrbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      filt_dmr <- subset(filt_depth, filt_depth$dmr_tool == dmrtool)
      for (deconv in unique(filt_depth$deconv_tool)) {
        
        # Filter over the deconvolution tool and run the statistical test to find the LoD
        filt_df <- subset(filt_dmr, filt_dmr$deconv_tool == deconv)
        stats <- compute_lod(filt_df, my_comparisons, correction.method = "BH")

        lod <- stats %>%
          arrange(as.numeric(group2)) %>%
          filter(as.numeric(p.adj) < 0.01) %>%
          slice(1) %>%
          pull(group2)
            
        # If lod is empty or NA, assign 1
        if (length(lod) == 0 || is.na(lod)) {
          lod <- 1
        }
        
        # Concatenate the results into the dataframe
        tmp <- data.frame(deconv_tool = deconv, dmr_tool = dmrtool, tumor_type = type, seq_depth = seqdepth, lod = lod, stringsAsFactors = F)
        lod_refree_rrbs <- rbind(lod_refree_rrbs, tmp)
      }
    }
  }
}
```


```{r}
lod_refbased_rrbs <- read.csv("../resources/lod_allstats.csv", row.names = 1)[1:5]
lod_refbased_rrbs <- subset(lod_refbased_rrbs, lod_refbased_rrbs$tumor_type %in% c("CRC","LC","PC"))
all.lods <- rbind(lod_refbased_rrbs, lod_refree_rrbs)  %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- all.lods %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(as.numeric(lod))) %>%
  arrange(Median)
plot_lods <- all.lods %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_lods, aes(
  x = deconv_tool, y = as.numeric(lod), 
  color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20") +
  labs(
    title = "RRBS-TT dataset",
    x = "",
    y = "LoD"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("LoD_RRBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

## RRBS-CL

We can now proceed with the RRBS-CL dataset.

### AUC-ROC values

We will first start computing and plotting the AUC-ROC values.


```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(in_silico_cfrrbs$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(in_silico_cfrrbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_cfrrbs, in_silico_cfrrbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)
          
          # If the dataframe is empty, skip
          if (dim(filt_df)[1] == 0) { next }
          
          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_cfrrbs <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(AUC=mean(AUC),.groups = "drop") %>%
    mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```

```{r}
# Rank the tools by mean RMSE
median_rank <- auc_cfrrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(AUC)) %>%
  arrange(desc(Median))
plot_auc <- auc_cfrrbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_auc, aes(x = deconv_tool, y = AUC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "RRBS-CL dataset",
    x = "",
    y = "AUC-ROC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
     "Reference-based" = "#009E73",
     "Reference-free" = "#E69F00"
   )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("AUC-ROC_RRBS-CL.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

We can now try to see how the reference-free methods perform at increasing sequencing depth

```{r warning=FALSE}
plot_scc_refree <- subset(auc_cfrrbs, auc_cfrrbs$deconv_tool %in% c("RefFreeCellMix","MeDeCom"))
plot_scc_refree$seq_depth <- factor(plot_scc_refree$seq_depth, levels = c("2M", "5M", "10M", "15M", "20M"))

# Define comparisons
my_comparisons <- list(
  c("2M", "5M"),
  c("5M", "10M"),
  c("10M","15M"),
  c("10M", "20M"),
  c("15M", "20M")
)
# Plot
p <- ggplot(data = plot_scc_refree, aes(
    x = seq_depth, y = AUC, fill = seq_depth)
  ) +
  #geom_jitter(size=1.5, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  facet_wrap(~deconv_tool, ncol = 5) +
  stat_compare_means(
      method = "wilcox.test", comparisons = my_comparisons,
      p.adjust.method = "BH", label = "adj.p", hide.ns = F,                    
      method.args = list(alternative="less"),
      label.y = c(0.97, 1, 1.02, 1.04, 1.06), size = 3
  ) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major.y = element_line(color = "grey90"),
          panel.grid.minor.y = element_line(color = "grey90")) +
  labs(title = "RRBS-CL dataset",
       x = "", y = "AUC") +
  scale_y_continuous(
    breaks = c(0.9, 1) # only show these breaks
  ) +
  scale_fill_manual(values = c("2M"="#FDE725FF","5M"="#7AD151FF","10M"="#2A788EFF","15M"="#414487FF","20M"="#440154FF"))  +
  coord_cartesian(ylim=c(0.9,1.1))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("robustness_AUC_seqdepth_refree_cfRRBS.pdf"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### SCC

Now, we compute the SCC.

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_cfrrbs <- in_silico_cfrrbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop") %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- scc_cfrrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(SCC)) %>%
  arrange(desc(Median))
plot_scc <- scc_cfrrbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_scc, aes(x = deconv_tool, y = SCC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "RRBS-CL dataset",
    x = "",
    y = "SCC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0,1)
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("SCC_RRBS-CL.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

We can now try to see how the reference-free methods perform at increasing sequencing depth

```{r warning=FALSE}
plot_scc_refree <- subset(scc_cfrrbs, scc_cfrrbs$deconv_tool %in% c("RefFreeCellMix","MeDeCom"))
plot_scc_refree$seq_depth <- factor(plot_scc_refree$seq_depth, levels = c("2M", "5M", "10M", "15M", "20M"))

# Define comparisons
my_comparisons <- list(
  c("2M", "5M"),
  c("5M", "10M"),
  c("10M","15M"),
  c("10M", "20M"),
  c("15M", "20M")
)
# Plot
p <- ggplot(data = plot_scc_refree, aes(
    x = seq_depth, y = SCC, fill = seq_depth)
  ) +
  #geom_jitter(size=1.5, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  facet_wrap(~deconv_tool, ncol = 5) +
  stat_compare_means(
      method = "wilcox.test", comparisons = my_comparisons,
      p.adjust.method = "BH", label = "adj.p", hide.ns = F,                    
      method.args = list(alternative="less"),
      label.y = c(1, 1.02, 1.04, 1.06, 1.08), size = 3
  ) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major.y = element_line(color = "grey90"),
          panel.grid.minor.y = element_line(color = "grey90")) +
  labs(title = "RRBS-CL dataset",
       x = "", y = "SCC") +
  scale_y_continuous(
    breaks = c(0.9, 1) # only show these breaks
  ) +
  scale_fill_manual(values = c("2M"="#FDE725FF","5M"="#7AD151FF","10M"="#2A788EFF","15M"="#414487FF","20M"="#440154FF"))  +
  coord_cartesian(ylim=c(0.9,1.1))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("robustness_SCC_seqdepth_refree_cfRRBS.pdf"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### RMSE 

Compute the RMSE.

```{r}
rmse_cfrrbs <- in_silico_cfrrbs %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, fraction_class) %>%
  summarize(RMSE = 1-rmse(expected_tf, predicted_tf)) %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- rmse_cfrrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(RMSE)) %>%
  arrange(desc(Median))
plot_rmse <- rmse_cfrrbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))
plot_rmse$fraction_class <- factor(plot_rmse$fraction_class, levels = c("low","medium","high"))

p <- ggplot(plot_rmse, aes(
  x = reorder_within(deconv_tool,-RMSE, fraction_class, median),
  y = RMSE, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "RRBS-CL dataset",
    x = "",
    y = "1-RMSE"
  ) +
  facet_wrap(~fraction_class, scales = "free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_x_reordered() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0.5,1)
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("RMSE_RRBS-CL.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

We can now try to see how the reference-free methods perform at increasing sequencing depth

```{r warning=FALSE}
plot_rmse_refree <- subset(rmse_cfrrbs, rmse_cfrrbs$deconv_tool %in% c("RefFreeCellMix","MeDeCom"))
plot_rmse_refree$seq_depth <- factor(plot_rmse_refree$seq_depth, levels = c("2M", "5M", "10M", "15M", "20M"))

# Define comparisons
my_comparisons <- list(
  c("2M", "5M"),
  c("5M", "10M"),
  c("10M","15M"),
  c("10M", "20M"),
  c("15M", "20M")
)
# Plot
p <- ggplot(data = plot_rmse_refree, aes(
    x = seq_depth, y = RMSE, fill = seq_depth)
  ) +
  #geom_jitter(size=1.5, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  facet_wrap(~deconv_tool, ncol = 5) +
  stat_compare_means(
      method = "wilcox.test", comparisons = my_comparisons,
      p.adjust.method = "BH", label = "adj.p", hide.ns = F,                    
      method.args = list(alternative="less"),
      label.y = c(1, 1.05, 1.10, 1.15, 1.20), size = 3
  ) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major.y = element_line(color = "grey90"),
          panel.grid.minor.y = element_line(color = "grey90")) +
  labs(title = "RRBS-CL dataset",
       x = "", y = "1-RMSE") +
  scale_y_continuous(
    breaks = c(0.6, 0.8, 1) # only show these breaks
  ) +
  scale_fill_manual(values = c("2M"="#FDE725FF","5M"="#7AD151FF","10M"="#2A788EFF","15M"="#414487FF","20M"="#440154FF"))  +
  coord_cartesian(ylim=c(0.6,1.25))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("robustness_RMSE_seqdepth_refree_cfRRBS.pdf"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### LoD

And finally compute the LoD score

```{r warning=FALSE}
# Define the comparisons and an empty dataframe to store the results
my_comparisons <- list(
  c("0", "1e-04"), 
  c("0", "0.001"), 
  c("0", "0.003"), 
  c("0", "0.007"),
  c("0", "0.01"),
  c("0", "0.025"),
  c("0", "0.05"),
  c("0", "0.1"),
  c("0", "0.25"),
  c("0", "0.5")
)
lod_refree_cfrrbs <- data.frame()
in_silico_refree_cfrrbs <- subset(in_silico_cfrrbs, in_silico_cfrrbs$deconv_tool %in% c("MeDeCom","RefFreeCellMix"))

# Loop over the type of tumors
for (type in unique(in_silico_refree_cfrrbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_refree_cfrrbs, in_silico_refree_cfrrbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      filt_dmr <- subset(filt_depth, filt_depth$dmr_tool == dmrtool)
      for (deconv in unique(filt_depth$deconv_tool)) {
        
        # Filter over the deconvolution tool and run the statistical test to find the LoD
        filt_df <- subset(filt_dmr, filt_dmr$deconv_tool == deconv)
        stats <- compute_lod(filt_df, my_comparisons, correction.method = "BH")

        lod <- stats %>%
          arrange(as.numeric(group2)) %>%
          filter(as.numeric(p.adj) < 0.01) %>%
          slice(1) %>%
          pull(group2)
            
        # If lod is empty or NA, assign 1
        if (length(lod) == 0 || is.na(lod)) {
          lod <- 1
        }
        
        # Concatenate the results into the dataframe
        tmp <- data.frame(deconv_tool = deconv, dmr_tool = dmrtool, tumor_type = type, seq_depth = seqdepth, lod = lod, stringsAsFactors = F)
        lod_refree_cfrrbs <- rbind(lod_refree_cfrrbs, tmp)
      }
    }
  }
}
```


```{r}
lod_refbased_cfrrbs <- read.csv("../resources/lod_allstats.csv", row.names = 1)[1:5]
lod_refbased_cfrrbs <- subset(lod_refbased_cfrrbs, lod_refbased_cfrrbs$tumor_type %in% c("NBL"))
all.lods <- rbind(lod_refbased_cfrrbs, lod_refree_cfrrbs)  %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by median RMSE
median_rank <- all.lods %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(as.numeric(lod))) %>%
  arrange(Median)
plot_lods <- all.lods %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_lods, aes(
  x = deconv_tool, y = as.numeric(lod), 
  color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20") +
  labs(
    title = "RRBS-CL dataset",
    x = "",
    y = "LoD"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("LoD_RRBS-CL.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```


## All datasets aggregated

```{r}
bench <- subset(bench, bench$mixture_type == "in_silico")
```


### AUC-ROC values

We will first start computing and plotting the AUC-ROC values.


```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(bench$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(bench$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(bench, bench$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)
          
          # If the dataframe is empty, skip
          if (dim(filt_df)[1] == 0) { next }
          
          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_all <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(AUC=mean(AUC),.groups = "drop") %>%
    mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- auc_all %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(AUC)) %>%
  arrange(desc(Median))
plot_auc <- auc_all %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_auc, aes(x = deconv_tool, y = AUC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "All datasets",
    x = "",
    y = "AUC-ROC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
     "Reference-based" = "#009E73",
     "Reference-free" = "#E69F00"
   )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("AUC-ROC_all_datasets.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```

### SCC

Now, we compute the SCC. The SCC is computed for each combination pooling all the observed tumor fractions together to get a value for each combination. The SCC is then min-max scaled.

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_all <- bench %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop") %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- scc_all %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(SCC)) %>%
  arrange(desc(Median))
plot_scc <- scc_all %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_scc, aes(x = deconv_tool, y = SCC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "All datasets",
    x = "",
    y = "SCC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0,1)
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("SCC_all_datasets.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```

### RMSE 

Compute the RMSE.

```{r}
rmse_all <- bench %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, fraction_class) %>%
  summarize(RMSE = 1-rmse(expected_tf, predicted_tf)) %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- rmse_all %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(RMSE)) %>%
  arrange(desc(Median))
plot_rmse <- rmse_all %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))
plot_rmse$fraction_class <- factor(plot_rmse$fraction_class, levels = c("low","medium","high"))

p <- ggplot(plot_rmse, aes(
  x = reorder_within(deconv_tool,-RMSE, fraction_class, median),
  y = RMSE, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "All datasets",
    x = "",
    y = "1-RMSE"
  ) +
  facet_wrap(~fraction_class, scales = "free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_x_reordered() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0.5,1)
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("RMSE_per_fraction_all_datasets.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```


```{r}
# Rank the tools by mean RMSE
median_rank <- rmse_all %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(RMSE)) %>%
  arrange(desc(Median))
plot_rmse <- rmse_all %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_rmse, aes(
  x = deconv_tool,
  y = RMSE, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "All datasets",
    x = "",
    y = "1-RMSE"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0.5,1)
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("RMSE_aggregated_all_datasets.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```

### LoD

```{r}
my_comp <- list(
  c("0", "1e-04"), 
  c("0", "0.001"), 
  c("0", "0.003"), 
  c("0", "0.007"),
  c("0", "0.01"),
  c("0", "0.025")
)
filt_df <- subset(bench, bench$expected_tf %in% c("0", "1e-04","0.001", "0.003", "0.007","0.01","0.025") & bench$dmr_tool=="rrbs20-200" & bench$seq_depth=="310M" & bench$deconv_tool=="MeDeCom" & bench$tumor_type=="LUAD")
stats <- compute_lod(filt_df, my_comp, correction.method = "BH")
p <- filt_df %>%
  group_by(deconv_tool) %>%
  ggplot(aes(x = as.factor(expected_tf), y = predicted_tf)) +
  geom_boxplot() +
  geom_point(size = 1) +
  labs(
    title = paste(),
    x = "",
    y = "Estimated tumoral fraction (%)"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5)  # add horizontal grid
  ) + 
  stat_pvalue_manual(stats, label = "star", xmin = "group1", xmax = "group2", y.position = "y.position")
p
```


And finally compute the LoD score

```{r warning=FALSE}
# Define the comparisons and an empty dataframe to store the results
my_comparisons <- list(
  c("0", "1e-04"), 
  c("0", "0.001"), 
  c("0", "0.003"), 
  c("0", "0.007"),
  c("0", "0.01"),
  c("0", "0.025"),
  c("0", "0.05"),
  c("0", "0.1"),
  c("0", "0.25"),
  c("0", "0.5")
)
lod_refree_all <- data.frame()
in_silico_refree_all <- subset(bench, bench$deconv_tool %in% c("MeDeCom","RefFreeCellMix"))

# Loop over the type of tumors
for (type in unique(in_silico_refree_all$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_refree_all, in_silico_refree_all$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      filt_dmr <- subset(filt_depth, filt_depth$dmr_tool == dmrtool)
      for (deconv in unique(filt_depth$deconv_tool)) {
        
        # Filter over the deconvolution tool and run the statistical test to find the LoD
        filt_df <- subset(filt_dmr, filt_dmr$deconv_tool == deconv)
        stats <- compute_lod(filt_df, my_comparisons, correction.method = "BH")

        lod <- stats %>%
          arrange(as.numeric(group2)) %>%
          filter(as.numeric(p.adj) < 0.01) %>%
          slice(1) %>%
          pull(group2)
        
        # p <- filt_df %>%
        #   group_by(deconv_tool) %>%
        #   ggplot(aes(x = as.factor(expected_tf), y = predicted_tf)) +
        #   geom_boxplot() +
        #   labs(
        #     title = paste("LoD | Tool:", deconv,"| DMR tool:", dmrtool,"| Depth:",seqdepth, "| Tumor:",type),
        #     x = "",
        #     y = "Estimated tumoral fraction (%)"
        #   ) +
        #   theme_classic() +
        #   theme(
        #     axis.text.x = element_text(angle = 45, hjust = 1),
        #     panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5)  # add horizontal grid
        #   ) + 
        #   stat_pvalue_manual(stats, label = "star", xmin = "group1", xmax = "group2", y.position = "y.position")
            
        # If lod is empty or NA, assign 1
        if (length(lod) == 0 || is.na(lod)) {
          lod <- 1
        }
        
        # Concatenate the results into the dataframe
        tmp <- data.frame(deconv_tool = deconv, dmr_tool = dmrtool, tumor_type = type, seq_depth = seqdepth, lod = lod, stringsAsFactors = F)
        lod_refree_all <- rbind(lod_refree_all, tmp)
      }
    }
  }
}
```


```{r}
lod_refbased_all <- read.csv("../resources/lod_allstats.csv", row.names = 1)[1:5]
lod_refbased_all <- subset(lod_refbased_all, lod_refbased_all$tumor_type %in% c("BRCA","BLCA","LUAD","LUSC"))

all.lods <- rbind(lod_refbased_all, lod_refree_all)  %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- all.lods %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(as.numeric(lod))) %>%
  arrange(Median)
plot_lods <- all.lods %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_lods, aes(
  x = deconv_tool, y = as.numeric(lod), 
  color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20") +
  labs(
    title = "All datasets",
    x = "",
    y = "LoD"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("LoD_all_datasets.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```


### Predicted vs expected fractions

```{r}
# Extract numeric positions of factor levels
plot_lf <- subset(bench, bench$expected_tf < 0.05)  %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
levels_tf <- sort(unique(plot_lf$expected_tf))
p <- ggplot(plot_lf, aes(
  x = as.factor(expected_tf), y = predicted_tf,
  color = RefType, fill = RefType)) +
  geom_boxplot(outliers = F, median.colour = "grey20", color = "grey20") +
  geom_line(data = data.frame(
    x = seq_along(levels_tf),
    y = levels_tf
  ), aes(x = x, y = y), color = "red", linetype = "dashed", inherit.aes = FALSE) +
  facet_wrap(~deconv_tool) +
  labs(
    title = "All datasets (fractions <= 0.05)",
    x = "Expected tumoral fraction",
    y = "Predicted tumoral fraction"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0,0.1)
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("all_datasets_predicted_vs_expected_lf.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```


```{r}
# Extract numeric positions of factor levels
plot_hf <- subset(bench, bench$expected_tf >= 0.05)  %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
levels_tf <- sort(unique(plot_hf$expected_tf))
p <- ggplot(plot_hf, aes(
  x = as.factor(expected_tf), y = predicted_tf,
  color = RefType, fill = RefType)) +
  geom_boxplot(outliers = F, median.colour = "grey20", color = "grey20") +
  geom_line(data = data.frame(
    x = seq_along(levels_tf),
    y = levels_tf
  ), aes(x = x, y = y), color = "red", linetype = "dashed", inherit.aes = FALSE) +
  facet_wrap(~deconv_tool) +
  labs(
    title = "All datasets (fractions > 0.05)",
    x = "Expected tumoral fraction",
    y = "Predicted tumoral fraction"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("all_datasets_predicted_vs_expected_hf.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```

## Calibration analysis

The sections above characterise error magnitude (RMSE), rank concordance
(SCC) and detection sensitivity/specificity (AUC, LoD), but none of them
directly test whether a tool's estimates are **calibrated** -- i.e. whether
`predicted_tf` tracks `expected_tf` on a 1:1 basis, or whether a tool is
systematically biased (e.g. reference-free tools consistently overestimating
tumour fraction, as suggested by the boxplots above). We address this with
three explicit, per-tool calibration analyses pooling all three datasets
(WGBS-TT, RRBS-TT, RRBS-CL) together: (1) a calibration regression of
observed on expected fraction (slope and intercept), (2) the same regression
restricted to the low-fraction range where clinical decisions are most
sensitive to bias, and (3) Bland-Altman agreement plots.

### Calibration slope and intercept

For each tool we fit `predicted_tf ~ expected_tf` by ordinary least squares
across all samples (all datasets, tumor types and sequencing depths pooled).
A perfectly calibrated tool has slope = 1 and intercept = 0. A slope < 1
indicates compression towards the mean (underestimation at high fractions,
overestimation at low fractions); a positive intercept at low/zero fractions
indicates a systematic overestimation bias, consistent with the LoD
definition itself (the fraction at which predictions become distinguishable
from the zero-fraction baseline).

```{r}
calibration_fit <- function(df) {
  fit <- lm(predicted_tf ~ expected_tf, data = df)
  coefs <- coef(fit)
  ci <- tryCatch(confint(fit), error = function(e) matrix(NA, 2, 2))
  data.frame(
    Intercept = coefs[["(Intercept)"]],
    Intercept_lwr = ci[1, 1], Intercept_upr = ci[1, 2],
    Slope = coefs[["expected_tf"]],
    Slope_lwr = ci[2, 1], Slope_upr = ci[2, 2],
    R2 = summary(fit)$r.squared,
    n = nrow(df)
  )
}

calibration_stats <- bench %>%
  group_by(deconv_tool) %>%
  group_modify(~ calibration_fit(.x)) %>%
  ungroup() %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom", "RefFreeCellMix"), "Reference-free", "Reference-based")) %>%
  arrange(desc(Slope))

calibration_stats
```

```{r}
p <- ggplot(calibration_stats, aes(x = reorder(deconv_tool, Slope), y = Slope, color = RefType)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = Slope_lwr, ymax = Slope_upr), size = 0.6) +
  coord_flip() +
  labs(
    title = "Calibration slope (predicted_tf ~ expected_tf), all datasets pooled",
    subtitle = "Dashed line = perfect calibration (slope = 1)",
    x = "", y = "Calibration slope"
  ) +
  theme_classic() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```

```{r}
output_file <- file.path(output_dir, "calibration_slope_all_datasets.svg")
ggsave(output_file, plot = p, width = 7, height = 5)
```

```{r}
p <- ggplot(calibration_stats, aes(x = reorder(deconv_tool, Intercept), y = Intercept, color = RefType)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = Intercept_lwr, ymax = Intercept_upr), size = 0.6) +
  coord_flip() +
  labs(
    title = "Calibration intercept (predicted_tf ~ expected_tf), all datasets pooled",
    subtitle = "Dashed line = no bias at zero tumour fraction (intercept = 0)",
    x = "", y = "Calibration intercept"
  ) +
  theme_classic() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```

```{r}
output_file <- file.path(output_dir, "calibration_intercept_all_datasets.svg")
ggsave(output_file, plot = p, width = 7, height = 5)
```

### Calibration at high tumour fractions

Because clinical relevance concentrates at high tumour fractions, we repeat
the calibration regression restricted to `expected_tf >= 0.05`, which is
where a systematic overestimation bias is most consequential (it directly
inflates apparent sensitivity/LoD).

```{r}
calibration_stats_lf <- bench %>%
  filter(expected_tf >= 0.05) %>%
  group_by(deconv_tool) %>%
  group_modify(~ calibration_fit(.x)) %>%
  ungroup() %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom", "RefFreeCellMix"), "Reference-free", "Reference-based")) %>%
  arrange(desc(Intercept))

calibration_stats_lf
```

```{r}
p <- ggplot(calibration_stats_lf, aes(x = reorder(deconv_tool, Intercept), y = Intercept, color = RefType)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = Intercept_lwr, ymax = Intercept_upr), size = 0.6) +
  coord_flip() +
  labs(
    title = "Calibration intercept, high tumour fractions only (expected_tf >= 0.05)",
    subtitle = "Dashed line = no bias at zero tumour fraction (intercept = 0)",
    x = "", y = "Calibration intercept"
  ) +
  theme_classic() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```

```{r}
output_file <- file.path(output_dir, "calibration_intercept_high_fractions.svg")
ggsave(output_file, plot = p, width = 7, height = 5)
```


```{r}
p <- ggplot(calibration_stats_lf, aes(x = reorder(deconv_tool, Slope), y = Slope, color = RefType)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = Slope_lwr, ymax = Slope_upr), size = 0.6) +
  coord_flip() +
  labs(
    title = "Calibration slope, high tumour fractions only (expected_tf >= 0.05)",
    subtitle = "Dashed line = perfect calibration (slope = 1)",
    x = "", y = "Calibration slope"
  ) +
  theme_classic() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```

```{r}
output_file <- file.path(output_dir, "calibration_slope_high_fractions.svg")
ggsave(output_file, plot = p, width = 7, height = 5)
```

### Bland-Altman agreement plots

For each tool we plot the difference (`predicted_tf - expected_tf`) against
the mean of the two (`(predicted_tf + expected_tf) / 2`), with the mean bias
and the 95% limits of agreement (mean bias +/- 1.96 SD). A tool with a mean
bias consistently above zero is systematically overestimating tumour
fraction; funnel-shaped spread indicates the bias/variance is not constant
across the range of tumour fractions.

```{r fig.width=12, fig.height=9}
bland_altman_df <- bench %>%
  mutate(
    diff_tf = predicted_tf - expected_tf,
    mean_tf = (predicted_tf + expected_tf) / 2,
    RefType = ifelse(deconv_tool %in% c("MeDeCom", "RefFreeCellMix"), "Reference-free", "Reference-based")
  )

bland_altman_stats <- bland_altman_df %>%
  group_by(deconv_tool) %>%
  summarise(
    bias = mean(diff_tf),
    sd_diff = sd(diff_tf),
    loa_lower = bias - 1.96 * sd_diff,
    loa_upper = bias + 1.96 * sd_diff,
    .groups = "drop"
  )

bland_altman_stats
```

```{r fig.width=12, fig.height=9}
p <- ggplot(bland_altman_df, aes(x = mean_tf, y = diff_tf, color = RefType)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40") +
  geom_hline(data = bland_altman_stats, aes(yintercept = bias), color = "black", linewidth = 0.6) +
  geom_hline(data = bland_altman_stats, aes(yintercept = loa_lower), color = "black", linetype = "dashed", linewidth = 0.4) +
  geom_hline(data = bland_altman_stats, aes(yintercept = loa_upper), color = "black", linetype = "dashed", linewidth = 0.4) +
  facet_wrap(~deconv_tool, scales = "free_y") +
  labs(
    title = "Bland-Altman: predicted_tf - expected_tf vs. mean(predicted_tf, expected_tf)",
    subtitle = "Solid line = mean bias; dashed lines = 95% limits of agreement",
    x = "Mean of predicted and expected tumour fraction",
    y = "Predicted - expected tumour fraction"
  ) +
  theme_classic() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  theme(legend.position = "top")
p
```

```{r}
output_file <- file.path(output_dir, "bland_altman_all_tools.png")
ggsave(output_file, plot = p, width = 12, height = 9)
```

### Calibration summary table

```{r}
calibration_summary <- calibration_stats %>%
  select(deconv_tool, RefType, Slope, Intercept, R2) %>%
  inner_join(
    calibration_stats_lf %>% select(deconv_tool, Slope_lf = Slope, Intercept_lf = Intercept),
    by = "deconv_tool"
  ) %>%
  inner_join(bland_altman_stats, by = "deconv_tool") %>%
  arrange(desc(Intercept))

calibration_summary
```

```{r}
write.csv(calibration_summary, file.path(output_dir, "calibration_summary.csv"), row.names = FALSE)
```

### Post-hoc recalibration

The calibration slope/intercept above can be used to *correct* each tool's
predictions after the fact: inverting the fitted line
`predicted_tf = Intercept + Slope * expected_tf` gives
`predicted_tf_corrected = (predicted_tf - Intercept) / Slope`, a per-tool
linear recalibration fitted on the same pooled data used above. This is a
diagnostic device: it isolates how much of each tool's
disagreement with `expected_tf` is a *fixable, linear* bias, versus
irreducible scatter that a simple recalibration cannot remove. We apply the
pooled (all-datasets) slope/intercept from the "Calibration slope and
intercept" section above. Because it is fitted and evaluated on the same
data, it is an in-sample estimate; see "Cross-validated recalibration"
below for the out-of-sample version.

```{r}
calib_params <- calibration_stats %>% select(deconv_tool, Intercept, Slope)

bench_corrected <- bench %>%
  left_join(calib_params, by = "deconv_tool") %>%
  mutate(predicted_tf_corrected = (predicted_tf - Intercept) / Slope) %>%
  select(-Intercept, -Slope)
```

#### Predicted vs. expected, after correction

Recalibrated counterpart of the "Predicted vs expected fractions" plots
above (fractions < 0.05); compare directly against those raw-prediction
plots.

```{r fig.width=12, fig.height=9}
plot_calib_corrected <- bench_corrected %>%
  filter(expected_tf < 0.05) %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom", "RefFreeCellMix"), "Reference-free", "Reference-based"))

levels_tf <- sort(unique(plot_calib_corrected$expected_tf))

p <- ggplot(plot_calib_corrected, aes(
  x = as.factor(expected_tf), y = predicted_tf_corrected,
  color = RefType, fill = RefType)) +
  geom_boxplot(outliers = FALSE, median.colour = "grey20", color = "grey20") +
  geom_line(data = data.frame(
    x = seq_along(levels_tf),
    y = levels_tf
  ), aes(x = x, y = y), color = "red", linetype = "dashed", inherit.aes = FALSE) +
  facet_wrap(~deconv_tool) +
  labs(
    title = "Recalibrated predicted vs. expected fraction (fractions < 0.05)",
    x = "Expected tumoral fraction",
    y = "Predicted tumoral fraction (recalibrated)"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  coord_cartesian(ylim = c(0, 0.1))
p
```

```{r}
output_file <- file.path(output_dir, "calibration_correction_predicted_vs_expected_lf.svg")
ggsave(output_file, plot = p, width = 8, height = 4)
```

#### Bland-Altman, after correction

Recalibrated counterpart of the Bland-Altman plot above.

```{r fig.width=12, fig.height=9}
bland_altman_corrected <- bench_corrected %>%
  mutate(
    RefType = ifelse(deconv_tool %in% c("MeDeCom", "RefFreeCellMix"), "Reference-free", "Reference-based"),
    diff_tf = predicted_tf_corrected - expected_tf,
    mean_tf = (predicted_tf_corrected + expected_tf) / 2
  )

bland_altman_corrected_stats <- bland_altman_corrected %>%
  group_by(deconv_tool) %>%
  summarise(
    bias = mean(diff_tf),
    sd_diff = sd(diff_tf),
    loa_lower = bias - 1.96 * sd_diff,
    loa_upper = bias + 1.96 * sd_diff,
    .groups = "drop"
  )

bland_altman_corrected_stats
```

```{r fig.width=12, fig.height=9}
p <- ggplot(bland_altman_corrected, aes(x = mean_tf, y = diff_tf, color = RefType)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40") +
  geom_hline(data = bland_altman_corrected_stats, aes(yintercept = bias), color = "black", linewidth = 0.6) +
  geom_hline(data = bland_altman_corrected_stats, aes(yintercept = loa_lower), color = "black", linetype = "dashed", linewidth = 0.4) +
  geom_hline(data = bland_altman_corrected_stats, aes(yintercept = loa_upper), color = "black", linetype = "dashed", linewidth = 0.4) +
  facet_wrap(~deconv_tool, scales = "free_y") +
  labs(
    title = "Bland-Altman, recalibrated predictions",
    subtitle = "Solid line = mean bias; dashed lines = 95% limits of agreement",
    x = "Mean of predicted and expected tumour fraction",
    y = "Predicted (recalibrated) - expected tumour fraction"
  ) +
  theme_classic() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  theme(legend.position = "top")
p
```

```{r}
output_file <- file.path(output_dir, "bland_altman_corrected.png")
ggsave(output_file, plot = p, width = 12, height = 9)
```

#### RMSE, before and after correction

Since RMSE is the metric most directly affected by a fixable linear bias, we
also report it before and after recalibration (pooled across datasets),
which quantifies how much of each tool's error was attributable to
calibration bias rather than irreducible scatter.

```{r}
rmse_compare <- bench_corrected %>%
  filter(expected_tf != 0) %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom", "RefFreeCellMix"), "Reference-free", "Reference-based")) %>%
  group_by(deconv_tool, RefType) %>%
  summarise(
    RMSE_raw = sqrt(mean((expected_tf - predicted_tf)^2)),
    RMSE_corrected = sqrt(mean((expected_tf - predicted_tf_corrected)^2)),
    .groups = "drop"
  ) %>%
  mutate(RMSE_reduction_pct = 100 * (RMSE_raw - RMSE_corrected) / RMSE_raw) %>%
  arrange(desc(RMSE_reduction_pct))

rmse_compare
```

```{r}
write.csv(rmse_compare, file.path(output_dir, "rmse_raw_vs_corrected.csv"), row.names = FALSE)
```

```{r fig.width=8, fig.height=6}
rmse_compare_plot <- rmse_compare %>%
  mutate(deconv_tool = factor(deconv_tool, levels = deconv_tool[order(RMSE_raw)]))

p <- ggplot(rmse_compare_plot) +
  geom_segment(aes(x = deconv_tool, xend = deconv_tool, y = RMSE_raw, yend = RMSE_corrected),
               color = "grey60") +
  geom_point(aes(x = deconv_tool, y = RMSE_raw, color = RefType), shape = 1, size = 3, stroke = 1.2) +
  geom_point(aes(x = deconv_tool, y = RMSE_corrected, color = RefType), shape = 16, size = 3) +
  coord_flip() +
  labs(
    title = "RMSE before (open circle) and after (filled circle) recalibration",
    x = "", y = "RMSE"
  ) +
  theme_classic() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```

```{r}
output_file <- file.path(output_dir, "rmse_raw_vs_corrected.svg")
ggsave(output_file, plot = p, width = 8, height = 6)
```

#### Cross-validated recalibration

The recalibration above is fitted and evaluated on the same data, so its
RMSE reduction is an in-sample estimate. To test whether the bias is
correctable on unseen data, we fit each tool's calibration line on part
of the data and apply it to the held-out part, using two grouped
cross-validation schemes:

- **Leave-one-tumor-type-out (LOTO)**: 8 folds; all samples (all
  sequencing depths) of one tumor type are held out together, so no
  material from the held-out tumor type is used for fitting.
- **Leave-one-dataset-out (LODO)**: 3 folds (WGBS-TT, RRBS-TT, RRBS-CL);
  tests whether a correction learned on other datasets transfers to a
  new one.

Each tumor type occurs in only one dataset. RMSE is computed on the
pooled out-of-fold predictions, excluding `expected_tf = 0` as above.

```{r}
cv_recalibrate <- function(df, fold_col) {
  folds <- unique(df[[fold_col]])
  bind_rows(lapply(folds, function(f) {
    train <- df[df[[fold_col]] != f, ]
    test  <- df[df[[fold_col]] == f, ]
    bind_rows(lapply(split(train, train$deconv_tool), function(tr) {
      fit <- coef(lm(predicted_tf ~ expected_tf, data = tr))
      test[test$deconv_tool == tr$deconv_tool[1], ] %>%
        mutate(predicted_tf_cv = (predicted_tf - fit[[1]]) / fit[[2]])
    })) %>%
      mutate(fold = f)
  }))
}

bench_cv <- bind_rows(
  cv_recalibrate(bench_corrected, "tumor_type") %>% mutate(cv_scheme = "LOTO"),
  cv_recalibrate(bench_corrected, "seq_method") %>% mutate(cv_scheme = "LODO")
)
stopifnot(nrow(bench_cv) == 2 * nrow(bench_corrected))
```

```{r}
rmse_cv <- bench_cv %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, cv_scheme) %>%
  summarise(RMSE_cv = sqrt(mean((expected_tf - predicted_tf_cv)^2)), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = cv_scheme, values_from = RMSE_cv, names_prefix = "RMSE_cv_")

# Share of held-out folds in which recalibration lowered the RMSE.
fold_improvement <- bench_cv %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, cv_scheme, fold) %>%
  summarise(improved = sqrt(mean((expected_tf - predicted_tf_cv)^2)) <
                       sqrt(mean((expected_tf - predicted_tf)^2)),
            .groups = "drop") %>%
  group_by(deconv_tool, cv_scheme) %>%
  summarise(folds_improved = sprintf("%d/%d", sum(improved), n()), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = cv_scheme, values_from = folds_improved, names_prefix = "folds_improved_")

rmse_compare_cv <- rmse_compare %>%
  select(deconv_tool, RefType, RMSE_raw, RMSE_corrected) %>%
  rename(RMSE_insample = RMSE_corrected) %>%
  inner_join(rmse_cv, by = "deconv_tool") %>%
  inner_join(fold_improvement, by = "deconv_tool") %>%
  mutate(
    reduction_insample_pct = 100 * (RMSE_raw - RMSE_insample) / RMSE_raw,
    reduction_LOTO_pct     = 100 * (RMSE_raw - RMSE_cv_LOTO) / RMSE_raw,
    reduction_LODO_pct     = 100 * (RMSE_raw - RMSE_cv_LODO) / RMSE_raw
  ) %>%
  arrange(desc(reduction_LOTO_pct))

rmse_compare_cv
```

```{r}
write.csv(rmse_compare_cv, file.path(output_dir, "rmse_raw_vs_corrected_cv.csv"), row.names = FALSE)
```

```{r fig.width=8, fig.height=6}
rmse_cv_plot <- rmse_compare_cv %>%
  select(deconv_tool, RefType, RMSE_raw, RMSE_insample, RMSE_cv_LOTO, RMSE_cv_LODO) %>%
  tidyr::pivot_longer(starts_with("RMSE_"), names_to = "estimate", values_to = "RMSE") %>%
  mutate(
    estimate = factor(estimate,
                      levels = c("RMSE_raw", "RMSE_insample", "RMSE_cv_LOTO", "RMSE_cv_LODO"),
                      labels = c("Raw", "Recalibrated (in-sample)",
                                 "Recalibrated (CV, leave-one-tumor-type-out)",
                                 "Recalibrated (CV, leave-one-dataset-out)")),
    deconv_tool = factor(deconv_tool, levels = rmse_compare$deconv_tool[order(rmse_compare$RMSE_raw)])
  )

p <- ggplot(rmse_cv_plot, aes(x = deconv_tool, y = RMSE)) +
  geom_line(aes(group = deconv_tool), color = "grey75") +
  geom_point(aes(shape = estimate, color = RefType), size = 2.8, stroke = 1.1) +
  scale_shape_manual(values = c(1, 16, 17, 15), name = NULL) +
  coord_flip() +
  labs(title = "RMSE before and after recalibration: in-sample vs. cross-validated",
       x = "", y = "RMSE") +
  theme_classic() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ), name = NULL) +
  guides(shape = guide_legend(nrow = 2)) +
  theme(legend.position = "bottom", legend.box = "vertical")
p
```

```{r}
output_file <- file.path(output_dir, "rmse_raw_vs_corrected_cv.svg")
ggsave(output_file, plot = p, width = 8, height = 6.5)
```
