---
title: "04. Analysis of reference free tools"
author: "Edoardo Giuili"
date: "2025-09-24"
output: html_document
---

## Introduction

In this tutorial we want to show how to compare the performance of the reference-free methods with the performance of the reference-based methods. Specifically, we want to see whether the former ones out- or underperform the latter.

### Setting up the environment

First, set up the main directory

```{r setup}
# Don't change this if you git cloned the entire GitHub repository and you are working in the notebooks/ folder
knitr::opts_knit$set(root.dir = './')
```

Import the necessary libraries for the analysis

```{r message=FALSE, warning=FALSE}
library(ggpubr)
library(ggplot2)
library(dplyr)
library(pROC)
library(tidytext)
```

Import the useful functions and the metrics

```{r}
source("../R/compute_auc.R")
source("../R/compute_lod.R")
# RMSE
rmse <- function(actual, predicted) {
  round(sqrt(mean((actual - predicted)^2)),4)
}
# Spearman's rank correlation coefficient (SCC)
scc <- function(actual, predicted) {
  cor(actual, predicted, method = "spearman", use = "pairwise.complete.obs")
}
```

Read the results generated by both reference-free and reference-based tools and concatenate them in a unique dataframe.

```{r}
bench_refree <- read.csv("../resources/refree_final_dataset.csv", row.names = 1)
bench_refree <- bench_refree %>% 
  filter(!is.na(bench_refree$predicted_tf)) %>%
  mutate(fraction_class = ifelse( expected_tf <= 0.025, "low", ifelse( expected_tf >0.025 & expected_tf <= 0.1, "medium", "high")))
bench_refbased <- read.csv("../resources/final_benchmarking_dataset.csv")
bench_refbased <- bench_refbased %>% 
  filter(!is.na(bench_refbased$predicted_tf))  %>%
  mutate(fraction_class = ifelse( expected_tf <= 0.025, "low", ifelse( expected_tf >0.025 & expected_tf <= 0.1, "medium", "high")))
names(bench_refree)[5] <- "dmr_tool"
bench <- rbind(bench_refbased, bench_refree)
```

Select the WGBS and RRBS datasets

```{r}
in_silico_wgbs <- subset(bench, bench$seq_method == "wgbs")
in_silico_rrbs <- subset(bench, bench$seq_method == "rrbs")
in_silico_cfrrbs <- subset(bench, bench$seq_method == "cfrrbs" & bench$mixture_type == "in_silico")
```


## WGBS-TT

As always, we will start our comparison analysing the WGBS-TT dataset, to then proceed with the RRBS-TT and RRBS-CL datasets. 

### AUC-ROC values

We will first start computing and plotting the AUC-ROC values.


```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(in_silico_wgbs$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(in_silico_wgbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_wgbs, in_silico_wgbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)
          
          # If the dataframe is empty, skip
          if (dim(filt_df)[1] == 0) { next }
          
          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_wgbs <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(AUC=mean(AUC),.groups = "drop") %>%
    mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```

```{r}
# Rank the tools by mean RMSE
median_rank <- auc_wgbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(AUC)) %>%
  arrange(desc(Median))
plot_auc <- auc_wgbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_auc, aes(x = deconv_tool, y = AUC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "WGBS-TT dataset",
    x = "",
    y = "AUC-ROC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
     "Reference-based" = "#009E73",
     "Reference-free" = "#E69F00"
   )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0,1)
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("AUC-ROC_WGBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### SCC

Now, we compute the SCC. The SCC is computed for each combination pooling all the observed tumor fractions together to get a value for each combination. The SCC is then min-max scaled.

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_wgbs <- in_silico_wgbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop") %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- scc_wgbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(SCC)) %>%
  arrange(desc(Median))
plot_scc <- scc_wgbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_scc, aes(x = deconv_tool, y = SCC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "WGBS-TT dataset",
    x = "",
    y = "SCC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0,1)
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("SCC_WGBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### RMSE 

Compute the RMSE.

```{r}
rmse_wgbs <- in_silico_wgbs %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, fraction_class) %>%
  summarize(RMSE = 1-rmse(expected_tf, predicted_tf)) %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- rmse_wgbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(RMSE)) %>%
  arrange(desc(Median))
plot_rmse <- rmse_wgbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))
plot_rmse$fraction_class <- factor(plot_rmse$fraction_class, levels = c("low","medium","high"))

p <- ggplot(plot_rmse, aes(
  x = reorder_within(deconv_tool,-RMSE, fraction_class, median),
  y = RMSE, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "WGBS-TT dataset",
    x = "",
    y = "1-RMSE"
  ) +
  facet_wrap(~fraction_class, scales = "free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_x_reordered() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0.5,1)
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("RMSE_WGBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```



### LoD

```{r}
my_comp <- list(
  c("0", "1e-04"), 
  c("0", "0.001"), 
  c("0", "0.003"), 
  c("0", "0.007"),
  c("0", "0.01"),
  c("0", "0.025")
)
filt_df <- subset(in_silico_wgbs, in_silico_wgbs$expected_tf %in% c("0", "1e-04","0.001", "0.003", "0.007","0.01","0.025") & in_silico_wgbs$dmr_tool=="rrbs20-200" & in_silico_wgbs$seq_depth=="310M" & in_silico_wgbs$deconv_tool=="MeDeCom" & in_silico_wgbs$tumor_type=="LUAD")
stats <- compute_lod(filt_df, my_comp, correction.method = "BH")
p <- filt_df %>%
  group_by(deconv_tool) %>%
  ggplot(aes(x = as.factor(expected_tf), y = predicted_tf)) +
  geom_boxplot() +
  geom_point(size = 1) +
  labs(
    title = paste(),
    x = "",
    y = "Estimated tumoral fraction (%)"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5)  # add horizontal grid
  ) + 
  stat_pvalue_manual(stats, label = "star", xmin = "group1", xmax = "group2", y.position = "y.position")
p
```


And finally compute the LoD score

```{r warning=FALSE}
# Define the comparisons and an empty dataframe to store the results
my_comparisons <- list(
  c("0", "1e-04"), 
  c("0", "0.001"), 
  c("0", "0.003"), 
  c("0", "0.007"),
  c("0", "0.01"),
  c("0", "0.025"),
  c("0", "0.05"),
  c("0", "0.1"),
  c("0", "0.25"),
  c("0", "0.5")
)
lod_refree_wgbs <- data.frame()
in_silico_refree_wgbs <- subset(in_silico_wgbs, in_silico_wgbs$deconv_tool %in% c("MeDeCom","RefFreeCellMix"))

# Loop over the type of tumors
for (type in unique(in_silico_refree_wgbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_refree_wgbs, in_silico_refree_wgbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      filt_dmr <- subset(filt_depth, filt_depth$dmr_tool == dmrtool)
      for (deconv in unique(filt_depth$deconv_tool)) {
        
        # Filter over the deconvolution tool and run the statistical test to find the LoD
        filt_df <- subset(filt_dmr, filt_dmr$deconv_tool == deconv)
        stats <- compute_lod(filt_df, my_comparisons, correction.method = "BH")

        lod <- stats %>%
          arrange(as.numeric(group2)) %>%
          filter(as.numeric(p.adj) < 0.01) %>%
          slice(1) %>%
          pull(group2)
        
        # p <- filt_df %>%
        #   group_by(deconv_tool) %>%
        #   ggplot(aes(x = as.factor(expected_tf), y = predicted_tf)) +
        #   geom_boxplot() +
        #   labs(
        #     title = paste("LoD | Tool:", deconv,"| DMR tool:", dmrtool,"| Depth:",seqdepth, "| Tumor:",type),
        #     x = "",
        #     y = "Estimated tumoral fraction (%)"
        #   ) +
        #   theme_classic() +
        #   theme(
        #     axis.text.x = element_text(angle = 45, hjust = 1),
        #     panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5)  # add horizontal grid
        #   ) + 
        #   stat_pvalue_manual(stats, label = "star", xmin = "group1", xmax = "group2", y.position = "y.position")
            
        # If lod is empty or NA, assign 1
        if (length(lod) == 0 || is.na(lod)) {
          lod <- 1
        }
        
        # Concatenate the results into the dataframe
        tmp <- data.frame(deconv_tool = deconv, dmr_tool = dmrtool, tumor_type = type, seq_depth = seqdepth, lod = lod, stringsAsFactors = F)
        lod_refree_wgbs <- rbind(lod_refree_wgbs, tmp)
      }
    }
  }
}
```


```{r}
lod_refbased_wgbs <- read.csv("../resources/lod_allstats.csv", row.names = 1)[1:5]
lod_refbased_wgbs <- subset(lod_refbased_wgbs, lod_refbased_wgbs$tumor_type %in% c("BRCA","BLCA","LUAD","LUSC"))

all.lods <- rbind(lod_refbased_wgbs, lod_refree_wgbs)  %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- all.lods %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(as.numeric(lod))) %>%
  arrange(Median)
plot_lods <- all.lods %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_lods, aes(
  x = deconv_tool, y = as.numeric(lod), 
  color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20") +
  labs(
    title = "WGBS-TT dataset",
    x = "",
    y = "LoD"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("LoD_WGBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```


## RRBS-TT

We can now proceed with the RRBS-TT dataset.

### AUC-ROC values

We will first start computing and plotting the AUC-ROC values.


```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(in_silico_rrbs$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(in_silico_rrbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_rrbs, in_silico_rrbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)
          
          # If the dataframe is empty, skip
          if (dim(filt_df)[1] == 0) { next }
          
          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_rrbs <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(AUC=mean(AUC),.groups = "drop") %>%
    mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```

```{r}
# Rank the tools by mean RMSE
median_rank <- auc_rrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(AUC)) %>%
  arrange(desc(Median))
plot_auc <- auc_rrbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_auc, aes(x = deconv_tool, y = AUC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "RRBS-TT dataset",
    x = "",
    y = "AUC-ROC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
     "Reference-based" = "#009E73",
     "Reference-free" = "#E69F00"
   )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("AUC-ROC_RRBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### SCC

Now, we compute the SCC.

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_rrbs <- in_silico_rrbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop") %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- scc_rrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(SCC)) %>%
  arrange(desc(Median))
plot_scc <- scc_rrbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_scc, aes(x = deconv_tool, y = SCC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "RRBS-TT dataset",
    x = "",
    y = "SCC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0,1)
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("SCC_RRBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### RMSE 

Compute the RMSE.

```{r}
rmse_rrbs <- in_silico_rrbs %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, fraction_class) %>%
  summarize(RMSE = 1-rmse(expected_tf, predicted_tf)) %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- rmse_rrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(RMSE)) %>%
  arrange(desc(Median))
plot_rmse <- rmse_rrbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))
plot_rmse$fraction_class <- factor(plot_rmse$fraction_class, levels = c("low","medium","high"))

p <- ggplot(plot_rmse, aes(
  x = reorder_within(deconv_tool,-RMSE, fraction_class, median),
  y = RMSE, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "RRBS-TT dataset",
    x = "",
    y = "1-RMSE"
  ) +
  facet_wrap(~fraction_class, scales = "free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_x_reordered() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0.5,1)
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("RMSE_RRBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### LoD

And finally compute the LoD score

```{r warning=FALSE}
# Define the comparisons and an empty dataframe to store the results
my_comparisons <- list(
  c("0", "1e-04"), 
  c("0", "0.001"), 
  c("0", "0.003"), 
  c("0", "0.007"),
  c("0", "0.01"),
  c("0", "0.025"),
  c("0", "0.05"),
  c("0", "0.1"),
  c("0", "0.25"),
  c("0", "0.5")
)
lod_refree_rrbs <- data.frame()
in_silico_refree_rrbs <- subset(in_silico_rrbs, in_silico_rrbs$deconv_tool %in% c("MeDeCom","RefFreeCellMix"))

# Loop over the type of tumors
for (type in unique(in_silico_refree_rrbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_refree_rrbs, in_silico_refree_rrbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      filt_dmr <- subset(filt_depth, filt_depth$dmr_tool == dmrtool)
      for (deconv in unique(filt_depth$deconv_tool)) {
        
        # Filter over the deconvolution tool and run the statistical test to find the LoD
        filt_df <- subset(filt_dmr, filt_dmr$deconv_tool == deconv)
        stats <- compute_lod(filt_df, my_comparisons, correction.method = "BH")

        lod <- stats %>%
          arrange(as.numeric(group2)) %>%
          filter(as.numeric(p.adj) < 0.01) %>%
          slice(1) %>%
          pull(group2)
            
        # If lod is empty or NA, assign 1
        if (length(lod) == 0 || is.na(lod)) {
          lod <- 1
        }
        
        # Concatenate the results into the dataframe
        tmp <- data.frame(deconv_tool = deconv, dmr_tool = dmrtool, tumor_type = type, seq_depth = seqdepth, lod = lod, stringsAsFactors = F)
        lod_refree_rrbs <- rbind(lod_refree_rrbs, tmp)
      }
    }
  }
}
```


```{r}
lod_refbased_rrbs <- read.csv("../resources/lod_allstats.csv", row.names = 1)[1:5]
lod_refbased_rrbs <- subset(lod_refbased_rrbs, lod_refbased_rrbs$tumor_type %in% c("CRC","LC","PC"))
all.lods <- rbind(lod_refbased_rrbs, lod_refree_rrbs)  %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- all.lods %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(as.numeric(lod))) %>%
  arrange(Median)
plot_lods <- all.lods %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_lods, aes(
  x = deconv_tool, y = as.numeric(lod), 
  color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20") +
  labs(
    title = "RRBS-TT dataset",
    x = "",
    y = "LoD"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("LoD_RRBS-TT.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

## RRBS-CL

We can now proceed with the RRBS-CL dataset.

### AUC-ROC values

We will first start computing and plotting the AUC-ROC values.


```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(in_silico_cfrrbs$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(in_silico_cfrrbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_cfrrbs, in_silico_cfrrbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)
          
          # If the dataframe is empty, skip
          if (dim(filt_df)[1] == 0) { next }
          
          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_cfrrbs <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(AUC=mean(AUC),.groups = "drop") %>%
    mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```

```{r}
# Rank the tools by mean RMSE
median_rank <- auc_cfrrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(AUC)) %>%
  arrange(desc(Median))
plot_auc <- auc_cfrrbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_auc, aes(x = deconv_tool, y = AUC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "RRBS-CL dataset",
    x = "",
    y = "AUC-ROC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
     "Reference-based" = "#009E73",
     "Reference-free" = "#E69F00"
   )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("AUC-ROC_RRBS-CL.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### SCC

Now, we compute the SCC.

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_cfrrbs <- in_silico_cfrrbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop") %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- scc_cfrrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(SCC)) %>%
  arrange(desc(Median))
plot_scc <- scc_cfrrbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_scc, aes(x = deconv_tool, y = SCC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "RRBS-CL dataset",
    x = "",
    y = "SCC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0,1)
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("SCC_RRBS-CL.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### RMSE 

Compute the RMSE.

```{r}
rmse_cfrrbs <- in_silico_cfrrbs %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, fraction_class) %>%
  summarize(RMSE = 1-rmse(expected_tf, predicted_tf)) %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- rmse_cfrrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(RMSE)) %>%
  arrange(desc(Median))
plot_rmse <- rmse_cfrrbs %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))
plot_rmse$fraction_class <- factor(plot_rmse$fraction_class, levels = c("low","medium","high"))

p <- ggplot(plot_rmse, aes(
  x = reorder_within(deconv_tool,-RMSE, fraction_class, median),
  y = RMSE, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "RRBS-CL dataset",
    x = "",
    y = "1-RMSE"
  ) +
  facet_wrap(~fraction_class, scales = "free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_x_reordered() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0.5,1)
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("RMSE_RRBS-CL.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```

### LoD

And finally compute the LoD score

```{r warning=FALSE}
# Define the comparisons and an empty dataframe to store the results
my_comparisons <- list(
  c("0", "1e-04"), 
  c("0", "0.001"), 
  c("0", "0.003"), 
  c("0", "0.007"),
  c("0", "0.01"),
  c("0", "0.025"),
  c("0", "0.05"),
  c("0", "0.1"),
  c("0", "0.25"),
  c("0", "0.5")
)
lod_refree_cfrrbs <- data.frame()
in_silico_refree_cfrrbs <- subset(in_silico_cfrrbs, in_silico_cfrrbs$deconv_tool %in% c("MeDeCom","RefFreeCellMix"))

# Loop over the type of tumors
for (type in unique(in_silico_refree_cfrrbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_refree_cfrrbs, in_silico_refree_cfrrbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      filt_dmr <- subset(filt_depth, filt_depth$dmr_tool == dmrtool)
      for (deconv in unique(filt_depth$deconv_tool)) {
        
        # Filter over the deconvolution tool and run the statistical test to find the LoD
        filt_df <- subset(filt_dmr, filt_dmr$deconv_tool == deconv)
        stats <- compute_lod(filt_df, my_comparisons, correction.method = "BH")

        lod <- stats %>%
          arrange(as.numeric(group2)) %>%
          filter(as.numeric(p.adj) < 0.01) %>%
          slice(1) %>%
          pull(group2)
            
        # If lod is empty or NA, assign 1
        if (length(lod) == 0 || is.na(lod)) {
          lod <- 1
        }
        
        # Concatenate the results into the dataframe
        tmp <- data.frame(deconv_tool = deconv, dmr_tool = dmrtool, tumor_type = type, seq_depth = seqdepth, lod = lod, stringsAsFactors = F)
        lod_refree_cfrrbs <- rbind(lod_refree_cfrrbs, tmp)
      }
    }
  }
}
```


```{r}
lod_refbased_cfrrbs <- read.csv("../resources/lod_allstats.csv", row.names = 1)[1:5]
lod_refbased_cfrrbs <- subset(lod_refbased_cfrrbs, lod_refbased_cfrrbs$tumor_type %in% c("NBL"))
all.lods <- rbind(lod_refbased_cfrrbs, lod_refree_cfrrbs)  %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by median RMSE
median_rank <- all.lods %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(as.numeric(lod))) %>%
  arrange(Median)
plot_lods <- all.lods %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_lods, aes(
  x = deconv_tool, y = as.numeric(lod), 
  color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20") +
  labs(
    title = "RRBS-CL dataset",
    x = "",
    y = "LoD"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("LoD_RRBS-CL.svg"))
ggsave(output_file, plot = p, width = 8, height = 6)
```


## All datasets aggregated

```{r}
bench <- subset(bench, bench$mixture_type == "in_silico")
```


### AUC-ROC values

We will first start computing and plotting the AUC-ROC values.


```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(bench$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(bench$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(bench, bench$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)
          
          # If the dataframe is empty, skip
          if (dim(filt_df)[1] == 0) { next }
          
          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_all <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(AUC=mean(AUC),.groups = "drop") %>%
    mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- auc_all %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(AUC)) %>%
  arrange(desc(Median))
plot_auc <- auc_all %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_auc, aes(x = deconv_tool, y = AUC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "All datasets",
    x = "",
    y = "AUC-ROC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
     "Reference-based" = "#009E73",
     "Reference-free" = "#E69F00"
   )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("AUC-ROC_all_datasets.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```

### SCC

Now, we compute the SCC. The SCC is computed for each combination pooling all the observed tumor fractions together to get a value for each combination. The SCC is then min-max scaled.

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_all <- bench %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop") %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- scc_all %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(SCC)) %>%
  arrange(desc(Median))
plot_scc <- scc_all %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_scc, aes(x = deconv_tool, y = SCC, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "All datasets",
    x = "",
    y = "SCC value"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0,1)
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("SCC_all_datasets.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```

### RMSE 

Compute the RMSE.

```{r}
rmse_all <- bench %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, fraction_class) %>%
  summarize(RMSE = 1-rmse(expected_tf, predicted_tf)) %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- rmse_all %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(RMSE)) %>%
  arrange(desc(Median))
plot_rmse <- rmse_all %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))
plot_rmse$fraction_class <- factor(plot_rmse$fraction_class, levels = c("low","medium","high"))

p <- ggplot(plot_rmse, aes(
  x = reorder_within(deconv_tool,-RMSE, fraction_class, median),
  y = RMSE, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "All datasets",
    x = "",
    y = "1-RMSE"
  ) +
  facet_wrap(~fraction_class, scales = "free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_x_reordered() +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0.5,1)
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("RMSE_per_fraction_all_datasets.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```


```{r}
# Rank the tools by mean RMSE
median_rank <- rmse_all %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(RMSE)) %>%
  arrange(desc(Median))
plot_rmse <- rmse_all %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_rmse, aes(
  x = deconv_tool,
  y = RMSE, color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20",) +
  labs(
    title = "All datasets",
    x = "",
    y = "1-RMSE"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0.5,1)
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("RMSE_aggregated_all_datasets.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```

### LoD

```{r}
my_comp <- list(
  c("0", "1e-04"), 
  c("0", "0.001"), 
  c("0", "0.003"), 
  c("0", "0.007"),
  c("0", "0.01"),
  c("0", "0.025")
)
filt_df <- subset(bench, bench$expected_tf %in% c("0", "1e-04","0.001", "0.003", "0.007","0.01","0.025") & bench$dmr_tool=="rrbs20-200" & bench$seq_depth=="310M" & bench$deconv_tool=="MeDeCom" & bench$tumor_type=="LUAD")
stats <- compute_lod(filt_df, my_comp, correction.method = "BH")
p <- filt_df %>%
  group_by(deconv_tool) %>%
  ggplot(aes(x = as.factor(expected_tf), y = predicted_tf)) +
  geom_boxplot() +
  geom_point(size = 1) +
  labs(
    title = paste(),
    x = "",
    y = "Estimated tumoral fraction (%)"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5)  # add horizontal grid
  ) + 
  stat_pvalue_manual(stats, label = "star", xmin = "group1", xmax = "group2", y.position = "y.position")
p
```


And finally compute the LoD score

```{r warning=FALSE}
# Define the comparisons and an empty dataframe to store the results
my_comparisons <- list(
  c("0", "1e-04"), 
  c("0", "0.001"), 
  c("0", "0.003"), 
  c("0", "0.007"),
  c("0", "0.01"),
  c("0", "0.025"),
  c("0", "0.05"),
  c("0", "0.1"),
  c("0", "0.25"),
  c("0", "0.5")
)
lod_refree_all <- data.frame()
in_silico_refree_all <- subset(bench, bench$deconv_tool %in% c("MeDeCom","RefFreeCellMix"))

# Loop over the type of tumors
for (type in unique(in_silico_refree_all$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_refree_all, in_silico_refree_all$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      filt_dmr <- subset(filt_depth, filt_depth$dmr_tool == dmrtool)
      for (deconv in unique(filt_depth$deconv_tool)) {
        
        # Filter over the deconvolution tool and run the statistical test to find the LoD
        filt_df <- subset(filt_dmr, filt_dmr$deconv_tool == deconv)
        stats <- compute_lod(filt_df, my_comparisons, correction.method = "BH")

        lod <- stats %>%
          arrange(as.numeric(group2)) %>%
          filter(as.numeric(p.adj) < 0.01) %>%
          slice(1) %>%
          pull(group2)
        
        # p <- filt_df %>%
        #   group_by(deconv_tool) %>%
        #   ggplot(aes(x = as.factor(expected_tf), y = predicted_tf)) +
        #   geom_boxplot() +
        #   labs(
        #     title = paste("LoD | Tool:", deconv,"| DMR tool:", dmrtool,"| Depth:",seqdepth, "| Tumor:",type),
        #     x = "",
        #     y = "Estimated tumoral fraction (%)"
        #   ) +
        #   theme_classic() +
        #   theme(
        #     axis.text.x = element_text(angle = 45, hjust = 1),
        #     panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5)  # add horizontal grid
        #   ) + 
        #   stat_pvalue_manual(stats, label = "star", xmin = "group1", xmax = "group2", y.position = "y.position")
            
        # If lod is empty or NA, assign 1
        if (length(lod) == 0 || is.na(lod)) {
          lod <- 1
        }
        
        # Concatenate the results into the dataframe
        tmp <- data.frame(deconv_tool = deconv, dmr_tool = dmrtool, tumor_type = type, seq_depth = seqdepth, lod = lod, stringsAsFactors = F)
        lod_refree_all <- rbind(lod_refree_all, tmp)
      }
    }
  }
}
```


```{r}
lod_refbased_all <- read.csv("../resources/lod_allstats.csv", row.names = 1)[1:5]
lod_refbased_all <- subset(lod_refbased_all, lod_refbased_all$tumor_type %in% c("BRCA","BLCA","LUAD","LUSC"))

all.lods <- rbind(lod_refbased_all, lod_refree_all)  %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
```


```{r}
# Rank the tools by mean RMSE
median_rank <- all.lods %>% 
  group_by(deconv_tool) %>% 
  summarise(Median = median(as.numeric(lod))) %>%
  arrange(Median)
plot_lods <- all.lods %>%
      mutate(deconv_tool = factor(deconv_tool, levels = median_rank$deconv_tool))

p <- ggplot(plot_lods, aes(
  x = deconv_tool, y = as.numeric(lod), 
  color = RefType, fill = RefType)) +
  geom_jitter(aes(color=RefType), size=1.5, alpha=0.8, width = 0.3) +
  geom_boxplot(alpha = 0.8, outliers = F, median.colour = "grey20",color = "grey20") +
  labs(
    title = "All datasets",
    x = "",
    y = "LoD"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("LoD_all_datasets.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```


### Predicted vs expected fractions

```{r}
# Extract numeric positions of factor levels
plot_lf <- subset(bench, bench$expected_tf < 0.05)  %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
levels_tf <- sort(unique(plot_lf$expected_tf))
p <- ggplot(plot_lf, aes(
  x = as.factor(expected_tf), y = predicted_tf,
  color = RefType, fill = RefType)) +
  geom_boxplot(outliers = F, median.colour = "grey20", color = "grey20") +
  geom_line(data = data.frame(
    x = seq_along(levels_tf),
    y = levels_tf
  ), aes(x = x, y = y), color = "red", linetype = "dashed", inherit.aes = FALSE) +
  facet_wrap(~deconv_tool) +
  labs(
    title = "All datasets (fractions <= 0.05)",
    x = "Expected tumoral fraction",
    y = "Predicted tumoral fraction"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  ylim(0,0.1)
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("all_datasets_predicted_vs_expected_lf.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```


```{r}
# Extract numeric positions of factor levels
plot_hf <- subset(bench, bench$expected_tf >= 0.05)  %>%
  mutate(RefType = ifelse(deconv_tool %in% c("MeDeCom","RefFreeCellMix"), "Reference-free", "Reference-based"))
levels_tf <- sort(unique(plot_hf$expected_tf))
p <- ggplot(plot_hf, aes(
  x = as.factor(expected_tf), y = predicted_tf,
  color = RefType, fill = RefType)) +
  geom_boxplot(outliers = F, median.colour = "grey20", color = "grey20") +
  geom_line(data = data.frame(
    x = seq_along(levels_tf),
    y = levels_tf
  ), aes(x = x, y = y), color = "red", linetype = "dashed", inherit.aes = FALSE) +
  facet_wrap(~deconv_tool) +
  labs(
    title = "All datasets (fractions > 0.05)",
    x = "Expected tumoral fraction",
    y = "Predicted tumoral fraction"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  )) +
  scale_fill_manual(values = c(
    "Reference-based" = "#009E73",
    "Reference-free" = "#E69F00"
  ))
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/04_refree_vs_refbased"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("all_datasets_predicted_vs_expected_hf.svg"))
ggsave(output_file, plot = p, width = 8, height = 4)
```
