---
title: '05. Additional plots'
author: "Edoardo Giuili"
date: "2025-09-24"
output: html_document
---

## Introduction

In this tutorial we will analyse the WGBS and RRBS and generate several different visualizations used in the benchmarking to interpret the data.

### Benchmarking metrics

It is important to highlight that to evaluate the performance of the deconvolution tools we considered 3 different key metrics: the root-mean-squared error (RMSE), the area under the curve (AUC-ROC) and the Spearman's rank correlation coefficient (ρ {\displaystyle \rho }). To create an overall bench.resmarking score against which to compare the deconvolution tools, we min-max scaled the metrics and computed the mean of the three metrics to obtain the overall score for each combination. Since RMSE is computed on a proportion of all the samples (specifically, only the samples with tumoral fraction \> 0% --\> 0.87), it will be weighted by this percentage.

### Setting up the environment

First, set up the main directory

```{r setup}
# Don't change this if you git cloned the entire GitHub repository and you are working in the notebooks/ folder
knitr::opts_knit$set(root.dir = '.')
```

Import the necessary libraries for the analysis

```{r message=FALSE, warning=FALSE}
library(ggpubr)
library(dplyr)
library(tidyverse)
library(stringr)
library(pROC)
```

Import the useful functions and the metrics

```{r}
source("../R/compute_auc.R")

# RMSE
rmse <- function(actual, predicted) {
  round(sqrt(mean((actual - predicted)^2)),4)
}
# Normalized RMSE (NRMSE)
nrmse <- function(actual, predicted) {
    round(sqrt(mean((actual - predicted))^2)/mean(actual+10e-8),4)
}
# Spearman's rank correlation coefficient (SCC)
scc <- function(actual, predicted) {
  cor(actual, predicted, method = "spearman", use = "pairwise.complete.obs")
}
```

Now, we will import all the results generated and described above. Note that here we expect that you have git cloned the repository and thus have the expected folder structure where in the folder resources/ there are the results files.

```{r}
df_raw <- read.csv("../resources/final_benchmarking_dataset.csv")
bench.res <- subset(df_raw, df_raw$mixture_type == "in_silico")
```

### Missing values

An important aspect to take into consideration is the number of missing values per tool.

```{r}
missing_percentage <- bench.res %>%
     group_by(deconv_tool) %>%
     summarise(
         total = n(),
         missing = sum(is.na(predicted_tf)),
         percent_missing = 100 * missing / total
     )
missing_percentage
```

Since the presence of missing values is very low (meaning that all the tools almost always produce an output), we will just remove them without penalizing the tools that produced missing values or imputing them.

```{r}
# Remove rows with NA values for further analysis 
bench.res <- bench.res %>%
  filter(!is.na(predicted_tf))
```


## Compute the metrics

For more information about the metrics, please refer to the (first notebook)["01_heatmaps_all_combinationts.Rmd"]. First, compute the RMSE.

```{r}
rmse_minmax <- bench.res %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
  mutate(RMSE = rmse(expected_tf, predicted_tf)) %>%
  ungroup() %>%
  group_by(dmr_tool, seq_depth, tumor_type, expected_tf) %>%
  mutate(
    RMSE_min = min(RMSE),
    RMSE_max = max(RMSE),
    RMSE_minmax = (RMSE - RMSE_min) / (RMSE_max - RMSE_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(RMSE = 1-mean(RMSE_minmax), .groups = "drop")
```

Then we compute the AUC for each tool.

```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(bench.res$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(bench.res$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(bench.res, bench.res$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)

          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_minmax <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(AUC=mean(AUC),.groups = "drop")
```

Finally, we compute the SCC.

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_minmax <- bench.res %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop")
```

Now we can merge the metrics together into a unique dataframe and normalize them. Then, aggregate them into an overall Score.

```{r}
# Merge the dataset
merged_metrics <- rmse_minmax %>%
  left_join(scc_minmax) %>%
  left_join(auc_minmax)

# Compute the number of zeros and non-zero observations
num.zeros <- nrow(bench.res[bench.res$expected_tf==0,])
num.nonzeros <- nrow(bench.res[bench.res$expected_tf!=0,])

# Compute the total number of observation
tot <- num.nonzeros + num.zeros

# Min-max scale the metrics across the tools
norm_merged_metrics <- merged_metrics %>% 
  group_by(dmr_tool, seq_depth, tumor_type) %>%
  mutate(normSCC  = ( SCC - min(SCC)     ) / ( max(SCC)  - min(SCC)  ),
         normAUC  = ( AUC - min(AUC)     ) / ( max(AUC)  - min(AUC)  ),
         normRMSE = ( (RMSE) - min(RMSE) ) / ( max(RMSE) - min(RMSE) ), 
         Score    = normSCC + normAUC + ( num.nonzeros/tot)*normRMSE )

# Normalize the Score
norm_merged_metrics <- norm_merged_metrics %>%
  group_by(dmr_tool, seq_depth, tumor_type) %>%
  mutate(normScore = ( Score - min(Score) ) / ( max(Score) - min(Score) ) )
```


## Plot the results using different visualizations

### Plot general ranking of the tools based on RMSE, SCC, AUC and overall Score

Plot min-max scaled metrics for all the datasets

```{r}
# Create a directory to save the plots (optional)
output_dir <- "../plots/05_additional_plots/rankings_per_dataset/"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Plot the ranking per dataset
for (dataset in c("WGBS", "RRBS", "cfRRBS")) {
      
      # Define the tumor types per dataset
      if (dataset == "WGBS") { 
        tum_types <- c("BRCA","LUAD","LUSC","BLCA") }
  
      else if (dataset == "RRBS") { 
        tum_types <- c("CRC","PC","LC") }
  
      else { tum_types <- c("NBL") }

      # Aggregate across seq depths and tumors and compute the mean Score
      df <- norm_merged_metrics %>% 
        filter(tumor_type %in% tum_types) %>%
        group_by(deconv_tool, dmr_tool, tumor_type) %>%
        summarise(Score = mean(normScore),
                  AUC   = mean(normAUC)  ,
                  RMSE  = mean(normRMSE) ,
                  SCC   = mean(normSCC)  ,
                  .groups = "drop") %>%
        group_by(deconv_tool, dmr_tool) %>%
        summarise(Score = mean(Score),
                  AUC   = mean(AUC)  ,
                  RMSE  = mean(RMSE) ,
                  SCC   = mean(SCC)  ,
                  .groups = "drop")
      
      for (metric in c("Score", "AUC", "SCC", "RMSE")) {
        
        # Rank the tools by mean Score across DMR tools
        mean_score <- df %>%
          group_by(deconv_tool) %>%
          summarise(Mean = mean(.data[[metric]]), .groups = 'drop') %>%
          arrange(Mean)
          
        # Reorder 'deconv_tool' factor levels globally
        df <- df %>%
          mutate(deconv_tool = factor(deconv_tool, levels = mean_score$deconv_tool))
          
        p <- ggplot(df, aes(y = deconv_tool, .data[[metric]], color = dmr_tool, shape = dmr_tool)) +
            geom_point(size = 3, alpha = 0.8) +
            labs(
                title = paste0(metric," in ", dataset),
                x = metric,
                y = ""
            ) +
            theme_classic() +
            theme(
                axis.text.x = element_text(angle = 45, hjust = 1),
                panel.grid.major.y = element_line(color = "grey90"),
                panel.grid.minor.y = element_line(color = "grey90")
            ) +
            scale_color_manual(values=c("limma"="#F8766D","wgbstools"="#00BA38","DMRfinder"="#619CFF"))
        print(p)
        
        # Save the plot as an image
        output_file <- file.path(output_dir, paste0(metric,"_",dataset,".svg"))
        ggsave(output_file, plot = p, width = 8, height = 6)
      }
}
```


### Plot ranking distribution per metric used

Here, we want to visualize how many times a tool is found in a given ranking position across all combinations. We will basically plot the ranking distribution of the tools per metric adopted.

```{r}
# Create a directory to save the plots (optional)
output_dir <- "../plots/05_additional_plots/ranking_distributions"
if (!dir.exists(output_dir)) dir.create(output_dir)

for (dataset in c("WGBS", "RRBS", "cfRRBS")) {
      
  # Define the tumor types per dataset
  if (dataset == "WGBS") { 
    tum_types <- c("BRCA","LUAD","LUSC","BLCA") }
  
  else if (dataset == "RRBS") { 
    tum_types <- c("CRC","PC","LC") }

  else { tum_types <- c("NBL") }

  for (metric in c("normAUC", "normRMSE", "normSCC", "normScore")) {
    
    # Rank the data
    data_ranked <- norm_merged_metrics %>%
      filter(tumor_type %in% tum_types) %>%
      group_by(dmr_tool, tumor_type, seq_depth) %>%
      mutate(Rank = rank(-!!sym(metric), ties.method = "min")) %>%
      ungroup()

    # Compute rank counts
    rank_counts <- data_ranked %>%
      group_by(deconv_tool, Rank) %>%
      summarise(Count = n(), .groups = 'drop')
    
    # Compute Weighted Score: Sum of (Count * Weight) to order the tools
    weighted_scores <- rank_counts %>%
      mutate(Weight = 1 / Rank,
             Weighted = Count * Weight) %>%
      group_by(deconv_tool) %>%
      summarise(Weighted_Score = sum(Weighted),
                .groups = 'drop')
    
    # Order tools based on Weighted Score (higher is better)
    order_tools <- weighted_scores %>%
      arrange(Weighted_Score) %>%
      pull(deconv_tool)
    
    # Set factor levels for plotting
    rank_counts$deconv_tool <- factor(rank_counts$deconv_tool, levels = order_tools)
  
    # Plotting
    p <- ggplot(rank_counts, aes(x = Rank, y = deconv_tool, size = Count)) +
      geom_point(color = "#3670A0", alpha = 0.8) +
      scale_size_continuous(range = c(3,10)) +
      labs(title = paste("Rank distribution based on", metric),
           x = "Rank (lower = better)",
           y = "Deconvolution Tool",
           size = "Count") +
      theme_classic(base_size = 14) +  # Clean white theme
      theme(
        panel.background = element_rect(fill = "white", color = NA),
        plot.background = element_rect(fill = "white", color = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
      ) +
      scale_x_continuous(breaks = seq(1, max(rank_counts$Rank))) +
      ylab("")
    print(p)
    
    # Save the plot as an image
    output_file <- file.path(output_dir, paste0("ranking_distribution_",metric,"_",dataset,".svg"))
    ggsave(output_file, plot = p, width = 8, height = 6)
  }
}
```

### Plot ranking distribution of the LoD

We can now plot the ranking distribution for the LoD as well. For more information on how we computed the LoDs of the tools, you can go over the (second notebook)[02_limit_of_detection.Rmd].

```{r}
# Create a directory to save the plots (optional)
output_dir <- "../plots/05_additional_plots/ranking_distributions"
if (!dir.exists(output_dir)) dir.create(output_dir)

# Import the results generated in 02_ notebook
all.stats <- read.csv("../resources/lod_allstats.csv")

for (dataset in c("WGBS", "RRBS", "cfRRBS")) {
      
  # Define the tumor types per dataset
  if (dataset == "WGBS") { 
    tum_types <- c("BRCA","LUAD","LUSC","BLCA") }
  
  else if (dataset == "RRBS") { 
    tum_types <- c("CRC","PC","LC") }

  else { tum_types <- c("NBL") }
  
  # Compute rank counts
  rank_counts <- all.stats %>%
    filter(tumor_type %in% tum_types) %>%
    group_by(deconv_tool, rank_raw) %>%
    summarise(Count = n(), .groups = 'drop')
  
  # Compute Weighted Score: Sum of (Count * Weight)
  weighted_scores <- rank_counts %>%
    mutate(Weight = 1 / rank_raw,
           Weighted = Count * Weight) %>%
    group_by(deconv_tool) %>%
    summarise(Weighted_Score = sum(Weighted),
              .groups = 'drop')
  
  # Order tools based on Weighted Score (higher is better)
  order_tools <- weighted_scores %>%
    arrange(Weighted_Score) %>%
    pull(deconv_tool)
  
  # Set factor levels for plotting
  rank_counts$deconv_tool <- factor(rank_counts$deconv_tool, levels = order_tools)
  
  # Plotting
  p <- ggplot(rank_counts, aes(x = rank_raw, y = deconv_tool, size = Count)) +
    geom_point(color = "#3670A0", alpha = 0.8) +
    scale_color_gradient() +
    scale_size_continuous(range = c(3,10)) +
    labs(title = paste("Ranking distribution based on LoD"),
         x = "Rank (lower = better)",
         y = "Deconvolution Tool",
         size = "Count") +
    theme_classic(base_size = 14) +  # Clean white theme
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank()
    ) +
    guides(color = "none") +
    scale_x_continuous(breaks = seq(1, max(rank_counts$rank_raw))) +
    ylab("")
  print(p)
  
  # Save the plot as an image
  output_file <- file.path(output_dir, paste0("ranking_distribution_LoD_",dataset,".svg"))
  ggsave(output_file, plot = p, width = 8, height = 6)
}
```


### Ranking dot plots for AUC-ROC score

We can now rank the tools based on the AUC-ROC score. We will use the raw score in this case, since we don not want to include it in combined scores.

```{r}
# Create a directory to save the plots (optional)
output_dir <- "../plots/05_additional_plots/dotplots_rank_AUC"
if (!dir.exists(output_dir)) dir.create(output_dir)

for (dataset in c("WGBS", "RRBS", "cfRRBS")) {
      
  # Define the tumor types per dataset
  if (dataset == "WGBS") { 
    tum_types <- c("BRCA","LUAD","LUSC","BLCA") }
  
  else if (dataset == "RRBS") { 
    tum_types <- c("CRC","PC","LC") }

  else { tum_types <- c("NBL") }
  
  df <- norm_merged_metrics %>% 
          filter(tumor_type %in% tum_types) %>%
          group_by(deconv_tool, seq_depth) %>%
          summarise(AUC = median(AUC))
  
  # Order the seq depths and assign the correct color palette
  if (dataset == "WGBS") {
    df$seq_depth <- factor(df$seq_depth, levels = c("62M", "183M", "310M", "620M"))
    option = "rocket" }
  else { 
    df$seq_depth <- factor(df$seq_depth, levels = c("2M", "5M", "10M", "15M", "20M"))
    option = "viridis"}
  
  # Order the tools by highest AUC
  tool_order <- df %>%
    group_by(deconv_tool) %>%
    summarise(mean_AUC = mean(AUC)) %>%
    arrange(mean_AUC) %>%
    pull(deconv_tool)

  df <- df %>%
    mutate(deconv_tool = factor(deconv_tool, levels = tool_order))

  # Dotplot of the tools AUC score
  p <- ggplot(df, aes(y = deconv_tool, x = AUC, fill = seq_depth)) +
        geom_point(size = 3, alpha = 0.8, color = "black", shape = 21) +
        labs(
            title = paste0(dataset, " dataset"),
            x = "AUC",
            y = ""
        ) +
        theme_classic() +
        theme(
              panel.grid.major.y = element_line(color = "grey90"),
              panel.grid.minor.y = element_line(color = "grey90")) +
        scale_fill_viridis_d(option = option, direction = -1)
  print(p)
  
  # Save the plot as an image
  output_file <- file.path(output_dir, paste0("rank_AUC_",dataset,".svg"))
  ggsave(output_file, plot = p, width = 8, height = 6)
}
```








