---
title: "06. Funkyheatmap and final ranking"
author: "Edoardo Giuili"
date: "2025-09-26"
output: html_document
---

## Introduction

In this tutorial we will plot the funkyheatmap published in the paper, where we summarized the performance of the tools across metrics, DMR tools, sequencing depth, scalability and overall ranking.

### Benchmarking metrics

It is important to highlight that to evaluate the performance of the deconvolution tools we considered 3 different key metrics: the root-mean-squared error (RMSE), the area under the curve (AUC-ROC) and the Spearman's rank correlation coefficient (Žü {\displaystyle \rho }). To create an overall benchmarking score against which to compare the deconvolution tools, we min-max scaled the metrics and computed the mean of the three metrics to obtain the overall score for each combination. Since RMSE is computed on a proportion of all the samples (specifically, only the samples with tumoral fraction \> 0% --\> 0.87), it will be weighted by this percentage.

### Setting up the environment

First, set up the main directory chaning the paths below

```{r setup}
knitr::opts_knit$set(root.dir = '~/OneDrive-UGent/Projects/002_Benchmarking/BenchmarkingAnalysis/')
# Define the directory containing the CSV files
dir_path <- "~/OneDrive-UGent/Projects/002_Benchmarking/BenchmarkingAnalysis/"
```

Import the necessary libraries for the analysis

```{r message=FALSE, warning=FALSE}
library(Metrics)
library(ggpubr)
library(dplyr)
library(tidyr)
library(funkyheatmap)
library(stringr)
library(pROC)
```

Import the useful functions and the metrics

```{r}
source("R/compute_auc.R")

# RMSE
rmse <- function(actual, predicted) {
  round(sqrt(mean((actual - predicted)^2)),4)
}
# Normalized RMSE (NRMSE)
nrmse <- function(actual, predicted) {
    round(sqrt(mean((actual - predicted))^2)/mean(actual+10e-8),4)
}
# Spearman's rank correlation coefficient (SCC)
scc <- function(actual, predicted) {
  cor(actual, predicted, method = "spearman", use = "pairwise.complete.obs")
}
```

Now, we will import all the results generated. Note that here we expect that you have git cloned the repository and thus have the expected folder structure where in the folder resources/ there are the results files.


```{r}
# Read the dataset and remove the na values
df <- read.csv(paste0(dir_path,"resources/final_benchmarking_dataset.csv"))
bench <- subset(df, !(is.na(predicted_tf)))

# Divide the dataframe into each different dataset
in_silico_wgbs <- subset(bench, bench$seq_method == "wgbs")
in_silico_rrbs <- subset(bench, bench$seq_method == "rrbs")
in_silico_cfrrbs <- subset(bench, bench$seq_method == "cfrrbs" & bench$mixture_type == "in_silico")
in_vitro_cfrrbs <- subset(bench, bench$mixture_type == "in_vitro" & !grepl("-subsamp", bench$sample) & bench$deconv_tool!= "Methyl_Resolver" & bench$expected_tf < 0.6)
```


## Generate the metrics per dataset

### WGBS

Compute the AUC values per tumoral fraction, then make the mean to have a mean AUC, then compute the min-max sclaed AUC across tools at each seq_depth, DMR tools and tumor_type.


```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(in_silico_wgbs$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(in_silico_wgbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_wgbs, in_silico_wgbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)

          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_wgbs <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(meanAUC=mean(AUC),.groups = "drop")  %>%
    group_by(dmr_tool, seq_depth, tumor_type) %>%
    mutate(
      AUC_min = min(meanAUC),
      AUC_max = max(meanAUC),
      AUC_minmax = (meanAUC - AUC_min) / (AUC_max - AUC_min)
    ) %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarise(AUC = AUC_minmax, .groups = "drop")
```

Now, we compute the SCC. The SCC is computed for each combination pooling all the observed tumor fractions together to get a value for each combination. The SCC is then min-max scaled.

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_wgbs <- in_silico_wgbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop") %>%
  ungroup() %>%
  group_by(dmr_tool, seq_depth, tumor_type) %>%
  mutate(
    SCC_min = min(SCC),
    SCC_max = max(SCC),
    SCC_minmax = (SCC - SCC_min) / (SCC_max - SCC_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = SCC_minmax, .groups = "drop")
```

Finally compute the RMSE.

```{r}
rmse_wgbs <- in_silico_wgbs %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
  mutate(RMSE = rmse(expected_tf, predicted_tf)) %>%
  ungroup() %>%
  group_by(dmr_tool, seq_depth, tumor_type, expected_tf) %>%
  mutate(
    RMSE_min = min(RMSE),
    RMSE_max = max(RMSE),
    RMSE_minmax = (RMSE - RMSE_min) / (RMSE_max - RMSE_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(RMSE = 1-mean(RMSE_minmax), .groups = "drop")
```

We can now combined the metrics into a unique, overall Score

```{r}
# Compute the number of zeros and non-zero observations
num.zeros <- nrow(in_silico_wgbs[in_silico_wgbs$expected_tf==0,])
num.nonzeros <- nrow(in_silico_wgbs[in_silico_wgbs$expected_tf!=0,])

# Compute the total number of observation
tot <- num.nonzeros + num.zeros

# Compute the Score
score_wgbs <- rmse_wgbs[,c("deconv_tool","dmr_tool","seq_depth","tumor_type")] 
score_wgbs$Score <- ( num.nonzeros/tot)*rmse_wgbs$RMSE + scc_wgbs$SCC + auc_wgbs$AUC
score_wgbs <- score_wgbs %>%
  group_by(dmr_tool,tumor_type,seq_depth) %>%
    mutate(
    Score_min = min(Score),
    Score_max = max(Score),
    Score_minmax = (Score - Score_min) / (Score_max - Score_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, tumor_type,seq_depth) %>%
  summarise(Score = Score_minmax, .groups = "drop")
```


### RRBS

Now we will do the same for the RRBS dataset.

```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(in_silico_rrbs$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(in_silico_rrbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_rrbs, in_silico_rrbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)

          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_rrbs <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(meanAUC=mean(AUC),.groups = "drop")  %>%
    group_by(dmr_tool, seq_depth, tumor_type) %>%
    mutate(
      AUC_min = min(meanAUC),
      AUC_max = max(meanAUC),
      AUC_minmax = (meanAUC - AUC_min) / (AUC_max - AUC_min)
    ) %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarise(AUC = AUC_minmax, .groups = "drop")
```

Now, we compute the SCC. The SCC is computed for each combination pooling all the observed tumor fractions together to get a value for each combination. The SCC is then min-max scaled.

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_rrbs <- in_silico_rrbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop") %>%
  ungroup() %>%
  group_by(dmr_tool, seq_depth, tumor_type) %>%
  mutate(
    SCC_min = min(SCC),
    SCC_max = max(SCC),
    SCC_minmax = (SCC - SCC_min) / (SCC_max - SCC_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = SCC_minmax, .groups = "drop")
```

Finally compute the RMSE.

```{r}
rmse_rrbs <- in_silico_rrbs %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
  mutate(RMSE = rmse(expected_tf, predicted_tf)) %>%
  ungroup() %>%
  group_by(dmr_tool, seq_depth, tumor_type, expected_tf) %>%
  mutate(
    RMSE_min = min(RMSE),
    RMSE_max = max(RMSE),
    RMSE_minmax = (RMSE - RMSE_min) / (RMSE_max - RMSE_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(RMSE = 1-mean(RMSE_minmax), .groups = "drop")
```

We can now combined the metrics into a unique, overall Score

```{r}
# Compute the number of zeros and non-zero observations
num.zeros <- nrow(in_silico_rrbs[in_silico_rrbs$expected_tf==0,])
num.nonzeros <- nrow(in_silico_rrbs[in_silico_rrbs$expected_tf!=0,])

# Compute the total number of observation
tot <- num.nonzeros + num.zeros

# Compute the Score
score_rrbs <- rmse_rrbs[,c("deconv_tool","dmr_tool","seq_depth","tumor_type")] 
score_rrbs$Score <- ( num.nonzeros/tot)*rmse_rrbs$RMSE + scc_rrbs$SCC + auc_rrbs$AUC
score_rrbs <- score_rrbs %>%
  group_by(dmr_tool, tumor_type, seq_depth) %>%
    mutate(
    Score_min = min(Score),
    Score_max = max(Score),
    Score_minmax = (Score - Score_min) / (Score_max - Score_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, tumor_type,seq_depth) %>%
  summarise(Score = Score_minmax, .groups = "drop")
```

### cfRRBS in silico

And for the cfRRBS in silico and in vitro

```{r}
# Create a new dataset dataframe where to store AUC-ROC results
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(in_silico_cfrrbs$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(in_silico_cfrrbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_silico_cfrrbs, in_silico_cfrrbs$tumor_type == type)
  for (seqdepth in unique(filt_tum$seq_depth)) {
    
    # Loop over the DMR tools    
    filt_depth <- subset(filt_tum, filt_tum$seq_depth == seqdepth)
    for (dmrtool in unique(filt_depth$dmr_tool)) {
      
      # Loop over the deconvolution tools
      for (deconv in unique(filt_depth$deconv_tool)) {
  
        # Loop over the tumoral fractions specified above
        for (fraction in unique_fractions) {
          filt_df <- filt_depth %>%
            filter(expected_tf %in% c(0,fraction) & 
                   dmr_tool == dmrtool &
                   deconv_tool == deconv)

          # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
          auc_single_combination <- compute_auc(filt_df, fraction, deconv)
          aucroc_data <- rbind(aucroc_data, auc_single_combination)
        }
      }
    }
  }
}
auc_cfrrbs <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarize(meanAUC=mean(AUC),.groups = "drop")  %>%
    group_by(dmr_tool, seq_depth, tumor_type) %>%
    mutate(
      AUC_min = min(meanAUC),
      AUC_max = max(meanAUC),
      AUC_minmax = (meanAUC - AUC_min) / (AUC_max - AUC_min)
    ) %>%
    group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
    summarise(AUC = AUC_minmax, .groups = "drop")
```

Now, we compute the SCC. The SCC is computed for each combination pooling all the observed tumor fractions together to get a value for each combination. The SCC is then min-max scaled.

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_cfrrbs <- in_silico_cfrrbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop") %>%
  ungroup() %>%
  group_by(dmr_tool, seq_depth, tumor_type) %>%
  mutate(
    SCC_min = min(SCC),
    SCC_max = max(SCC),
    SCC_minmax = (SCC - SCC_min) / (SCC_max - SCC_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = SCC_minmax, .groups = "drop")
```

Finally compute the RMSE.

```{r}
rmse_cfrrbs <- in_silico_cfrrbs %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, expected_tf) %>%
  mutate(RMSE = rmse(expected_tf, predicted_tf)) %>%
  ungroup() %>%
  group_by(dmr_tool, seq_depth, tumor_type, expected_tf) %>%
  mutate(
    RMSE_min = min(RMSE),
    RMSE_max = max(RMSE),
    RMSE_minmax = (RMSE - RMSE_min) / (RMSE_max - RMSE_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(RMSE = 1-mean(RMSE_minmax), .groups = "drop")
```

We can now combined the metrics into a unique, overall Score

```{r}
# Compute the number of zeros and non-zero observations
num.zeros <- nrow(in_silico_wgbs[in_silico_cfrrbs$expected_tf==0,])
num.nonzeros <- nrow(in_silico_wgbs[in_silico_cfrrbs$expected_tf!=0,])

# Compute the total number of observation
tot <- num.nonzeros + num.zeros

# Compute the Score
score_cfrrbs <- rmse_cfrrbs[,c("deconv_tool","dmr_tool","seq_depth","tumor_type")] 
score_cfrrbs$Score <- ( num.nonzeros/tot)*rmse_cfrrbs$RMSE + scc_cfrrbs$SCC + auc_cfrrbs$AUC
score_cfrrbs <- score_cfrrbs %>%
  group_by(dmr_tool,tumor_type,seq_depth) %>%
    mutate(
    Score_min = min(Score),
    Score_max = max(Score),
    Score_minmax = (Score - Score_min) / (Score_max - Score_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, tumor_type,seq_depth) %>%
  summarise(Score = Score_minmax, .groups = "drop")
```

### cfRRBS in vitro

Compute the AUC values per tumoral fraction, then make the make to have a mean AUC, then compute the min-max sclaed AUC across tools at each seq_depth, DMR tools and tumor_type.


```{r message=FALSE, warning=FALSE}
aucroc_data <- data.frame()
# Select the samples with tumoral fractions > 0%
unique_fractions <- unique(in_vitro_cfrrbs$expected_tf)
unique_fractions <- subset(unique_fractions, unique_fractions != 0)

# Loop over the type of tumors
for (type in unique(in_vitro_cfrrbs$tumor_type)) {
  
  # Loop over sequencing depths
  filt_tum <- subset(in_vitro_cfrrbs, in_vitro_cfrrbs$tumor_type == type)
  for (dmrtool in unique(filt_tum$dmr_tool)) {
      
    # Loop over the deconvolution tools
    for (deconv in unique(filt_tum$deconv_tool)) {
  
      # Loop over the tumoral fractions specified above
      for (fraction in unique_fractions) {
        
        filt_df <- filt_tum %>%
          filter(expected_tf %in% c(0,fraction) & 
                 dmr_tool == dmrtool &
                 deconv_tool == deconv)

        # Compute the AUC-ROC score per combination and concatanate it to the final dataframe
        auc_single_combination <- compute_auc(filt_df, fraction, deconv)
        aucroc_data <- rbind(aucroc_data, auc_single_combination)
      }
    }
  }
}
auc_cfrrbs_invitro <- aucroc_data %>%
    group_by(deconv_tool, dmr_tool, tumor_type, expected_tf) %>%
    summarize(AUC=mean(auc),.groups = "drop") %>%
    group_by(deconv_tool, dmr_tool, tumor_type) %>%
    summarize(meanAUC=mean(AUC),.groups = "drop")  %>%
    group_by(dmr_tool, tumor_type) %>%
    mutate(
      AUC_min = min(meanAUC),
      AUC_max = max(meanAUC),
      AUC_minmax = (meanAUC - AUC_min) / (AUC_max - AUC_min)
    ) %>%
    group_by(deconv_tool, dmr_tool, tumor_type) %>%
    summarise(AUC = AUC_minmax, .groups = "drop")

```


```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_cfrrbs_invitro <- in_vitro_cfrrbs %>%
  group_by(deconv_tool, dmr_tool, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop") %>%
  ungroup() %>%
  group_by(dmr_tool, tumor_type) %>%
  mutate(
    SCC_min = min(SCC),
    SCC_max = max(SCC),
    SCC_minmax = (SCC - SCC_min) / (SCC_max - SCC_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, tumor_type) %>%
  summarise(SCC = mean(SCC_minmax), .groups = "drop")
```


```{r}
rmse_cfrrbs_invitro <- in_vitro_cfrrbs %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, dmr_tool, tumor_type, expected_tf) %>%
  mutate(RMSE = rmse(expected_tf, predicted_tf)) %>%
  ungroup() %>%
  group_by(dmr_tool, tumor_type, expected_tf) %>%
  mutate(
    RMSE_min = min(RMSE),
    RMSE_max = max(RMSE),
    RMSE_minmax = (RMSE - RMSE_min) / (RMSE_max - RMSE_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, tumor_type) %>%
  summarise(RMSE = 1-mean(RMSE_minmax), .groups = "drop")
```


```{r}
# Compute the number of zeros and non-zero observations
num.zeros <- nrow(in_vitro_cfrrbs[in_vitro_cfrrbs$expected_tf==0,])
num.nonzeros <- nrow(in_vitro_cfrrbs[in_vitro_cfrrbs$expected_tf!=0,])

# Compute the total number of observation
tot <- num.nonzeros + num.zeros

# Compute the Score
score_cfrrbs_invitro <- rmse_cfrrbs_invitro[,c("deconv_tool","dmr_tool","tumor_type")] 
score_cfrrbs_invitro$Score <- ( num.nonzeros/tot)*rmse_cfrrbs_invitro$RMSE + scc_cfrrbs_invitro$SCC + auc_cfrrbs_invitro$AUC
score_cfrrbs_invitro <- score_cfrrbs_invitro %>%
  group_by(dmr_tool,tumor_type) %>%
    mutate(
    Score_min = min(Score),
    Score_max = max(Score),
    Score_minmax = (Score - Score_min) / (Score_max - Score_min)
  ) %>%
  group_by(deconv_tool, dmr_tool, tumor_type) %>%
  summarise(Score = Score_minmax, .groups = "drop")
```


## Aggregate the metrics

We will now aggregate the metrics of the datasets according to different variables. We will aggregate per metric to have an overall per metric performance. We will also aggregate per DMR tool, per sequencing depth and per dataset. We will give a different weigth to each dataset to balance the difference in terms of sequencing technology. Since there is only 1 WGBS dataset and 3 RRBS-like datasets, we will give the weight of 1 to the WGBS, 0.5 to the RRBS and 0.25 to both the in-silico and in-vitro cfRRBS dataset.

### Aggregate per metric

Here we will calculate the mean RMSE per deconvolution tool per dataset and then aggregate the mean RMSE using the weighted mean.

```{r}
# Compute the mean RMSE for WGBS
aggr_rmse_wgbs <- rmse_wgbs %>%
  group_by(deconv_tool) %>%
  summarise(RMSE = mean(RMSE))
aggr_rmse_wgbs$weight <- 1

# For RRBS
aggr_rmse_rrbs <- rmse_rrbs %>%
  group_by(deconv_tool) %>%
  summarise(RMSE = mean(RMSE))
aggr_rmse_rrbs$weight <- 0.5

# For in-silico cfRRBS
aggr_rmse_cfrrbs <- rmse_cfrrbs %>%
  group_by(deconv_tool) %>%
  summarise(RMSE = mean(RMSE))
aggr_rmse_cfrrbs$weight <- 0.25

# For in-vitro cfRRBS
aggr_rmse_cfrrbs_invitro <- rmse_cfrrbs_invitro %>%
  group_by(deconv_tool) %>%
  summarise(RMSE = mean(RMSE))
aggr_rmse_cfrrbs_invitro$weight <- 0.25

# Concatanate the dataframes into 1 
aggr_rmse <- bind_rows(
  aggr_rmse_wgbs,
  aggr_rmse_rrbs,
  aggr_rmse_cfrrbs,
  aggr_rmse_cfrrbs_invitro
)

# Compute the weighted mean
funkyheat_permetric_rmse <- aggr_rmse %>%
  group_by(deconv_tool) %>%
  summarise(RMSE = weighted.mean(RMSE, w = weight)) %>%
  arrange(desc(RMSE))

```


Do the same for the SCC

```{r}
# Compute the mean SCC for WGBS
aggr_scc_wgbs <- scc_wgbs %>%
  group_by(deconv_tool) %>%
  summarise(SCC = mean(SCC))
aggr_scc_wgbs$weight <- 1

# RRBS
aggr_scc_rrbs <- scc_rrbs %>%
  group_by(deconv_tool) %>%
  summarise(SCC = mean(SCC))
aggr_scc_rrbs$weight <- 0.5

# in-silico cfRRBS
aggr_scc_cfrrbs <- scc_cfrrbs %>%
  group_by(deconv_tool) %>%
  summarise(SCC = mean(SCC))
aggr_scc_cfrrbs$weight <- 0.25

# in-vitro cfRRBS
aggr_scc_cfrrbs_invitro <- scc_cfrrbs_invitro %>%
  group_by(deconv_tool) %>%
  summarise(SCC = mean(SCC))
aggr_scc_cfrrbs_invitro$weight <- 0.25

# Concatanate
aggr_scc <- bind_rows(
  aggr_scc_wgbs,
  aggr_scc_rrbs,
  aggr_scc_cfrrbs,
  aggr_scc_cfrrbs_invitro
)

# Weighted mean
funkyheat_permetric_scc <- aggr_scc %>%
  group_by(deconv_tool) %>%
  summarise(SCC = weighted.mean(SCC, w = weight)) %>%
  arrange(desc(SCC))
```


And the same for the AUC-ROC score

```{r}
# Compute the mean AUC-ROC score for WGBS
aggr_auc_wgbs <- auc_wgbs %>%
  group_by(deconv_tool) %>%
  summarise(AUC = mean(AUC))
aggr_auc_wgbs$weight <- 1

# RRBS
aggr_auc_rrbs <- auc_rrbs %>%
  group_by(deconv_tool) %>%
  summarise(AUC = mean(AUC))
aggr_auc_rrbs$weight <- 0.5

# in-silico cfRRBS
aggr_auc_cfrrbs <- auc_cfrrbs %>%
  group_by(deconv_tool) %>%
  summarise(AUC = mean(AUC))
aggr_auc_cfrrbs$weight <- 0.25

# in-vitro cfRRBS
aggr_auc_cfrrbs_invitro <- auc_cfrrbs_invitro %>%
  group_by(deconv_tool) %>%
  summarise(AUC = mean(AUC))
aggr_auc_cfrrbs_invitro$weight <- 0.25

# Concatanate
aggr_auc <- bind_rows(
  aggr_auc_wgbs,
  aggr_auc_rrbs,
  aggr_auc_cfrrbs,
  aggr_auc_cfrrbs_invitro
)

# Weighted mean
funkyheat_permetric_auc <- aggr_auc %>%
  group_by(deconv_tool) %>%
  summarise(AUC = weighted.mean(AUC, w = weight)) %>%
  arrange(desc(AUC))
```


### Aggregate per DMRtool

We will now do the same as above, but aggregating per DMR tool. In this case, we will use the overall Score as a metric of performance for the aggregation per DMR tool.

```{r}
# Compute the mean Score for WGBS
aggr_score_wgbs <- score_wgbs %>%
  group_by(deconv_tool,dmr_tool) %>%
  summarise(Score = mean(Score)) %>%
  pivot_wider(
    names_from = dmr_tool,
    values_from = Score
  )
aggr_score_wgbs$weight <- 1

# RRBS
aggr_score_rrbs <- score_rrbs %>%
  group_by(deconv_tool,dmr_tool) %>%
  summarise(Score = mean(Score)) %>%
  pivot_wider(
    names_from = dmr_tool,
    values_from = Score
  )
aggr_score_rrbs$weight <- 0.5

# in-silico cfRRBS
aggr_score_cfrrbs <- score_cfrrbs %>%
  group_by(deconv_tool,dmr_tool) %>%
  summarise(Score = mean(Score)) %>%
  pivot_wider(
    names_from = dmr_tool,
    values_from = Score
  )
aggr_score_cfrrbs$weight <- 0.25

# in-vitro cfRRBS
aggr_score_cfrrbs_invitro <- score_cfrrbs_invitro %>%
  group_by(deconv_tool,dmr_tool) %>%
  summarise(Score = mean(Score)) %>%
  pivot_wider(
    names_from = dmr_tool,
    values_from = Score
  )
aggr_score_cfrrbs_invitro$weight <- 0.25

# Concatanate
aggr_score <- bind_rows(
  aggr_score_wgbs,
  aggr_score_rrbs,
  aggr_score_cfrrbs,
  aggr_score_cfrrbs_invitro
)

# Weighted mean
funkyheat_perdmrtool_score <- aggr_score %>%
  group_by(deconv_tool) %>%
  summarise(DMRfinder = weighted.mean(DMRfinder, w = weight),
            limma = weighted.mean(limma, w = weight),
            wgbstools = weighted.mean(wgbstools, w = weight)) %>%
  arrange(desc(wgbstools))
```


### Aggregate per dataset

We can now aggregate per dataset, again using the overall Score.

```{r}
# Compute the mean Score for WGBS
aggr_score_wgbs <- score_wgbs %>%
  group_by(deconv_tool) %>%
  summarise(WGBS = mean(Score))

# RRBS
aggr_score_rrbs <- score_rrbs %>%
  group_by(deconv_tool) %>%
  summarise(RRBS = mean(Score))

# cfRRBS in silico
aggr_score_cfrrbs <- score_cfrrbs %>%
  group_by(deconv_tool) %>%
  summarise(cfRRBS_insilico = mean(Score))

# cfRRBS in vitro
aggr_score_cfrrbs_invitro <- score_cfrrbs_invitro %>%
  group_by(deconv_tool) %>%
  summarise(cfRRBS_invitro = mean(Score))

# Concatanate
funkyheat_perdataset_score <- aggr_score_wgbs %>%
  inner_join(aggr_score_rrbs, by = "deconv_tool") %>%
  inner_join(aggr_score_cfrrbs, by = "deconv_tool") %>%
  inner_join(aggr_score_cfrrbs_invitro, by = "deconv_tool")

```


### Aggregate per dataset per seq_depth

Now we will aggregate per sequencing depth. Since WGBS and RRBS have different sequencing depths, we will aggregate within each dataset

```{r}
# Compute the mean Score for WGBS seq depth
aggr_score_perseqdepth_wgbs <- score_wgbs %>%
  group_by(deconv_tool, seq_depth) %>%
  summarise(WGBS = mean(Score)) %>%
  pivot_wider(
    names_from = seq_depth,
    values_from = WGBS,
    names_glue = "WGBS_{seq_depth}"
  )

# RRBS
aggr_score_perseqdepth_rrbs <- score_rrbs %>%
  group_by(deconv_tool, seq_depth) %>%
  summarise(RRBS = mean(Score)) %>%
  pivot_wider(
    names_from = seq_depth,
    values_from = RRBS,
    names_glue = "RRBS_{seq_depth}"
  )

# cfRRBS in silico
aggr_score_perseqdepth_cfrrbs <- score_cfrrbs %>%
  group_by(deconv_tool, seq_depth) %>%
  summarise(cfRRBS_insilico = mean(Score)) %>%
  pivot_wider(
    names_from = seq_depth,
    values_from = cfRRBS_insilico,
    names_glue = "cfRRBS_{seq_depth}"
  )

# cfRRBS in vitro
funkyheat_perdataset_perseqdepth_score <- aggr_score_perseqdepth_wgbs %>%
  inner_join(aggr_score_perseqdepth_rrbs, by = "deconv_tool") %>%
  inner_join(aggr_score_perseqdepth_cfrrbs, by = "deconv_tool")
```

### Scalability

An important aspect in the evaluation of deconvolution tools is the scalability computed as the mean of the running time score and memory usage score. Since the scalability difference between the deconvolution tools is not that strong, we will give a low weigth (0.1) when we will aggregate it with the other scores to get the final overall Score.

```{r}
# Read the dataset and subset the useful columns
scalability.df <- read.csv(paste0(dir_path, "resources/scalability.clean.csv"), row.names = 1)
scalability.df <- scalability.df[,c("name","dmr","samples","realtime_sec","peak_rss_mb")]
names(scalability.df)[1] <- "deconv_tool"

# Change the deconvolution tools names with the appropriate ones
scalability.df <- scalability.df %>%
  mutate(deconv_tool = recode(deconv_tool,
                       "CELFIE" = "CelFiE",
                       "EPIDISH_RPC" = "EpiDISH_RPC",
                       "EPIDISH_eq" = "EpiDISH_CP_eq",
                       "EPIDISH_ineq" = "EpiDISH_CP_ineq",
                       "EPISCORE" = "EpiSCORE",
                       "METDECODE" = "MetDecode",
                       "METH_ATLAS" = "meth_atlas",
                       "PRMETH" = "PRMeth"))

# Compute the mean running time and memory usage across the different combinations
average_runtime <- scalability.df %>%
  group_by(deconv_tool) %>%
  summarise(Runtime = round(mean(realtime_sec),2))  %>%
  arrange(Runtime)
average_memory <- scalability.df %>%
  group_by(deconv_tool) %>%
  summarise(Memory = round(mean(peak_rss_mb),0))  %>%
  arrange(Memory)

# Min-max scale both running time and memory
scalability_memory <- data.frame(
  deconv_tool = average_memory$deconv_tool,
  Memory = 1 - ( average_memory$Memory - min(average_memory$Memory) ) / ( max(average_memory$Memory) - min(average_memory$Memory) ) )
scalability_time <- data.frame(
  deconv_tool = average_runtime$deconv_tool,
  Time = 1 - ( average_runtime$Runtime - min(average_runtime$Runtime) ) / ( max(average_runtime$Runtime) - min(average_runtime$Runtime) ) )

# Concatanate the datasets and make a mean score
aggr_scalab_score = merge(scalability_memory, scalability_time, by = "deconv_tool")
aggr_scalab_score$Score = (aggr_scalab_score$Memory + aggr_scalab_score$Time) / 2
aggr_scalab_score <- aggr_scalab_score[, c("deconv_tool","Score")]
```

## Compute overall score

We can finally compute the overall Score across the different datasets, including the scalability Score.

```{r}
# WGBS
overall_wgbs_score <- score_wgbs %>%
  group_by(deconv_tool) %>%
  summarise(Score = mean(Score))

# RRBS
overall_rrbs_score <- score_rrbs %>%
  group_by(deconv_tool) %>%
  summarise(Score = mean(Score))

# cfRRBS in silico
overall_cfrrbs_score <- score_cfrrbs %>%
  group_by(deconv_tool) %>%
  summarise(Score = mean(Score))

# cfRRBS in vitro
overall_cfrrbs_invitro_score <- score_cfrrbs_invitro %>%
  group_by(deconv_tool) %>%
  summarise(Score = mean(Score))

# Assign scores
overall_wgbs_score$weight <- 1
overall_rrbs_score$weight <- 0.5
overall_cfrrbs_score$weight <- 0.25
overall_cfrrbs_invitro_score$weight <- 0.25
aggr_scalab_score$weight <- 0.1
```

Combine the scores using a weighted mean

```{r}
# Combine all score dataframes into one
all_scores <- bind_rows(
  overall_wgbs_score,
  overall_rrbs_score,
  overall_cfrrbs_score,
  overall_cfrrbs_invitro_score,
  aggr_scalab_score
)

# Compute weighted mean per deconv_tool
overall_score <- all_scores %>%
  group_by(deconv_tool) %>%
  summarise(
    Overall_Score = sum(Score * weight) / sum(weight),
    .groups = "drop"
  ) %>%
  arrange(desc(Overall_Score)) %>%
  mutate(Rank = rev(rank(Overall_Score)))
```




## Plot funkyheatmap

Now we are ready to plot the funkyheatmap. However, we have to prepare the dataset in such a format which is accepted by the library. We will first concatanate all the datasets previously generated.

```{r}
# Concatanate the datasets
data_funkyheat <- inner_join(funkyheat_permetric_auc, funkyheat_permetric_rmse, by= "deconv_tool" ) %>%
  inner_join(funkyheat_permetric_scc, by = "deconv_tool") %>%
  inner_join(funkyheat_perdmrtool_score, by = "deconv_tool") %>%
  inner_join(funkyheat_perdataset_score, by = "deconv_tool") %>%
  inner_join(funkyheat_perdataset_perseqdepth_score, by = "deconv_tool") %>%
  inner_join(average_memory, by = "deconv_tool") %>%
  inner_join(average_runtime, by = "deconv_tool") %>%
  inner_join(overall_score, by = "deconv_tool")

# Add the python/R logos depending on the tool
data_funkyheat$Availability <- c("python-logo.png","R_logo.png","R_logo.png",
                                 "python-logo.png","python-logo.png",
                                 "python-logo.png","R_logo.png","R_logo.png",
                                 "R_logo.png","R_logo.png")
data_funkyheat$Availability <- paste0(dir_path,"resources/", data_funkyheat$Availability)

# Add the core statistical method depending on the tool and order the dataset based on the overall score
data_funkyheat$Statistical_Method <- c("EM","CP/QP","CP/QP","NNLS+\nAdam","NNLS","NNLS","NMF","RPC","RPC","v-SVR")
data_funkyheat <- data_funkyheat[order(data_funkyheat$Overall_Score, decreasing = T),]
```

Prepare the column info necessary for the funkyheatmap. Here we define the palette to use for each group.

```{r}
palettes <- tribble(
  ~palette, ~colours,
  # blues palette
  "overall", grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(9, "Greys")[-1]))(101),
  "benchmark", grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(9, "Blues") %>% c("#011636")))(101),
  "scaling", grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(9, "Reds")[-8:-9]))(101),
  "stability", grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(9, "YlOrBr")[-7:-9]))(101),
  "qc", grDevices::colorRampPalette(rev(RColorBrewer::brewer.pal(9, "Greens")[-1] %>% c("#00250f")))(101),
  "white6black4", c(rep("white", 3), rep("black", 7)),
  # "column_annotation", c(overall = "#555555", benchmark = "#4292c6", scaling = "#f6483a", stability = "#fe9929", qc = "#41ab5d")
)
```

And here we define the groups to plot, that have to match the final dataframe (data_funkyheatmap) column names

```{r}
column_info <- tribble(
  ~id,                  ~group,               ~name,                  ~geom,        ~palette,    ~options,
  "deconv_tool",        "Method",             "",                     "text",       "overall",   lst(hjust = 0, width = 6),
  "Statistical_Method", "Method",             "Statistical method",   "text",       "overall",   lst(width = 2),
  "Availability",       "Method",             "Availability",         "image",      "overall",   lst(width = 1.5),
  "RMSE",               "Summary",            "1-RMSE",               "bar",        "benchmark",        lst(),
  "SCC",                "Summary",            "SCC",                  "bar",        "benchmark",        lst(),
  "AUC",                "Summary",            "AUC",                  "bar",        "benchmark",        lst(),
  "DMRfinder",          "Per DMRtool",        "DMRfinder",            "funkyrect",  "benchmark", lst(),
  "limma",              "Per DMRtool",        "limma",                "funkyrect",  "benchmark", lst(),
  "wgbstools",          "Per DMRtool",        "wgbstools",            "funkyrect",  "benchmark", lst(),
  "WGBS",               "Per dataset",        "WGBS",                 "funkyrect",  "benchmark", lst(),
  "RRBS",               "Per dataset",        "RRBS",                 "funkyrect",  "benchmark", lst(),
  "cfRRBS_insilico",    "Per dataset",        "cfRRBS (in-silico)",   "funkyrect",  "benchmark", lst(),
  "cfRRBS_invitro",     "Per dataset",        "cfRRBS (in-vitro)",    "funkyrect",  "benchmark", lst(),
  "WGBS_62M",           "Per SeqDepth (WGBS)","62M",                  "funkyrect",  "benchmark", lst(),
  "WGBS_183M",          "Per SeqDepth (WGBS)","183M",                 "funkyrect",  "benchmark", lst(),
  "WGBS_310M",          "Per SeqDepth (WGBS)","320M",                 "funkyrect",  "benchmark", lst(),
  "WGBS_620M",          "Per SeqDepth (WGBS)","620M",                 "funkyrect",  "benchmark", lst(),
  "RRBS_2M",            "Per SeqDepth (RRBS)","2M",                   "funkyrect",  "benchmark", lst(),
  "RRBS_5M",            "Per SeqDepth (RRBS)","5M",                   "funkyrect",  "benchmark", lst(),
  "RRBS_10M",           "Per SeqDepth (RRBS)","10M",                  "funkyrect",  "benchmark", lst(),
  "RRBS_15M",           "Per SeqDepth (RRBS)","15M",                  "funkyrect",  "benchmark", lst(),
  "RRBS_20M",           "Per SeqDepth (RRBS)","20M",                  "funkyrect",  "benchmark", lst(),
  "cfRRBS_2M",          "Per SeqDepth (cfRRBS)","2M",                   "funkyrect",  "benchmark", lst(),
  "cfRRBS_5M",          "Per SeqDepth (cfRRBS)","5M",                 "funkyrect",  "benchmark", lst(),
  "cfRRBS_10M",         "Per SeqDepth (cfRRBS)","10M",                "funkyrect",  "benchmark", lst(),
  "cfRRBS_15M",         "Per SeqDepth (cfRRBS)","15M",                "funkyrect",  "benchmark", lst(),
  "cfRRBS_20M",         "Per SeqDepth (cfRRBS)","20M",                "funkyrect",  "benchmark", lst(),
  "Memory",             "Scalability",        "Memory (MB)",          "rect",       "scaling",   lst(width = 1.7),
  "Memory",             "Scalability",        "Memory (MB)",          "text",       "black",     lst(overlay = TRUE),
  "Runtime",            "Scalability",        "Runtime (sec)",        "rect",       "scaling",   lst(width = 1.7),
  "Runtime",            "Scalability",        "Runtime (sec)",        "text",       "black",     lst(overlay = TRUE),
  "Overall_Score",      "Final Ranking",      "Overall score",        "bar",        "stability", lst(width = 2.4),
  "Rank",               "Final Ranking",      "Ranking",              "text",       "stability", lst(width = 1.8),


)
```

Define the subgroups/subsections to plot in the funkyheatmap

```{r}
column_groups <- tibble(
  Experiment = c("Method", "General performance", "General performance", "General performance","Performance per sequencing depth", "Performance per sequencing depth", "Performance per sequencing depth", "Scalability","Final Ranking"),
  Category = c("", "Per metric", "Per DMRtool", "Per dataset", "WGBS" , "RRBS", "cfRRBS", "",""),
  group = c("Method", "Summary", "Per DMRtool", "Per dataset", "Per SeqDepth (WGBS)", "Per SeqDepth (RRBS)", "Per SeqDepth (cfRRBS)", "Scalability","Final Ranking"),
  palette = c("overall","benchmark", "benchmark", "benchmark", "benchmark", "benchmark", "benchmark", "scaling","stability")
)
```


And finally plot it

```{r}
g <- funky_heatmap(data_funkyheat, 
                   column_info = column_info,
                   column_groups = column_groups,
                   palettes = palettes
                   ) 
g
```

And save it in the correct location in svg format for a better visualization.

```{r}
output_dir <- paste0(dir_path,"plots/06_funkyheatmap")
if (!dir.exists(output_dir)) dir.create(output_dir)

output_file <- file.path(output_dir, paste0("funkyheatmap.svg"))
ggsave(output_file, g, width = g$width, height = g$height)
```




