---
title: "Tools scalability"
author: "Edoardo Giuili"
date: "2025-09-26"
output: html_document
---

## Introduction

In this tutorial we will go over the scalability results in terms of running time and memory usage. To compute the scalability of the tools we used the following strategy. We selected 5, 25, 50 and 100 in silico mixtures from the NBL cfRRBS dataset and we run each tool using sequentially 25, 50, 100, 250 and 500 DMRs. Second, we extracted the memory peak and the real time computed automatically by the Nextflow pipeline for each individual deconvolution tool. We run this analysis allocating to each deconvolution tool 8 CPUs and 16GB of memory, the minimum required resources to run the pipeline.

### Setting up the environment

First, set up the main directory

```{r setup}
# Don't change this if you git cloned the entire GitHub repository and you are working in the notebooks/ folder
knitr::opts_knit$set(root.dir = './')
```

Import the necessary libraries for the analysis

```{r message=FALSE, warning=FALSE}
library(ggpubr)
library(ggplot2)
library(dplyr)
library(patchwork)
library(stringr)
```

Import the results

```{r}
res.scal <- read.csv("../resources/scalability.clean.csv")
```


## Plot the results

### Boxplots of memory and realtime

We can plot the running time and memory distribution among the different combinations using boxplots. We first have to convert the memory and running time into the correct format.

```{r}
# Helper function to convert memory to MB
convert_mem <- function(x) {
  x <- str_trim(x)
  val <- as.numeric(str_extract(x, "[0-9.]+"))
  unit <- str_extract(x, "MB|GB")
  ifelse(unit == "GB", val * 1024, val)
}

convert_time <- function(x) {
  x <- tolower(trimws(x))
  total <- numeric(length(x))

  for (i in seq_along(x)) {
    str <- x[i]
    sec <- as.numeric(stringr::str_extract(str, "\\d+\\.?\\d*(?=s)"))
    min <- as.numeric(stringr::str_extract(str, "\\d+\\.?\\d*(?=m(?!s))")) # avoid matching ms
    ms  <- as.numeric(stringr::str_extract(str, "\\d+\\.?\\d*(?=ms)"))

    sec[is.na(sec)] <- 0
    min[is.na(min)] <- 0
    ms[is.na(ms)]   <- 0

    total[i] <- min * 60 + sec + ms / 1000
  }

  return(total)
}
```


```{r}
# Apply conversions to memory
data_clean_memory <- res.scal %>%
  mutate(
    realtime_sec = convert_mem(peak_rss)
  )
# Calculate median duration per method and sort
order_memory <- data_clean_memory %>%
  group_by(name) %>%
  summarise(med = mean(realtime_sec, na.rm = TRUE)) %>%
  arrange(med) %>%
  pull(name)
# Set factor levels for duration plot
data_clean_memory$name <- factor(data_clean_memory$name, levels = order_memory)

# Repeat for the time
data_clean_time <- res.scal %>%
  mutate(
    realtime_sec = convert_time(realtime)
  )
order_time <- data_clean_time %>%
  group_by(name) %>%
  summarise(med = median(realtime_sec, na.rm = TRUE)) %>%
  arrange(med) %>%
  pull(name)
data_clean_time$name <- factor(data_clean_time$name, levels = order_time)
```

And now plot the results combined

```{r}
# Plot: Boxplot of duration
p1 <- ggplot(data_clean_time, aes(x = name, y = realtime_sec)) +
  geom_boxplot() +
  labs(title = "", x = "", y = "Real time (seconds)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(breaks = seq(0,110,by=20))


# Plot: Boxplot of peak RSS
p2 <- ggplot(data_clean_memory, aes(x = name, y = peak_rss_mb)) +
  geom_boxplot() +
  labs(title = "", x = "", y = "Memory usage (MB)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Show plots
combined.plot <- p1 + p2
combined.plot
```

```{r}
output_dir <- "../plots/07_scalability"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("memory_time_boxplots_8CPU_16GB.svg"))
ggsave(output_file, combined.plot, width = 8, height = 5)
```


### Compare the memory at different number of DMRs and samples

We can also check how the memory changes increasing the number of DMRs and samples.

```{r}
tool_colors <- c(
  "CELFIE"             ="#A6CEE3", 
  "CIBERSORT"          ="#1F78B4",
  "EPIDISH_eq"      ="#B2DF8A", 
  "EPIDISH_ineq"    ="#33A02C",
  "EPIDISH_RPC"        ="#FB9A99", 
  "EPISCORE"           ="#E31A1C",
  "METDECODE"          ="#FDBF6F", 
  "METH_ATLAS"         ="#FF7F00",
  "PRMETH"             ="#CAB2D6", 
  "UXM"                ="#6A3D9A"
)
```


```{r}
# Summarize by mean or median
summary_df_dmr <- data_clean_memory %>%
  group_by(name, dmr) %>%
  summarise(mean_rss = mean(peak_rss_mb), .groups = "drop")
summary_df_samples <- data_clean_memory %>%
  group_by(name, samples) %>%
  summarise(mean_rss = mean(peak_rss_mb), .groups = "drop")

p1 <- ggplot(summary_df_dmr, aes(x = dmr, y = mean_rss, color = name)) +
  geom_line() +
  geom_point() +
  labs(title = "",
       x = "Number of DMRs", y = "Memory usage (MB)") +
  theme_classic() +
  theme(legend.position = "none") +
  scale_color_manual(values = tool_colors) +
  scale_x_continuous(breaks = c(25, 100, 250, 500)) +
  ylim(0,800)

p2 <- ggplot(summary_df_samples, aes(x = samples, y = mean_rss, color = name)) +
  geom_line() +
  geom_point() +
  labs(title = "",
       x = "Number of samples", y = "") +
  theme_classic() +
  scale_color_manual(values = tool_colors) +
  scale_x_continuous(breaks = c(5, 20, 50, 100)) +
  ylim(0,800)

# Show plots
combined.plot <- p1 + p2
combined.plot
```


```{r}
output_dir <- "../plots/07_scalability"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("memory_samples_dmrs_8CPU_16GB.svg"))
ggsave(output_file, combined.plot, width = 8, height = 4)
```


### Compare the real time at different number of DMRs and samples

And we can visualize the same for the running time

```{r}
# Summarize by mean or median
summary_df_dmr <- data_clean_time %>%
  group_by(name, dmr) %>%
  summarise(mean_time = mean(realtime_sec), .groups = "drop")
summary_df_samples <- data_clean_time %>%
  group_by(name, samples) %>%
  summarise(mean_time = mean(realtime_sec), .groups = "drop")

p1 <- ggplot(summary_df_dmr, aes(x = dmr, y = mean_time, color = name)) +
  geom_line() +
  geom_point() +
  labs(title = "",
       x = "Number of DMRs", y = "Real time (seconds)") +
  theme_classic() +
  theme(legend.position = "none") +
  scale_color_manual(values = tool_colors) +
  scale_x_continuous(breaks = c(25, 100, 250, 500)) +
  scale_y_continuous(breaks = seq(0,70,by=10)) + 
  ylim(0,30)

p2 <- ggplot(summary_df_samples, aes(x = samples, y = mean_time, color = name)) +
  geom_line() +
  geom_point() +
  labs(title = "",
       x = "Number of samples", y = "") +
  theme_classic() +
  scale_x_continuous(breaks = c(5, 20, 50, 100)) + 
  scale_color_manual(values = tool_colors) +
  scale_y_continuous(breaks = seq(0,70,by=10)) +
  ylim(0,30)

# Show plots
combined.plot <- p1 + p2
combined.plot
```



```{r}
output_dir <- "../plots/07_scalability"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("time_samples_dmrs_8CPU_16GB.svg"))
ggsave(output_file, combined.plot, width = 8, height = 4)
```








