---
title: "08. Preciseness of tumor predictions (RMSE and SCC)"
author: "Edoardo Giuili"
date: "2025-10-02"
output: html_document
---

## Introduction

In this tutorial we want to show how to analyse the preciseness of the tools in predicting tumoral fractions, evaluated in this case using only RMSE and SCC metrics.

### Setting up the environment

First, set up the main directory

```{r setup}
# Don't change this if you git cloned the entire GitHub repository and you are working in the notebooks/ folder
knitr::opts_knit$set(root.dir = './')
```

Import the necessary libraries for the analysis

```{r message=FALSE, warning=FALSE}
library(ggpubr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(tidytext)
```

Now, we will import all the results generated on the in silico mixtures. Note that here we expect that you have git cloned the repository and thus have the expected folder structure where in the folder resources/ there are the results files. We will filter out the expected fractions equal to 0% and divide the expected fractions into low (<1%), medium (1%<=x<=5%) and high (>=10%). 

```{r}
df_raw <- read.csv("../resources/final_benchmarking_dataset.csv")
bench.res <- subset(df_raw, df_raw$mixture_type == "in_silico") %>%
  filter(!is.na(predicted_tf) & expected_tf != 0) %>%
  mutate(fraction_class = ifelse( expected_tf <= 0.025, "low", ifelse( expected_tf >0.025 & expected_tf <= 0.1, "medium", "high") ) )
```

Import the useful functions and the metrics and specify the colors

```{r}
# RMSE
rmse <- function(actual, predicted) {
  round(sqrt(mean((actual - predicted)^2)),4)
}
# Spearman's rank correlation coefficient (SCC)
scc <- function(actual, predicted) {
  cor(actual, predicted, method = "spearman", use = "pairwise.complete.obs")
}
tool_colors <- c(
  "CelFiE"             ="#A6CEE3", 
  "CIBERSORT"          ="#1F78B4",
  "Houseman_eq"        ="#B2DF8A", 
  "Houseman_ineq"      ="#33A02C",
  "EpiDISH"            ="#FB9A99", 
  "EpiSCORE"           ="#E31A1C",
  "MetDecode"          ="#FDBF6F", 
  "meth_atlas"         ="#FF7F00",
  "PRMeth"             ="#CAB2D6", 
  "UXM"                ="#6A3D9A"
)
```


## Analyse WGBS dataset

We will start analysing the WGBS in silico mixtures, to then proceed to the RRBS datasets. First, we will subset the results to only keep the WGBS samples,

```{r}
bench_wgbs <- subset(bench.res,bench.res$seq_method=="wgbs")
```

We can now compute the RMSE and SCC. Differently from other notebooks, here we will use the raw RMSE and SCC without min-max scaling it across deconvolution tools. This is necessary since we want to check, within each deconvolution tool, whether the performance changes varying the sequencing depth and expected fractions.

```{r}
rmse_raw_wgbs <- bench_wgbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, fraction_class) %>%
  summarize("1-RMSE" = 1-rmse(expected_tf, predicted_tf))
```

Finally, we compute the SCC. The SCC is computed for each combination pooling all the observed tumor fractions together to get a value for each combination. 

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_raw_wgbs <- bench_wgbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop")
```


We can now combine the RMSE and SCC metrics into a unique dataframe

```{r}
df.wgbs.metrics <- scc_raw_wgbs %>%
  left_join(rmse_raw_wgbs)
```


### Plot the data of WGBS-TT

#### Boxplots of RMSE

```{r}
# Compute the SCC
df.scc <- scc_raw_wgbs %>% group_by(deconv_tool) %>% summarise(SCC = mean(SCC))

# Set fraction_class as an ordered factor
rmse_raw_wgbs$fraction_class <- factor(
  rmse_raw_wgbs$fraction_class,
  levels = c("low", "medium", "high")
)

# Build fin.scc dataframe
fin.scc <- data.frame(
  deconv_tool   = rep(df.scc$deconv_tool, 3),
  SCC           = rep(df.scc$SCC, 3),
  fraction_class = rep(c("low", "medium", "high"), each = length(df.scc$deconv_tool))
)

# Ensure consistent factor levels
fin.scc$fraction_class <- factor(fin.scc$fraction_class, levels = c("low", "medium", "high"))

# Order tools based on SCC in the "low" fraction_class
tool_order <- fin.scc %>%
  filter(fraction_class == "low") %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)

# Apply consistent ordering to factors
rmse_raw_wgbs$deconv_tool <- factor(rmse_raw_wgbs$deconv_tool, levels = tool_order)
fin.scc$deconv_tool       <- factor(fin.scc$deconv_tool, levels = tool_order)
```


```{r}
# Plot
p1 <- ggplot(data = rmse_raw_wgbs, aes(
    x = reorder_within(deconv_tool,-`1-RMSE`, fraction_class, median), 
    y = `1-RMSE`, fill = deconv_tool)
  ) +
  geom_jitter(size=1, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  scale_color_manual(name = "", values = c("SCC" = "red")) +
  facet_wrap(~ fraction_class, axes = "all", scales = "free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(title = "WGBS-TT dataset",
       x = "", y = "1 - RMSE",
       color = "SCC") +
  scale_x_reordered() +
  scale_fill_manual(values = tool_colors) +
  scale_color_manual(values = tool_colors)
p1
```


```{r}
# Create a directory to save the plots (optional)
output_dir <- "../plots/08_preciseness"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Save the plot as an image
output_file <- file.path(output_dir, paste0("preciseness_WGBS_dataset_boxplots.svg"))
ggsave(output_file, plot = p1, width = 8, height = 4)
```

#### Boxplots of SCC

```{r}
# Compute the SCC
median_scc <- scc_raw_wgbs %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)

scc_raw_wgbs$deconv_tool <- factor(scc_raw_wgbs$deconv_tool, levels = median_scc)
```


```{r}
# Plot
p1 <- ggplot(data = scc_raw_wgbs, aes(
    x = deconv_tool, y = SCC, fill = deconv_tool)
  ) +
  geom_jitter(size=1.5, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  labs(title = "WGBS-TT dataset",
       x = "", y = "SCC") +
  scale_fill_manual(values = tool_colors) +
  scale_color_manual(values = tool_colors)
p1
```

```{r}
# Create a directory to save the plots (optional)
output_dir <- "../plots/08_preciseness"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Save the plot as an image
output_file <- file.path(output_dir, paste0("preciseness_SCC_WGBS_dataset_boxplots.svg"))
ggsave(output_file, plot = p1, width = 8, height = 4)
```

## Analyse RRBS-TT dataset

And now we analyze the RRBS in silico mixtures 

```{r}
bench_rrbs <- subset(bench.res,bench.res$seq_method=="rrbs")
```

We can now compute the RMSE and SCC. Differently from other notebooks, here we will use the raw RMSE and SCC without min-max scaling it across deconvolution tools. This is necessary since we want to check, within each deconvolution tool, whether the performance changes varying the sequencing depth and expected fractions.

```{r}
rmse_raw_rrbs <- bench_rrbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, fraction_class) %>%
  summarize("1-RMSE" = 1-rmse(expected_tf, predicted_tf))
```

Finally, we compute the SCC. The SCC is computed for each combination pooling all the observed tumor fractions together to get a value for each combination. 

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_raw_rrbs <- bench_rrbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop")
```


We can now combine the RMSE and SCC metrics into a unique dataframe

```{r}
df.rrbs.metrics <- scc_raw_rrbs %>%
  left_join(rmse_raw_rrbs)
```


### Plot the data of RRBS

#### Boxplots of RMSE

```{r}
# Compute the SCC
df.scc <- scc_raw_rrbs %>% group_by(deconv_tool) %>% summarise(SCC = mean(SCC))

# Set fraction_class as an ordered factor
rmse_raw_rrbs$fraction_class <- factor(
  rmse_raw_rrbs$fraction_class,
  levels = c("low", "medium", "high")
)

# Build fin.scc dataframe
fin.scc <- data.frame(
  deconv_tool   = rep(df.scc$deconv_tool, 3),
  SCC           = rep(df.scc$SCC, 3),
  fraction_class = rep(c("low", "medium", "high"), each = length(df.scc$deconv_tool))
)

# Ensure consistent factor levels
fin.scc$fraction_class <- factor(fin.scc$fraction_class, levels = c("low", "medium", "high"))

# Order tools based on SCC in the "low" fraction_class
tool_order <- fin.scc %>%
  filter(fraction_class == "low") %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)

# Apply consistent ordering to factors
rmse_raw_rrbs$deconv_tool <- factor(rmse_raw_rrbs$deconv_tool, levels = tool_order)
fin.scc$deconv_tool       <- factor(fin.scc$deconv_tool, levels = tool_order)
```


```{r}
# Plot
p1 <- ggplot(data = rmse_raw_rrbs, aes(
    x = reorder_within(deconv_tool,-`1-RMSE`, fraction_class, median), 
    y = `1-RMSE`, fill = deconv_tool)
  ) +
  geom_jitter(size=1, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  facet_wrap(~ fraction_class, axes = "all", scales = "free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(title = "RRBS-TT",
       x = "", y = "1 - RMSE",
       color = "SCC") +
  scale_x_reordered() +
  scale_fill_manual(values = tool_colors) +
  scale_color_manual(values = tool_colors) +
  scale_color_manual(values = tool_colors)
p1
```


```{r}
# Create a directory to save the plots (optional)
output_dir <- "../plots/08_preciseness"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Save the plot as an image
output_file <- file.path(output_dir, paste0("preciseness_RRBS_dataset_boxplots.svg"))
ggsave(output_file, plot = p1, width = 8, height = 4)
```

#### Boxplots of SCC

```{r}
# Compute the SCC
median_scc <- scc_raw_rrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)

scc_raw_rrbs$deconv_tool <- factor(scc_raw_rrbs$deconv_tool, levels = median_scc)
```


```{r}
# Plot
p1 <- ggplot(data = scc_raw_rrbs, aes(
    x = deconv_tool, y = SCC, fill = deconv_tool)
  ) +
  geom_jitter(size=1.5, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  labs(title = "RRBS-TT",
       x = "", y = "SCC") +
  scale_fill_manual(values = tool_colors) +
  scale_color_manual(values = tool_colors)
p1
```

```{r}
# Create a directory to save the plots (optional)
output_dir <- "../plots/08_preciseness"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Save the plot as an image
output_file <- file.path(output_dir, paste0("preciseness_SCC_RRBS_dataset_boxplots.svg"))
ggsave(output_file, plot = p1, width = 8, height = 4)
```


## Analyse RRBS-CL dataset

And now we analyze the cfRRBS in silico mixtures 

```{r}
bench_cfrrbs <- subset(bench.res,bench.res$seq_method=="cfrrbs")
```

We can now compute the RMSE and SCC. Differently from other notebooks, here we will use the raw RMSE and SCC without min-max scaling it across deconvolution tools. This is necessary since we want to check, within each deconvolution tool, whether the performance changes varying the sequencing depth and expected fractions.

```{r}
rmse_raw_cfrrbs <- bench_cfrrbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, fraction_class) %>%
  summarize("1-RMSE" = 1-rmse(expected_tf, predicted_tf))
```

Finally, we compute the SCC. The SCC is computed for each combination pooling all the observed tumor fractions together to get a value for each combination. 

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_raw_cfrrbs <- bench_cfrrbs %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop")
```


We can now combine the RMSE and SCC metrics into a unique dataframe

```{r}
df.cfrrbs.metrics <- scc_raw_cfrrbs %>%
  left_join(rmse_raw_cfrrbs)
```


### Plot the data of RRBS

#### Boxplots of RMSE

```{r}
# Compute the SCC
df.scc <- scc_raw_cfrrbs %>% group_by(deconv_tool) %>% summarise(SCC = mean(SCC))

# Set fraction_class as an ordered factor
rmse_raw_cfrrbs$fraction_class <- factor(
  rmse_raw_cfrrbs$fraction_class,
  levels = c("low", "medium", "high")
)

# Build fin.scc dataframe
fin.scc <- data.frame(
  deconv_tool   = rep(df.scc$deconv_tool, 3),
  SCC           = rep(df.scc$SCC, 3),
  fraction_class = rep(c("low", "medium", "high"), each = length(df.scc$deconv_tool))
)

# Ensure consistent factor levels
fin.scc$fraction_class <- factor(fin.scc$fraction_class, levels = c("low", "medium", "high"))

# Order tools based on SCC in the "low" fraction_class
tool_order <- fin.scc %>%
  filter(fraction_class == "low") %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)

# Apply consistent ordering to factors
rmse_raw_cfrrbs$deconv_tool <- factor(rmse_raw_cfrrbs$deconv_tool, levels = tool_order)
fin.scc$deconv_tool       <- factor(fin.scc$deconv_tool, levels = tool_order)
```


```{r}
# Plot
p1 <- ggplot(data = rmse_raw_cfrrbs, aes(
    x = reorder_within(deconv_tool,-`1-RMSE`, fraction_class, median), 
    y = `1-RMSE`, fill = deconv_tool)
  ) +
  geom_jitter(size=1, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  facet_wrap(~fraction_class, scales = "free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(title = "RRBS-CL dataset",
       x = "", y = "1 - RMSE",
       color = "SCC") +
  scale_x_reordered() +
  scale_fill_manual(values = tool_colors) +
  scale_color_manual(values = tool_colors)
p1
```


```{r}
# Create a directory to save the plots (optional)
output_dir <- "../plots/08_preciseness"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Save the plot as an image
output_file <- file.path(output_dir, paste0("preciseness_cfRRBS_dataset_boxplots.svg"))
ggsave(output_file, plot = p1, width = 8, height = 4)
```


#### Boxplots of SCC

```{r}
# Compute the SCC
median_scc <- scc_raw_cfrrbs %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)

scc_raw_cfrrbs$deconv_tool <- factor(scc_raw_cfrrbs$deconv_tool, levels = median_scc)
```


```{r}
# Plot
p1 <- ggplot(data = scc_raw_cfrrbs, aes(
    x = deconv_tool, y = SCC, fill = deconv_tool)
  ) +
  geom_jitter(size=1.5, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  labs(title = "RRBS-CL dataset",
       x = "", y = "SCC") +
  scale_fill_manual(values = tool_colors) +
  scale_color_manual(values = tool_colors)
p1
```

```{r}
# Create a directory to save the plots (optional)
output_dir <- "../plots/08_preciseness"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Save the plot as an image
output_file <- file.path(output_dir, paste0("preciseness_SCC_cfRRBS_dataset_boxplots.svg"))
ggsave(output_file, plot = p1, width = 8, height = 4)
```

## Analyse all datasets combined

We finally analyse all the datasets combined. We can now compute the RMSE and SCC. Differently from other notebooks, here we will use the raw RMSE and SCC without min-max scaling it across deconvolution tools. This is necessary since we want to check, within each deconvolution tool, whether the performance changes varying the sequencing depth and expected fractions.

```{r}
rmse_raw_aggr <- bench.res %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type, fraction_class) %>%
  summarize("1-RMSE" = 1-rmse(expected_tf, predicted_tf))
```

Finally, we compute the SCC. The SCC is computed for each combination pooling all the observed tumor fractions together to get a value for each combination. 

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_raw_aggr <- bench.res %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop")
```


We can now combine the RMSE and SCC metrics into a unique dataframe

```{r}
df.aggr.metrics <- scc_raw_aggr %>%
  left_join(rmse_raw_aggr)
```


### Plot all datasets aggregated

#### Boxplots of RMSE

```{r}
# Set fraction_class as an ordered factor
rmse_raw_aggr$fraction_class <- factor(
  rmse_raw_aggr$fraction_class,
  levels = c("low", "medium", "high")
)
```


```{r}
# Plot
p1 <- ggplot(data = rmse_raw_aggr, aes(
    x = reorder_within(deconv_tool,-`1-RMSE`, fraction_class, median), 
    y = `1-RMSE`, fill = deconv_tool)
  ) +
  geom_jitter(size=1, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  facet_wrap(~ fraction_class, axes = "all", scales = "free_x") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(title = "All datasets",
       x = "", y = "1 - RMSE") +
  scale_x_reordered() +
  scale_fill_manual(values = tool_colors) +
  scale_color_manual(values = tool_colors) +
  ylim(0.5,1)
p1
```


```{r}
# Create a directory to save the plots (optional)
output_dir <- "../plots/08_preciseness"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Save the plot as an image
output_file <- file.path(output_dir, paste0("preciseness_all_dataset_boxplots.svg"))
ggsave(output_file, plot = p1, width = 8, height = 4)
```

#### Boxplots of SCC

```{r}
# Compute the SCC
median_scc <- scc_raw_aggr %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)

scc_raw_aggr$deconv_tool <- factor(scc_raw_aggr$deconv_tool, levels = median_scc)
```


```{r}
# Plot
p1 <- ggplot(data = scc_raw_aggr, aes(
    x = deconv_tool, y = SCC, fill = deconv_tool)
  ) +
  geom_jitter(size=1.5, alpha=0.8, width = 0.3, aes(color = deconv_tool)) +
  geom_boxplot(outliers = FALSE, alpha = 0.8) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(
      face = "bold",
      background = element_rect(fill = "grey90", color = NA, inherit.blank = FALSE))
  ) +
  labs(title = "All datasets",
       x = "", y = "SCC") +
  scale_fill_manual(values = tool_colors) +
  scale_color_manual(values = tool_colors)
p1
```

```{r}
# Create a directory to save the plots (optional)
output_dir <- "../plots/08_preciseness"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Save the plot as an image
output_file <- file.path(output_dir, paste0("preciseness_SCC_all_dataset_boxplots.svg"))
ggsave(output_file, plot = p1, width = 4, height = 4)
```

