---
title: "09. Analysis of reference free tools when the sample size changes"
author: "Edoardo Giuili"
date: "2026-04-08"
output: html_document
---

## Introduction

In this tutorial, we demonstrate how to compare the performance of reference-free methods as the sample size—i.e., the number of bulk samples to be deconvolved—changes. Since reference-free methods deconvolve all bulk samples jointly, the number of samples may influence the performance of these tools as it increases or decreases.

### Setting up the environment

First, set up the main directory

```{r setup}
# Don't change this if you git cloned the entire GitHub repository and you are working in the notebooks/ folder
knitr::opts_knit$set(root.dir = './')
```

Import the necessary libraries for the analysis

```{r message=FALSE, warning=FALSE}
library(ggpubr)
library(ggplot2)
library(dplyr)
library(pROC)
library(tidytext)
library(tidyr)
```

Import the useful functions and the metrics

```{r}
source("../R/compute_auc.R")
source("../R/compute_lod.R")
# RMSE
rmse <- function(actual, predicted) {
  round(sqrt(mean((actual - predicted)^2)),4)
}
# Spearman's rank correlation coefficient (SCC)
scc <- function(actual, predicted) {
  cor(actual, predicted, method = "spearman", use = "pairwise.complete.obs")
}
```

Read the results generated by both reference-free and reference-based tools and concatenate them in a unique dataframe.

```{r}
bench_refree <- read.csv("../resources/refree_different_sample_sizes.csv")
bench_refree <- bench_refree %>% 
  filter(!is.na(bench_refree$predicted_tf)) %>%
  mutate(fraction_class = ifelse( expected_tf <= 0.025, "low", ifelse( expected_tf >0.025 & expected_tf <= 0.1, "medium", "high")))
```


## Performance calculation for RRBS-NBL dataset

To check the performance of ref-free tools at lower number of samples, we run
deconvolution usinf MeDeCom and RefFreeCellMix on 100, 50, 25 and 5 samples from
the RRBS-CL dat, and in particular from the NBL tumor type.


```{r}
tool_colors <- c(
  "MeDeCom"        = "#66C2A5",  # teal
  "RefFreeCellMix" = "#666666"   # dark grey-blue
)
```

### RMSE and SCC

We now compute the SCC and the RMSE

```{r}
scc_cfrrbs <- bench_refree %>%
  group_by(deconv_tool, sample_size) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop")
```


Compute the RMSE.

```{r}
rmse_cfrrbs <- bench_refree %>%
  filter(expected_tf != 0) %>%
  group_by(deconv_tool, sample_size) %>%
  summarize(RMSE = 1-rmse(expected_tf, predicted_tf))
```

Aggregate the results

```{r}
aggr.res <- merge(scc_cfrrbs, rmse_cfrrbs, by = c("deconv_tool", "sample_size"))
aggr.res
```


```{r}
# Ensure sample_size is ordered
df_long <- aggr.res %>%
  mutate(sample_size = factor(sample_size, levels = sort(unique(sample_size)))) %>%
  pivot_longer(cols = c(SCC, RMSE), names_to = "metric", values_to = "value")
df_long$value <- round(df_long$value, 4)

p_combined <- ggplot(df_long, aes(x = sample_size, y = value,
                                 color = deconv_tool, group = deconv_tool)) +
  geom_line() +
  geom_point(size = 3) +
  facet_wrap(~metric, scales = "free_y",
             labeller = labeller(metric = c(SCC = "SCC",
                                            RMSE = "1-RMSE"))) +
  labs(title = "",
       x = "Sample Size",
       y = "Value",
       color = "Deconvolution Tool") +
  theme_classic() +
  scale_color_manual(values = tool_colors)

print(p_combined)
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/09_refree_different_sample_size"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

output_file <- file.path(output_dir, paste0("rmse_scc_different_sample_size.pdf"))
ggsave(output_file, plot = p_combined, width = 8, height = 4)
```



















