---
title: "10. Analysis of reference-based tools performance in deconvolving in-silico mixtures using individual tumor samples"
output: html_document
date: "2025-09-23"
author: "Edoardo Giuili"
---

## Introduction

In this notebook we will analyse reference-based deconvolution tools performance
in deconvolving in-silico mixtures generated mixing healthy cfDNA and individual
tumor samples (unpooled samples), instead of pooling the samples from the same 
tumor type into a unique BAM file (pooled samples). This analysis wants to check 
whether the pooling of tumor samples into a unique one had a strong influence on
the ranking of the tools due to a potential reduction of the intra-patient 
heterogeneity due to the pooling. For this, we generated 10 replicates for 5 
tumor fractions per tumor sample at a sequencing depth of 183M reads. Given that 
we have 8 samples in total, this sums up to 300 in-silico mixtures.

### Setting up the environment

First, set up the main directory

```{r setup}
# Don't change this if you git cloned the entire GitHub repository and you are working in the notebooks/ folder
knitr::opts_knit$set(root.dir = './')
```

Import the necessary libraries for the analysis

```{r message=FALSE, warning=FALSE}
library(ggpubr)
library(dplyr)
library(tidyverse)
library(ComplexHeatmap)
library(stringr)
library(pROC)
library(car)
library(tidytext)
library(patchwork)
```

Import the useful functions and the metrics

```{r}
source("../R/compute_auc.R")

# RMSE
rmse <- function(actual, predicted) {
  round(sqrt(mean((actual - predicted)^2)),4)
}
# Spearman's rank correlation coefficient (SCC)
scc <- function(actual, predicted) {
  cor(actual, predicted, method = "spearman", use = "pairwise.complete.obs")
}
tool_colors <- c(
  "CelFiE"             ="#A6CEE3", 
  "CIBERSORT"          ="#1F78B4",
  "Houseman_eq"        ="#B2DF8A", 
  "Houseman_ineq"      ="#33A02C",
  "EpiDISH"            ="#FB9A99", 
  "EpiSCORE"           ="#E31A1C",
  "MetDecode"          ="#FDBF6F", 
  "meth_atlas"         ="#FF7F00",
  "PRMeth"             ="#CAB2D6", 
  "UXM"                ="#6A3D9A"
)
```

Now, we will import all the results generated and described above. Note that 
here we expect that you have git cloned the repository and thus have the expected 
folder structure where in the folder resources/ there are the results files.
From the old results, generated on the in-silico mixtures from pooling tumor 
samples together, we will only select the predictions made on the 183M sequencing
depth combination and on the tumor fractions specified below, which are the 5 
tumor fractions generated using the tumor samples non-pooled.

```{r}
df_raw <- read.csv("../resources/final_benchmarking_dataset.csv")
df_unmerged <- read.csv("../resources/refbased_results_unmerged_insilico_mixtures.csv")
bench.res <- subset(df_raw, 
                      df_raw$mixture_type == "in_silico" & 
                      df_raw$seq_method == "wgbs" & 
                      df_raw$seq_depth == "183M" &
                      df_raw$expected_tf %in% c("0","0.001", "0.007", "0.01", "0.05", "0.5"))
bench.res.zeros <- subset(bench.res, bench.res$expected_tf=="0")
bench.res.zeros$batch <- "unpooled_samples"
bench.res$batch <- "pooled_samples"
df_unmerged$batch <- "unpooled_samples"
fin.res <- rbind(bench.res, df_unmerged, bench.res.zeros)
```

### Missing values

An important aspect to take into consideration is the number of missing values per tool.

```{r}
missing_percentage <- fin.res %>%
     group_by(deconv_tool) %>%
     summarise(
         total = n(),
         missing = sum(is.na(predicted_tf)),
         percent_missing = 100 * missing / total
     )
missing_percentage
```

Since the presence of missing values is very low, we will just remove them 
without penalizing the tools that produced missing values or imputing them.

```{r}
# Remove rows with NA values for further analysis 
fin.res <- fin.res %>%
  filter(!is.na(predicted_tf))
```

## Plot the results

We can now compare the results obtained on the in-silico mixtures generated from
the pooled samples vs the results obtained using the unpooled samples.

### Rank the tools by median SCC

```{r}
# Compute the metrics: modify the RMSE= 1) nrmse 2) rmse 3) rmse of logit(values)
scc_raw_wgbs <- fin.res %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop")
```

#### BLCA

Let's now check the results on the BLCA samples, and then proceed with the 3
remaining tumor types.

```{r}
scc_raw_wgbs_blca1 <- subset(scc_raw_wgbs, scc_raw_wgbs$tumor_type =="BLCA_1")

# Compute the SCC
median_scc <- scc_raw_wgbs_blca1 %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)
scc_raw_wgbs_blca1$deconv_tool <- factor(scc_raw_wgbs_blca1$deconv_tool, levels = median_scc)

p1 <- ggplot(scc_raw_wgbs_blca1, aes(x = deconv_tool, y = SCC, fill = deconv_tool)) +
  geom_boxplot() +
  labs(x = "", y = "SCC") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.2, vjust = 0.2)) +
  scale_fill_manual(values = tool_colors) + 
  ggtitle("BLCA_1 (unpooled)")
```


```{r}
scc_raw_wgbs_blca4 <- subset(scc_raw_wgbs, scc_raw_wgbs$tumor_type =="BLCA_4")

# Compute the SCC
median_scc <- scc_raw_wgbs_blca4 %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)
scc_raw_wgbs_blca4$deconv_tool <- factor(scc_raw_wgbs_blca4$deconv_tool, levels = median_scc)

p2 <- ggplot(scc_raw_wgbs_blca4, aes(x = deconv_tool, y = SCC, fill = deconv_tool)) +
  geom_boxplot() +
  labs(x = "", y = "SCC") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.2, vjust = 0.2)) +
  scale_fill_manual(values = tool_colors) + 
  ggtitle("BLCA_4 (unpooled)")
```


```{r}
scc_raw_wgbs_blca <- subset(scc_raw_wgbs, scc_raw_wgbs$tumor_type =="BLCA")

# Compute the SCC
median_scc <- scc_raw_wgbs_blca %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)
scc_raw_wgbs_blca$deconv_tool <- factor(scc_raw_wgbs_blca$deconv_tool, levels = median_scc)

p3 <- ggplot(scc_raw_wgbs_blca, aes(x = deconv_tool, y = SCC, fill = deconv_tool)) +
  geom_boxplot() +
  labs(x = "", y = "SCC") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.2, vjust = 0.2)) +
  scale_fill_manual(values = tool_colors) + 
  ggtitle("BLCA (pooled)")
```

```{r}
# Combine
p <- (p1 / p2 / p3) +
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "right"))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/10_performance_refbased_unmerged"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
file_output <- paste0(output_dir,"/ranking_scc_blca.pdf")
ggsave(file_output, p, width = 12, height = 10)
```


#### BRCA


```{r}
scc_raw_wgbs_brca3 <- subset(scc_raw_wgbs, scc_raw_wgbs$tumor_type =="BRCA_3")

# Compute the SCC
median_scc <- scc_raw_wgbs_brca3 %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)
scc_raw_wgbs_brca3$deconv_tool <- factor(scc_raw_wgbs_brca3$deconv_tool, levels = median_scc)

p1 <- ggplot(scc_raw_wgbs_brca3, aes(x = deconv_tool, y = SCC, fill = deconv_tool)) +
  geom_boxplot() +
  labs(x = "", y = "SCC") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.2, vjust = 0.2)) +
  scale_fill_manual(values = tool_colors) + 
  ggtitle("BRCA_3 (unpooled)")
```


```{r}
scc_raw_wgbs_brca5 <- subset(scc_raw_wgbs, scc_raw_wgbs$tumor_type =="BRCA_5")

# Compute the SCC
median_scc <- scc_raw_wgbs_brca5 %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)
scc_raw_wgbs_brca5$deconv_tool <- factor(scc_raw_wgbs_brca5$deconv_tool, levels = median_scc)

p2 <- ggplot(scc_raw_wgbs_brca5, aes(x = deconv_tool, y = SCC, fill = deconv_tool)) +
  geom_boxplot() +
  labs(x = "", y = "SCC") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.2, vjust = 0.2)) +
  scale_fill_manual(values = tool_colors) + 
  ggtitle("BRCA_5 (unpooled)")
```


```{r}
scc_raw_wgbs_brca <- subset(scc_raw_wgbs, scc_raw_wgbs$tumor_type =="BRCA")

# Compute the SCC
median_scc <- scc_raw_wgbs_brca %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)
scc_raw_wgbs_brca$deconv_tool <- factor(scc_raw_wgbs_brca$deconv_tool, levels = median_scc)

p3 <- ggplot(scc_raw_wgbs_brca, aes(x = deconv_tool, y = SCC, fill = deconv_tool)) +
  geom_boxplot() +
  labs(x = "", y = "SCC") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.2, vjust = 0.2)) +
  scale_fill_manual(values = tool_colors) + 
  ggtitle("BRCA (pooled)")
```

```{r}
# Combine
p <- (p1 / p2 / p3) +
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "right"))
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/10_performance_refbased_unmerged"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
file_output <- paste0(output_dir,"/ranking_scc_brca.pdf")
ggsave(file_output, p, width = 12, height = 10)
```


#### LUSC


```{r}
scc_raw_wgbs_lusc4 <- subset(scc_raw_wgbs, scc_raw_wgbs$tumor_type =="LUSC_4")

# Compute the SCC
median_scc <- scc_raw_wgbs_lusc4 %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)
scc_raw_wgbs_lusc4$deconv_tool <- factor(scc_raw_wgbs_lusc4$deconv_tool, levels = median_scc)

p1 <- ggplot(scc_raw_wgbs_lusc4, aes(x = deconv_tool, y = SCC, fill = deconv_tool)) +
  geom_boxplot() +
  labs(x = "", y = "SCC") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.2, vjust = 0.2)) +
  scale_fill_manual(values = tool_colors) + 
  ggtitle("LUSC_4 (unpooled)")
```


```{r}
scc_raw_wgbs_lusc5 <- subset(scc_raw_wgbs, scc_raw_wgbs$tumor_type =="LUSC_5")

# Compute the SCC
median_scc <- scc_raw_wgbs_lusc5 %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)
scc_raw_wgbs_lusc5$deconv_tool <- factor(scc_raw_wgbs_blca4$deconv_tool, levels = median_scc)

p2 <- ggplot(scc_raw_wgbs_lusc5, aes(x = deconv_tool, y = SCC, fill = deconv_tool)) +
  geom_boxplot() +
  labs(x = "", y = "SCC") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.2, vjust = 0.2)) +
  scale_fill_manual(values = tool_colors) + 
  ggtitle("LUSC_5 (unpooled)")
```


```{r}
scc_raw_wgbs_lusc <- subset(scc_raw_wgbs, scc_raw_wgbs$tumor_type =="LUSC")

# Compute the SCC
median_scc <- scc_raw_wgbs_lusc %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)
scc_raw_wgbs_lusc$deconv_tool <- factor(scc_raw_wgbs_lusc$deconv_tool, levels = median_scc)

p3 <- ggplot(scc_raw_wgbs_lusc, aes(x = deconv_tool, y = SCC, fill = deconv_tool)) +
  geom_boxplot() +
  labs(x = "", y = "SCC") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.2, vjust = 0.2)) +
  scale_fill_manual(values = tool_colors) + 
  ggtitle("LUSC (pooled)")
```


```{r}
# Combine
p <- (p1 / p2 / p3) +
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "right"))
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/10_performance_refbased_unmerged"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
file_output <- paste0(output_dir,"/ranking_scc_lusc.pdf")
ggsave(file_output, p, width = 12, height = 10)
```


#### LUAD


```{r}
scc_raw_wgbs_luad1 <- subset(scc_raw_wgbs, scc_raw_wgbs$tumor_type =="LUAD_1")

# Compute the SCC
median_scc <- scc_raw_wgbs_luad1 %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)
scc_raw_wgbs_luad1$deconv_tool <- factor(scc_raw_wgbs_luad1$deconv_tool, levels = median_scc)

p1 <- ggplot(scc_raw_wgbs_luad1, aes(x = deconv_tool, y = SCC, fill = deconv_tool)) +
  geom_boxplot() +
  labs(x = "", y = "SCC") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.2, vjust = 0.2)) +
  scale_fill_manual(values = tool_colors) + 
  ggtitle("LUAD_1 (unpooled)")
```


```{r}
scc_raw_wgbs_luad5 <- subset(scc_raw_wgbs, scc_raw_wgbs$tumor_type =="LUAD_5")

# Compute the SCC
median_scc <- scc_raw_wgbs_luad5 %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)
scc_raw_wgbs_luad5$deconv_tool <- factor(scc_raw_wgbs_luad5$deconv_tool, levels = median_scc)

p2 <- ggplot(scc_raw_wgbs_luad5, aes(x = deconv_tool, y = SCC, fill = deconv_tool)) +
  geom_boxplot() +
  labs(x = "", y = "SCC") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.2, vjust = 0.2)) +
  scale_fill_manual(values = tool_colors) + 
  ggtitle("LUAD_5 (unpooled)")
```


```{r}
scc_raw_wgbs_luad <- subset(scc_raw_wgbs, scc_raw_wgbs$tumor_type =="LUAD")

# Compute the SCC
median_scc <- scc_raw_wgbs_luad %>% 
  group_by(deconv_tool) %>% 
  summarise(SCC = median(SCC)) %>%
  arrange(desc(SCC)) %>%
  pull(deconv_tool)
scc_raw_wgbs_luad$deconv_tool <- factor(scc_raw_wgbs_luad$deconv_tool, levels = median_scc)

p3 <- ggplot(scc_raw_wgbs_luad, aes(x = deconv_tool, y = SCC, fill = deconv_tool)) +
  geom_boxplot() +
  labs(x = "", y = "SCC") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.2, vjust = 0.2)) +
  scale_fill_manual(values = tool_colors) + 
  ggtitle("LUAD (pooled)")
```


```{r}
# Combine
p <- (p1 / p2 / p3) +
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "right"))
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/10_performance_refbased_unmerged"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
file_output <- paste0(output_dir,"/ranking_scc_luad.pdf")
ggsave(file_output, p, width = 12, height = 10)
```


### Check ranks correlation for pooled and unpooled samples (SCC)

We can now check the correlation between the ranks of the tools when using the
in silico mixtures generated with the pooled and unpooled samples. The ranks
are obtained according to decreasing Spearman's rank correlation coefficient (SCC).

#### BLCA

```{r}
make_rank_plot <- function(data, tumor_label, pooled_rank) {

  df_rank <- data %>%
    filter(tumor_type == tumor_label) %>%
    group_by(deconv_tool) %>%
    summarise(score = median(SCC, na.rm = TRUE), .groups = "drop") %>%
    mutate(rank_target = rank(-score)) %>%
    inner_join(pooled_rank, by = "deconv_tool")

  cor_val <- cor(df_rank$rank_blca,
                 df_rank$rank_target,
                 method = "spearman",
                 use = "complete.obs")

  p <- ggplot(df_rank, aes(rank_blca, rank_target)) +
    geom_point(size = 3) +
    geom_text(aes(label = deconv_tool), vjust = -0.5, size = 6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    theme_bw(base_size = 25) +
    scale_x_continuous(breaks = 1:10, limits = c(1, 10)) +
    scale_y_continuous(breaks = 1:10, limits = c(1, 10)) +
    labs(
      x = "Rank in BLCA (pooled)",
      y = paste0("Rank in ", tumor_label)
    ) +
    annotate(
      "text",
      x = 2, y = 10,
      label = paste0("Spearman coeff = ", round(cor_val, 3)),
      hjust = 0, size = 7
    )

  list(plot = p, cor = cor_val)
}
```


```{r}
blca_pooled_rank <- scc_raw_wgbs %>%
  filter(tumor_type == "BLCA") %>%
  group_by(deconv_tool) %>%
  summarise(score = median(SCC, na.rm = TRUE), .groups = "drop") %>%
  mutate(rank_blca = rank(-score))

res_1 <- make_rank_plot(scc_raw_wgbs, "BLCA_4", blca_pooled_rank)
res_2 <- make_rank_plot(scc_raw_wgbs, "BLCA_1", blca_pooled_rank)
p <- res_1$plot / res_2$plot
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/10_performance_refbased_unmerged"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
file_output <- paste0(output_dir,"/correlation_scc_ranks_blca.pdf")
ggsave(file_output, p, width = 12, height = 12)
```

#### BRCA

```{r}
make_rank_plot <- function(data, tumor_label, pooled_rank) {

  df_rank <- data %>%
    filter(tumor_type == tumor_label) %>%
    group_by(deconv_tool) %>%
    summarise(score = median(SCC, na.rm = TRUE), .groups = "drop") %>%
    mutate(rank_target = rank(-score)) %>%
    inner_join(pooled_rank, by = "deconv_tool")

  cor_val <- cor(df_rank$rank_blca,
                 df_rank$rank_target,
                 method = "spearman",
                 use = "complete.obs")

  p <- ggplot(df_rank, aes(rank_blca, rank_target)) +
    geom_point(size = 3) +
    geom_text(aes(label = deconv_tool), vjust = -0.5, size = 6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    theme_bw(base_size = 25) +
    scale_x_continuous(breaks = 1:10, limits = c(1, 10)) +
    scale_y_continuous(breaks = 1:10, limits = c(1, 10)) +
    labs(
      x = "Rank in BRCA (pooled)",
      y = paste0("Rank in ", tumor_label)
    ) +
    annotate(
      "text",
      x = 2, y = 10,
      label = paste0("Spearman coeff = ", round(cor_val, 3)),
      hjust = 0, size = 7
    )

  list(plot = p, cor = cor_val)
}
```


```{r}
blca_pooled_rank <- scc_raw_wgbs %>%
  filter(tumor_type == "BRCA") %>%
  group_by(deconv_tool) %>%
  summarise(score = median(SCC, na.rm = TRUE), .groups = "drop") %>%
  mutate(rank_blca = rank(-score))

res_1 <- make_rank_plot(scc_raw_wgbs, "BRCA_3", blca_pooled_rank)
res_2 <- make_rank_plot(scc_raw_wgbs, "BRCA_5", blca_pooled_rank)
p <- res_1$plot / res_2$plot
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/10_performance_refbased_unmerged"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
file_output <- paste0(output_dir,"/correlation_scc_ranks_brca.pdf")
ggsave(file_output, p, width = 12, height = 12)
```

#### LUAD

```{r}
make_rank_plot <- function(data, tumor_label, pooled_rank) {

  df_rank <- data %>%
    filter(tumor_type == tumor_label) %>%
    group_by(deconv_tool) %>%
    summarise(score = median(SCC, na.rm = TRUE), .groups = "drop") %>%
    mutate(rank_target = rank(-score)) %>%
    inner_join(pooled_rank, by = "deconv_tool")

  cor_val <- cor(df_rank$rank_blca,
                 df_rank$rank_target,
                 method = "spearman",
                 use = "complete.obs")

  p <- ggplot(df_rank, aes(rank_blca, rank_target)) +
    geom_point(size = 3) +
    geom_text(aes(label = deconv_tool), vjust = -0.5, size = 6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    theme_bw(base_size = 25) +
    scale_x_continuous(breaks = 1:10, limits = c(1, 10)) +
    scale_y_continuous(breaks = 1:10, limits = c(1, 10)) +
    labs(
      x = "Rank in LUAD (pooled)",
      y = paste0("Rank in ", tumor_label)
    ) +
    annotate(
      "text",
      x = 2, y = 10,
      label = paste0("Spearman coeff = ", round(cor_val, 3)),
      hjust = 0, size = 7
    )

  list(plot = p, cor = cor_val)
}
```


```{r}
blca_pooled_rank <- scc_raw_wgbs %>%
  filter(tumor_type == "LUAD") %>%
  group_by(deconv_tool) %>%
  summarise(score = median(SCC, na.rm = TRUE), .groups = "drop") %>%
  mutate(rank_blca = rank(-score))

res_1 <- make_rank_plot(scc_raw_wgbs, "LUAD_1", blca_pooled_rank)
res_2 <- make_rank_plot(scc_raw_wgbs, "LUAD_5", blca_pooled_rank)
p <- res_1$plot / res_2$plot
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/10_performance_refbased_unmerged"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
file_output <- paste0(output_dir,"/correlation_scc_ranks_luad.pdf")
ggsave(file_output, p, width = 12, height = 12)
```

#### LUSC

```{r}
make_rank_plot <- function(data, tumor_label, pooled_rank) {

  df_rank <- data %>%
    filter(tumor_type == tumor_label) %>%
    group_by(deconv_tool) %>%
    summarise(score = median(SCC, na.rm = TRUE), .groups = "drop") %>%
    mutate(rank_target = rank(-score)) %>%
    inner_join(pooled_rank, by = "deconv_tool")

  cor_val <- cor(df_rank$rank_blca,
                 df_rank$rank_target,
                 method = "spearman",
                 use = "complete.obs")

  p <- ggplot(df_rank, aes(rank_blca, rank_target)) +
    geom_point(size = 3) +
    geom_text(aes(label = deconv_tool), vjust = -0.5, size = 6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    theme_bw(base_size = 25) +
    scale_x_continuous(breaks = 1:10, limits = c(1, 10)) +
    scale_y_continuous(breaks = 1:10, limits = c(1, 10)) +
    labs(
      x = "Rank in LUSC (pooled)",
      y = paste0("Rank in ", tumor_label)
    ) +
    annotate(
      "text",
      x = 2, y = 10,
      label = paste0("Spearman coeff = ", round(cor_val, 3)),
      hjust = 0, size = 7
    )

  list(plot = p, cor = cor_val)
}
```


```{r}
blca_pooled_rank <- scc_raw_wgbs %>%
  filter(tumor_type == "LUSC") %>%
  group_by(deconv_tool) %>%
  summarise(score = median(SCC, na.rm = TRUE), .groups = "drop") %>%
  mutate(rank_blca = rank(-score))

res_1 <- make_rank_plot(scc_raw_wgbs, "LUSC_4", blca_pooled_rank)
res_2 <- make_rank_plot(scc_raw_wgbs, "LUSC_5", blca_pooled_rank)
p <- res_1$plot / res_2$plot
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/10_performance_refbased_unmerged"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
file_output <- paste0(output_dir,"/correlation_scc_ranks_lusc.pdf")
ggsave(file_output, p, width = 12, height = 12)
```


### Check correlation between RMSE and tumor purity

We now want to investigate whether tumor purity of the samples analysed in this
study and estimated in the [following study](https://www.nature.com/articles/ncomms9971#Bib1) 
has an association with the RMSE.

```{r}
rmse_raw_wgbs <- fin.res %>%
  filter(tumor_type %in% c("BLCA_1","BLCA_4","BRCA_3","BRCA_5",
                           "LUAD_1","LUAD_5","LUSC_4","LUSC_5")) %>%
  group_by(deconv_tool, dmr_tool, seq_depth, tumor_type) %>%
  summarize("1-RMSE" = 1-rmse(expected_tf, predicted_tf))
tumor_purity <- read.csv("../resources/tumor_purity.csv")
purity.res.rmse <- merge(rmse_raw_wgbs, tumor_purity, by = c("tumor_type"))
```

Here we investigate the association with the tumor purity estimation using the
CPE metric, which is computed as a consensus metric of other 4 metrics (see
study above).

```{r}
# Basic scatter plot with facets
p <- ggplot(purity.res.rmse, aes(x = CPE, y = `1-RMSE`)) +
  geom_jitter(aes(color = tumor_type), size = 1, alpha = 0.8) +
  geom_boxplot(aes(group = tumor_type), median.colour = "grey20",color = "grey20", outliers = F) +
  theme_bw() +
  theme(text = element_text(size = 12)) +
  labs(x = "Purity (CPE)", y = "1-RMSE") + 
  stat_cor(
    aes(label = paste(..r.label.., ..p.label.., sep = "~`,`~")),
    method = "spearman",
    label.x = 0.9,
    label.y = 0.1
  )
p
```


```{r}
# Create the main directory to save the plots
output_dir <- "../plots/10_performance_refbased_unmerged"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
file_output <- paste0(output_dir,"/cpe_rmse_corr.pdf")
ggsave(file_output, p, width = 10, height = 6)
```


Here we investigate the association with the tumor purity estimation using the
ESTIMATE metric, which estimates the tumor purity using a set of genes specific 
for stromal and immune cells (see study above).


```{r}
# Basic scatter plot with facets
p <- ggplot(purity.res.rmse, aes(x = ESTIMATE, y = `1-RMSE`)) +
  geom_jitter(aes(color = tumor_type), size = 1, alpha = 0.8) +
  geom_boxplot(aes(group = tumor_type), median.colour = "grey20",color = "grey20", outliers = F) +
  theme_bw() +
  theme(text = element_text(size = 12)) +
  labs(x = "Purity (ESTIMATE)", y = "1-RMSE") + 
  stat_cor(
    aes(label = paste(..r.label.., ..p.label.., sep = "~`,`~")),
    method = "spearman",
    label.x = 0.9,
    label.y = 0.1
  )
p
```

```{r}
# Create the main directory to save the plots
output_dir <- "../plots/10_performance_refbased_unmerged"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = T)

# Combine and draw
file_output <- paste0(output_dir,"/estimate_rmse_corr.pdf")
ggsave(file_output, p, width = 10, height = 6)
```










