---
title: "11. Ranking robustness: rank-aggregation sensitivity and bootstrap stability"
author: "Edoardo Giuili"
date: "2026-07-16"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Introduction

This notebook explores how robust the final deconvolution-tool ranking
is. It is purely additive to `06_funkyheatmap.Rmd`: the existing
**aggregate-then-rank** ranking (min-max scale each metric within a
combination, average, then rank the averages -- i.e. `overall_score` in
06) is *not* recomputed or reimplemented here. Instead we execute 06's
own code verbatim (via `knitr::purl()` + `source()`) and reuse its
objects directly, so there is exactly one implementation of that logic
in the whole repository.

On top of that fixed reference ranking, this notebook adds:

1.  A complementary **rank-then-aggregate (case-based)** scheme, in the
    sense of Maier-Hein et al. 2018 ("Why rankings of biomedical image
    analysis competitions should be interpreted with care", *Nat
    Commun*): for each metric (RMSE, SCC, AUC, LoD) and each *case* (a
    combination = one tumor_type x seq_depth within a dataset), tools
    are ranked directly on the **raw, unscaled** metric value -- no
    min-max scaling anywhere, since ranks are already scale-free. The
    per-metric ranks within a case are averaged into one consensus
    per-case rank, and those per-case ranks are aggregated across cases
    (mean, default; median exposed as a sensitivity check) using the
    same dataset weights as 06. This is the classical "rank, then
    aggregate" (case-based) scheme, contrasted with 06's "aggregate,
    then rank" (metric-based) scheme, which min-max scales and averages
    the metrics *before* ranking.
2.  **Kendall's tau** between the two schemes, as a single robustness
    number.
3.  A **bootstrap over combinations** (resampling the (dataset x
    tumor_type x seq_depth) blocks with replacement, 1000x), recomputing
    both rankings each time, to report each tool's rank distribution and
    % of resamples ranked 1st.
4.  A **Friedman test + Nemenyi post-hoc** on the complete combination x
    tool blocks.
5.  Plots: a bootstrap rank-distribution "blob" plot (dot size =
    frequency of rank, in the style of Fig 2B) and a scheme-comparison
    plot.

The **resampling/blocking unit** throughout is a *combination*: one
(dataset x tumor_type x seq_depth) triple. Every combination is a
complete block containing a value (a `Score`, and separately the raw
RMSE/SCC/AUC used for the case-based scheme) for all 12 deconvolution
tools (both reference-based and the two reference-free tools, MeDeCom
and RefFreeCellMix).

### Setting up the environment

```{r setup-07}
knitr::opts_knit$set(root.dir = './')
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(PMCMRplus)
library(tibble)
library(purrr)
library(ggrepel)
set.seed(42)
```

## Step 1: reuse 06's aggregate-then-rank ranking, unchanged

We do not copy or reimplement any of 06's scoring logic. Instead, we
extract 06's R code chunks with `knitr::purl()` and `source()` them,
which executes the notebook's own code exactly as written. We only need
everything **up to and including the `overall_score` computation**; the
chunks after that in 06 deal purely with plotting the funkyheatmap
(building the `funky_heatmap()` figure, which needs the `funkyheatmap`
package and logo image files) and are not needed here, so we stop
sourcing right before the `palettes <- tribble(...)` chunk.

```{r message=FALSE, warning=FALSE}
notebook_06_path <- "06_funkyheatmap.Rmd"
purled_06_path <- tempfile(fileext = ".R")

invisible(knitr::purl(notebook_06_path, output = purled_06_path,
                      documentation = 0, quiet = TRUE))

lines_06 <- readLines(purled_06_path)
# Everything from here on in 06 is funkyheatmap plotting code (funky_heatmap(),
# logos, ggsave); we stop just before it so we don't need the funkyheatmap
# package or the logo PNGs to obtain the reference ranking.
plot_cutoff <- grep("^palettes <- tribble\\(", lines_06)[1]
stopifnot(!is.na(plot_cutoff))
lines_06_scoring <- lines_06[seq_len(plot_cutoff - 1)]

# 06 reuses the variable name `aucroc_data` for each of the three datasets in
# turn (wgbs, then rrbs, then cfrrbs), so only the last one would survive in
# env_06. To also get the raw (pre-min-max-scaled) per-case AUC values needed
# for the case-based scheme below -- without re-running 06's expensive nested
# AUC loop a second time -- we snapshot `aucroc_data` into a dataset-specific
# name right before each point where 06 itself consumes it (i.e. right before
# each `auc_<dataset> <- aucroc_data %>% ...` line). This does not alter any
# of 06's own computations, it just keeps an extra copy of an intermediate
# object 06 already produces.
snapshot_targets <- c("auc_wgbs <- aucroc_data", "auc_rrbs <- aucroc_data", "auc_cfrrbs <- aucroc_data")
snapshot_names <- c("aucroc_data_wgbs", "aucroc_data_rrbs", "aucroc_data_cfrrbs")
for (i in seq_along(snapshot_targets)) {
  target_idx <- grep(paste0("^", snapshot_targets[i]), lines_06_scoring, fixed = FALSE)[1]
  stopifnot(!is.na(target_idx))
  lines_06_scoring <- append(
    lines_06_scoring,
    paste0(snapshot_names[i], " <- aucroc_data"),
    after = target_idx - 1
  )
}

env_06 <- new.env()
source(textConnection(lines_06_scoring), local = env_06)
```

This gives us, from 06, unchanged: `bench` (the harmonized ref-based +
ref-free long table), `in_silico_wgbs` / `in_silico_rrbs` /
`in_silico_cfrrbs` (the raw per-sample predictions), `aucroc_data_wgbs`
/ `_rrbs` / `_cfrrbs` (raw per-case, per-tumor-fraction AUC, i.e. 06's
own `aucroc_data` snapshotted per dataset), `lod.stats` (raw per-sample
LoD statistics), `score_wgbs` / `score_rrbs` / `score_cfrrbs` (06's
per-combination, within-combination min-max-scaled `Score`), and
`overall_score` (06's final aggregate-then-rank ranking with dataset
weights WGBS = 0.45, RRBS = 0.225, cfRRBS = 0.225, plus scalability =
0.1). The case-based scheme below uses the raw AUC/SCC/RMSE/LoD values
(i.e. computed the same way as 06, just read off *before* 06's
min-max-scaling step) rather than
`score_wgbs`/`score_rrbs`/`score_cfrrbs`.

```{r}
reference_ranking <- env_06$overall_score %>%
  arrange(desc(Overall_Score)) %>%
  mutate(Rank_reference = rank(-Overall_Score, ties.method = "average")) %>%
  select(deconv_tool, Overall_Score, Rank_reference)

reference_ranking
```

We reuse 06's exact coded weights (0.45 / 0.225 / 0.225 / 0.1) below for
the rank-then-aggregate scheme, so both schemes are driven by identical
weights.

```{r}
weights_used <- tibble::tibble(
  dataset = c("wgbs", "rrbs", "cfrrbs", "scalability"),
  weight = c(0.45, 0.225, 0.225, 0.1)
)
weights_used
```

## Step 2: build the long combination-level table

### 2a. Score-level table (used for the Friedman/Nemenyi block matrix)

Each of `score_wgbs`, `score_rrbs`, `score_cfrrbs` is a per-combination
table: one row per (deconv_tool, tumor_type, seq_depth), with `Score`
min-max scaled within that combination. This is only used later for the
Friedman/Nemenyi test (which needs a single combination x tool value
matrix); the case-based ranking scheme itself is built from the raw
per-metric values in 2b, not from `Score`.

```{r}
combo_scores <- bind_rows(
  env_06$score_wgbs   %>% mutate(dataset = "wgbs",   weight = 0.45),
  env_06$score_rrbs   %>% mutate(dataset = "rrbs",   weight = 0.225),
  env_06$score_cfrrbs %>% mutate(dataset = "cfrrbs", weight = 0.225)
) %>%
  mutate(
    seq_depth = as.character(seq_depth),
    combination_id = paste(dataset, tumor_type, seq_depth, sep = "_")
  ) %>%
  select(combination_id, dataset, tumor_type, seq_depth, deconv_tool, Score, weight)

n_tools <- n_distinct(combo_scores$deconv_tool)

combo_completeness <- combo_scores %>%
  count(combination_id, dataset, weight, name = "n_tools_present")

cat("Total combinations (dataset x tumor_type x seq_depth):",
    n_distinct(combo_scores$combination_id), "\n")
cat("Total tools:", n_tools, "\n")
cat("Combinations with all", n_tools, "tools present (complete blocks):",
    sum(combo_completeness$n_tools_present == n_tools), "/",
    nrow(combo_completeness), "\n")

head(combo_scores)
```

### 2b. Raw per-metric, per-case table (used for the case-based scheme)

For the case-based scheme we need the **raw, unscaled** RMSE / SCC / AUC
per case, i.e. exactly the intermediate values 06 computes right before
its min-max-scaling step (`meanAUC` before `AUC_minmax`, `SCC` before
`SCC_minmax`, per-case mean `RMSE` before `RMSE_minmax`). 06 itself does
not keep these intermediates around (they get overwritten by the scaled
version in the same pipe), so we recompute them here using 06's own
harmonized data (`env_06$in_silico_wgbs/rrbs/cfrrbs`) and 06's own
snapshotted `aucroc_data_wgbs/rrbs/cfrrbs` (see above) -- i.e. the same
inputs and the same per-case grouping as 06, just reading the value off
one pipe step earlier, so no min-max scaling is applied anywhere in this
scheme.

RMSE is an error (lower = better); SCC and AUC are higher = better. LoD
(the detectable tumor fraction, lower = better) is available in 06's
`lod.stats` table (built at 06 lines 223-224 and extended per dataset)
at per-(deconv_tool, dmr_tool, tumor_type, seq_depth) granularity. 06
only pools it to a single per-(deconv_tool, dataset) value when building
`Score`, but the raw `lod.stats` is actually resolved per case, so for
the case-based scheme we use it at the case level: we average `lod` over
`dmr_tool` to get one LoD per (deconv_tool, tumor_type, seq_depth), then
rank it within each case alongside RMSE/SCC/AUC. This lets LoD
contribute genuine between-case variation to the per-case consensus
rank, rather than being folded in as a dataset-level constant.

```{r}
rmse_case <- function(actual, predicted) sqrt(mean((actual - predicted)^2))

compute_raw_metrics <- function(in_silico_df, aucroc_df) {
  raw_auc <- aucroc_df %>%
    group_by(deconv_tool, seq_depth, tumor_type, expected_tf) %>%
    summarise(AUC_per_fraction = mean(auc), .groups = "drop") %>%
    group_by(deconv_tool, seq_depth, tumor_type) %>%
    summarise(AUC = mean(AUC_per_fraction), .groups = "drop")

  raw_scc <- in_silico_df %>%
    group_by(deconv_tool, seq_depth, tumor_type) %>%
    summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop")

  raw_rmse <- in_silico_df %>%
    filter(expected_tf != 0) %>%
    group_by(deconv_tool, seq_depth, tumor_type, expected_tf) %>%
    summarise(RMSE_per_fraction = rmse_case(expected_tf, predicted_tf), .groups = "drop") %>%
    group_by(deconv_tool, seq_depth, tumor_type) %>%
    summarise(RMSE = mean(RMSE_per_fraction), .groups = "drop")

  raw_auc %>%
    inner_join(raw_scc, by = c("deconv_tool", "seq_depth", "tumor_type")) %>%
    inner_join(raw_rmse, by = c("deconv_tool", "seq_depth", "tumor_type"))
}

raw_metrics_wgbs   <- compute_raw_metrics(env_06$in_silico_wgbs,   env_06$aucroc_data_wgbs)
raw_metrics_rrbs   <- compute_raw_metrics(env_06$in_silico_rrbs,   env_06$aucroc_data_rrbs)
raw_metrics_cfrrbs <- compute_raw_metrics(env_06$in_silico_cfrrbs, env_06$aucroc_data_cfrrbs)

# Raw (pre-scaling) LoD, per case: 06's `lod.stats` is resolved per
# (deconv_tool, dmr_tool, tumor_type, seq_depth). We collapse over dmr_tool to
# one LoD per (deconv_tool, tumor_type, seq_depth) so it can be ranked within
# each case alongside the other metrics.
raw_lod_case <- env_06$lod.stats %>%
  mutate(lod = as.numeric(lod), seq_depth = as.character(seq_depth)) %>%
  group_by(deconv_tool, tumor_type, seq_depth) %>%
  summarise(LoD = mean(lod, na.rm = TRUE), .groups = "drop")

raw_metrics_long <- bind_rows(
  raw_metrics_wgbs   %>% mutate(dataset = "wgbs",   weight = 0.45),
  raw_metrics_rrbs   %>% mutate(dataset = "rrbs",   weight = 0.225),
  raw_metrics_cfrrbs %>% mutate(dataset = "cfrrbs", weight = 0.225)
) %>%
  mutate(seq_depth = as.character(seq_depth)) %>%
  left_join(raw_lod_case, by = c("deconv_tool", "tumor_type", "seq_depth")) %>%
  mutate(combination_id = paste(dataset, tumor_type, seq_depth, sep = "_")) %>%
  select(combination_id, dataset, tumor_type, seq_depth, deconv_tool, weight, RMSE, SCC, AUC, LoD)

head(raw_metrics_long)
```

## Step 3: rank-then-aggregate (case-based) scheme

For each case (combination) we rank tools separately on each metric's
raw value -- **higher-is-better** for SCC and AUC, **lower-is-better**
for RMSE and LoD -- then average the (up to) 4 per-metric ranks into a
single per-case consensus rank. Those per-case ranks are then aggregated
across cases with the same dataset weights as 06 (weighted mean,
default; weighted median exposed as a sensitivity check via `agg_fun`).
This mirrors the case-based ("rank, then aggregate") scheme as used e.g.
in Maier-Hein et al. 2018.

```{r}
per_case_consensus_rank <- function(metrics_long) {
  # Rank keeping NAs as NA (na.last = "keep") so a metric that is missing for a
  # tool in a case does not get silently ranked "worst"; case_rank then
  # averages only the metrics actually available in that case.
  metrics_long %>%
    group_by(combination_id) %>%
    mutate(
      rank_RMSE = rank(RMSE,  ties.method = "average", na.last = "keep"),  # lower RMSE = better
      rank_SCC  = rank(-SCC,  ties.method = "average", na.last = "keep"),  # higher SCC = better
      rank_AUC  = rank(-AUC,  ties.method = "average", na.last = "keep"),  # higher AUC = better
      rank_LoD  = rank(LoD,   ties.method = "average", na.last = "keep")   # lower LoD = better
    ) %>%
    ungroup() %>%
    rowwise() %>%
    mutate(case_rank = mean(c(rank_RMSE, rank_SCC, rank_AUC, rank_LoD), na.rm = TRUE)) %>%
    ungroup()
}

weighted_median <- function(x, w) {
  ord <- order(x)
  x <- x[ord]; w <- w[ord]
  cum_w <- cumsum(w) / sum(w)
  x[which(cum_w >= 0.5)[1]]
}

rank_then_aggregate <- function(metrics_long, agg_fun = mean) {
  ranked <- per_case_consensus_rank(metrics_long)

  ranked %>%
    group_by(deconv_tool) %>%
    summarise(
      Agg_Rank = if (identical(agg_fun, mean)) {
        sum(case_rank * weight) / sum(weight)
      } else if (identical(agg_fun, median)) {
        weighted_median(case_rank, weight)
      } else {
        agg_fun(case_rank, weight)
      },
      .groups = "drop"
    ) %>%
    arrange(Agg_Rank) %>%
    mutate(Rank_rta = rank(Agg_Rank, ties.method = "average"))
}

rta_ranking_mean <- rank_then_aggregate(raw_metrics_long, agg_fun = mean)
rta_ranking_mean
```

Sensitivity check with `median` instead of `mean` (per Maier-Hein et
al., mean-based aggregation after ranking is expected to be the more
robust of the two):

```{r}
rta_ranking_median <- rank_then_aggregate(raw_metrics_long, agg_fun = median)
rta_ranking_median
```

## Step 4: compare the two schemes

```{r}
scheme_comparison <- reference_ranking %>%
  select(deconv_tool, Rank_reference) %>%
  inner_join(rta_ranking_mean %>% select(deconv_tool, Rank_rta), by = "deconv_tool") %>%
  inner_join(
    rta_ranking_median %>% select(deconv_tool, Rank_rta_median = Rank_rta),
    by = "deconv_tool"
  ) %>%
  arrange(Rank_reference)

scheme_comparison
```

Kendall's tau between the aggregate-then-rank (reference) and
rank-then-aggregate (mean) schemes, as a single robustness number:

```{r}
kendall_tau <- cor(
  scheme_comparison$Rank_reference,
  scheme_comparison$Rank_rta,
  method = "kendall"
)

kendall_test <- cor.test(
  scheme_comparison$Rank_reference,
  scheme_comparison$Rank_rta,
  method = "kendall"
)

cat(sprintf("Kendall's tau (aggregate-then-rank vs. rank-then-aggregate, mean): %.3f (p = %.4g)\n",
            kendall_tau, kendall_test$p.value))

kendall_tau_median <- cor(
  scheme_comparison$Rank_reference,
  scheme_comparison$Rank_rta_median,
  method = "kendall"
)
cat(sprintf("Kendall's tau (aggregate-then-rank vs. rank-then-aggregate, median): %.3f\n",
            kendall_tau_median))
```

## Step 5: bootstrap over combinations

We resample the `combination_id` blocks with replacement (1000
replicates). Because the same combination can be drawn more than once in
a replicate, each draw is tagged with a unique suffix (`__b1`, `__b2`,
...) so duplicated blocks are treated as separate blocks rather than
being silently collapsed by a `group_by`. Both schemes
(aggregate-then-rank and rank-then-aggregate) are recomputed from
scratch on every bootstrap replicate.

```{r}
combination_ids <- unique(combo_scores$combination_id)
stopifnot(setequal(combination_ids, unique(raw_metrics_long$combination_id)))
n_boot <- 1000

aggregate_then_rank_once <- function(scores_df) {
  # Mirrors 06's overall_score logic exactly (weighted mean of Score across
  # combinations, then rank), but operating on a resampled long table where
  # combination_id may contain duplicate-draw suffixes.
  scores_df %>%
    group_by(deconv_tool) %>%
    summarise(Overall_Score = sum(Score * weight) / sum(weight), .groups = "drop") %>%
    arrange(desc(Overall_Score)) %>%
    mutate(Rank_reference = rank(-Overall_Score, ties.method = "average"))
}

run_bootstrap_replicate <- function(i) {
  # Draw the SAME resample of combinations for both schemes, so any
  # difference between them on a given replicate reflects the aggregation
  # order, not different resampled data.
  drawn <- sample(combination_ids, length(combination_ids), replace = TRUE)
  draw_key <- tibble(
    combination_id = drawn,
    draw_id = paste0(drawn, "__b", seq_along(drawn))
  )

  boot_scores <- combo_scores %>%
    inner_join(draw_key, by = "combination_id", relationship = "many-to-many") %>%
    mutate(combination_id = draw_id) %>%
    select(-draw_id)

  boot_metrics <- raw_metrics_long %>%
    inner_join(draw_key, by = "combination_id", relationship = "many-to-many") %>%
    mutate(combination_id = draw_id) %>%
    select(-draw_id)

  ref_rank <- aggregate_then_rank_once(boot_scores) %>%
    transmute(deconv_tool, replicate = i, scheme = "aggregate_then_rank", Rank = Rank_reference)
  rta_rank <- rank_then_aggregate(boot_metrics, agg_fun = mean) %>%
    transmute(deconv_tool, replicate = i, scheme = "rank_then_aggregate", Rank = Rank_rta)

  bind_rows(ref_rank, rta_rank)
}

boot_results <- purrr::map_dfr(seq_len(n_boot), run_bootstrap_replicate)
```

```{r}
rank_distribution <- boot_results %>%
  count(scheme, deconv_tool, Rank, name = "freq") %>%
  group_by(scheme, deconv_tool) %>%
  mutate(pct = 100 * freq / sum(freq)) %>%
  ungroup()

pct_rank1 <- boot_results %>%
  filter(Rank == 1) %>%
  count(scheme, deconv_tool, name = "n_rank1") %>%
  mutate(pct_rank1 = 100 * n_rank1 / n_boot) %>%
  select(-n_rank1)

rank1_summary <- reference_ranking %>%
  select(deconv_tool, Rank_reference) %>%
  left_join(pct_rank1 %>% filter(scheme == "aggregate_then_rank") %>%
              select(deconv_tool, pct_rank1_aggregate_then_rank = pct_rank1),
            by = "deconv_tool") %>%
  left_join(pct_rank1 %>% filter(scheme == "rank_then_aggregate") %>%
              select(deconv_tool, pct_rank1_rank_then_aggregate = pct_rank1),
            by = "deconv_tool") %>%
  mutate(across(starts_with("pct_rank1"), ~replace_na(.x, 0))) %>%
  arrange(Rank_reference)

rank1_summary
```

## Step 6: Friedman test + Nemenyi post-hoc

We run the Friedman test (deconvolution tools = treatments, combinations
= blocks) followed by the Nemenyi post-hoc test on complete blocks only,
for **both** schemes: once on 06's min-max-scaled `Score` (reference,
aggregate-then-rank), and once on the raw per-case consensus rank used
for the case-based scheme (Step 3), since the two are on different
scales and shouldn't be tested with a single shared matrix.

```{r}
build_block_matrix <- function(long_df, value_col) {
  block_matrix_df <- long_df %>%
    select(combination_id, deconv_tool, all_of(value_col)) %>%
    pivot_wider(names_from = deconv_tool, values_from = all_of(value_col))

  complete_rows <- complete.cases(block_matrix_df %>% select(-combination_id))
  block_matrix_complete <- block_matrix_df[complete_rows, ]
  block_matrix <- as.matrix(block_matrix_complete %>% select(-combination_id))
  rownames(block_matrix) <- block_matrix_complete$combination_id

  list(
    matrix = block_matrix,
    n_total = nrow(block_matrix_df),
    n_complete = sum(complete_rows)
  )
}

run_friedman_nemenyi <- function(block_info, label) {
  cat(sprintf("[%s] Combinations available: %d. Complete blocks (all %d tools present): %d.\n",
              label, block_info$n_total, n_tools, block_info$n_complete))

  if (block_info$n_complete < 2) {
    cat(sprintf("[%s] Fewer than 2 complete blocks available -- Friedman/Nemenyi skipped.\n", label))
    return(invisible(NULL))
  }

  friedman_result <- friedman.test(block_info$matrix)
  print(friedman_result)
  print(PMCMRplus::frdAllPairsNemenyiTest(block_info$matrix))
  invisible(friedman_result)
}
```

### 6a. Aggregate-then-rank (reference, min-max-scaled `Score`)

```{r}
block_info_score <- build_block_matrix(combo_scores, "Score")
n_total_combinations <- block_info_score$n_total
n_complete_combinations <- block_info_score$n_complete
invisible(run_friedman_nemenyi(block_info_score, "aggregate-then-rank (Score)"))
```

### 6b. Rank-then-aggregate (case-based, raw per-case consensus rank)

```{r}
case_ranks_wide <- per_case_consensus_rank(raw_metrics_long) %>%
  select(combination_id, deconv_tool, case_rank)

block_info_case_rank <- build_block_matrix(case_ranks_wide, "case_rank")
invisible(run_friedman_nemenyi(block_info_case_rank, "rank-then-aggregate (case_rank)"))
```

## Step 7: plots

### Bootstrap rank-distribution blob plot

Dot size = frequency (%) with which a tool received a given rank across
the 1000 bootstrap replicates, faceted by aggregation scheme, in the
style of Fig 2B. Tools are ordered by their reference
(aggregate-then-rank) rank.

```{r fig.width=9, fig.height=6}
tool_order <- reference_ranking %>% arrange(Rank_reference) %>% pull(deconv_tool)

rank_distribution_plot_df <- rank_distribution %>%
  mutate(
    deconv_tool = factor(deconv_tool, levels = rev(tool_order)),
    scheme = dplyr::recode(scheme,
                     aggregate_then_rank = "Aggregate-then-rank (reference)",
                     rank_then_aggregate = "Rank-then-aggregate (mean rank)")
  )

p_blob <- ggplot(rank_distribution_plot_df, aes(x = Rank, y = deconv_tool, size = pct)) +
  geom_point(alpha = 0.75, color = "#2c7fb8") +
  scale_size_continuous(name = "% of\nresamples", range = c(0.5, 12)) +
  scale_x_continuous(breaks = seq_len(n_tools)) +
  facet_wrap(~scheme) +
  labs(
    title = "Bootstrap rank-distribution across combinations (1000 resamples)",
    x = "Rank (1 = best)",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())

p_blob
```

```{r}
output_dir <- "../plots/11_ranking_robustness"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
ggsave(file.path(output_dir, "bootstrap_rank_distribution_blobplot.svg"),
       p_blob, width = 12, height = 6)
```

### Scheme-comparison plot

Reference (aggregate-then-rank) rank vs. rank-then-aggregate (mean) rank
for each tool, with the identity line shown for reference; points
falling on the diagonal indicate perfect agreement between the two
schemes.

```{r fig.width=7, fig.height=6}
p_scheme <- ggplot(scheme_comparison, aes(x = Rank_reference, y = Rank_rta, label = deconv_tool)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(size = 3, color = "#e34a33") +
  ggrepel::geom_text_repel(size = 3.5, seed = 42) +
  scale_x_continuous(breaks = seq_len(n_tools)) +
  scale_y_continuous(breaks = seq_len(n_tools)) +
  labs(
    title = sprintf("Aggregate-then-rank vs. rank-then-aggregate (Kendall's tau = %.3f)", kendall_tau),
    x = "Rank -- aggregate-then-rank (reference)",
    y = "Rank -- rank-then-aggregate (mean rank)"
  ) +
  coord_equal() +
  theme_minimal(base_size = 12)

p_scheme
```

```{r}
ggsave(file.path(output_dir, "scheme_comparison.svg"), p_scheme, width = 8, height = 6)
```

## Summary

```{r}
cat(sprintf(
  "Reference ranking (06, aggregate-then-rank) vs. rank-then-aggregate (mean): Kendall's tau = %.3f\n",
  kendall_tau
))
cat(sprintf(
  "Reference ranking vs. rank-then-aggregate (median, sensitivity check): Kendall's tau = %.3f\n",
  kendall_tau_median
))
cat(sprintf(
  "Bootstrap: %d replicates over %d combinations (dataset x tumor_type x seq_depth).\n",
  n_boot, length(combination_ids)
))
cat(sprintf(
  "Friedman/Nemenyi: %d / %d combinations used as complete blocks.\n",
  n_complete_combinations, n_total_combinations
))
```
