---
title: "11. Ranking robustness: rank-aggregation sensitivity and bootstrap stability"
author: "Edoardo Giuili"
date: "2026-07-16"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Introduction

This notebook tests how robust the final tool ranking is. The reference
**aggregate-then-rank** ranking (`overall_score` in
`06_funkyheatmap.Rmd`) is not reimplemented: we run 06's own code and
reuse its objects. On top of it, we add:

1.  A **rank-then-aggregate (case-based)** scheme (Maier-Hein et al.
    2018, *Nat Commun*): within each case (dataset x tumor_type x
    seq_depth), tools are ranked on the raw RMSE, SCC, AUC and LoD
    values, the per-metric ranks are averaged, and the per-case ranks
    are aggregated across cases with 06's dataset weights (mean; median
    as a sensitivity check).
2.  **Kendall's tau** between the two schemes.
3.  A **bootstrap over combinations** (1000x), reporting each tool's
    rank distribution and % of resamples ranked 1st.
4.  A **hierarchy-aware analysis**: a cluster bootstrap over biological
    sources (tumor types), plus leave-one-tumor-type-out and
    leave-one-dataset-out.
5.  A **dataset-weight sensitivity analysis**: both rankings with equal
    dataset weights.
6.  A **Friedman test + Nemenyi post-hoc**.
7.  Plots of the above.

Unless stated otherwise, the resampling/blocking unit is a
*combination* (dataset x tumor_type x seq_depth). All 36 combinations
are complete blocks with values for all 12 tools.

### Setting up the environment

```{r setup-07}
knitr::opts_knit$set(root.dir = './')
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(PMCMRplus)
library(tibble)
library(purrr)
library(ggrepel)
set.seed(42)
```

## Step 1: reuse 06's aggregate-then-rank ranking, unchanged

We extract 06's code with `knitr::purl()` and `source()` it up to the
`overall_score` computation, stopping before the funkyheatmap plotting
code (`palettes <- tribble(...)`).

```{r message=FALSE, warning=FALSE}
notebook_06_path <- "06_funkyheatmap.Rmd"
purled_06_path <- tempfile(fileext = ".R")

invisible(knitr::purl(notebook_06_path, output = purled_06_path,
                      documentation = 0, quiet = TRUE))

lines_06 <- readLines(purled_06_path)
# Stop before 06's plotting code (needs funkyheatmap and logo files).
plot_cutoff <- grep("^palettes <- tribble\\(", lines_06)[1]
stopifnot(!is.na(plot_cutoff))
lines_06_scoring <- lines_06[seq_len(plot_cutoff - 1)]

# 06 overwrites `aucroc_data` for each dataset. Keep a per-dataset copy (for
# the raw per-case AUC below) without changing any of 06's computations.
snapshot_targets <- c("auc_wgbs <- aucroc_data", "auc_rrbs <- aucroc_data", "auc_cfrrbs <- aucroc_data")
snapshot_names <- c("aucroc_data_wgbs", "aucroc_data_rrbs", "aucroc_data_cfrrbs")
for (i in seq_along(snapshot_targets)) {
  target_idx <- grep(paste0("^", snapshot_targets[i]), lines_06_scoring, fixed = FALSE)[1]
  stopifnot(!is.na(target_idx))
  lines_06_scoring <- append(
    lines_06_scoring,
    paste0(snapshot_names[i], " <- aucroc_data"),
    after = target_idx - 1
  )
}

env_06 <- new.env()
source(textConnection(lines_06_scoring), local = env_06)
```

```{r}
reference_ranking <- env_06$overall_score %>%
  arrange(desc(Overall_Score)) %>%
  mutate(Rank_reference = rank(-Overall_Score, ties.method = "average")) %>%
  select(deconv_tool, Overall_Score, Rank_reference)

reference_ranking
```

Both schemes use 06's dataset weights:

```{r}
weights_used <- tibble::tibble(
  dataset = c("wgbs", "rrbs", "cfrrbs", "scalability"),
  weight = c(0.45, 0.225, 0.225, 0.1)
)
weights_used
```

## Step 2: build the long combination-level table

### 2a. Score-level table (used for the Friedman/Nemenyi block matrix)

06's `score_*` tables: one min-max-scaled `Score` per tool and
combination. Used for the aggregate-then-rank scheme and its Friedman
test.

```{r}
combo_scores <- bind_rows(
  env_06$score_wgbs   %>% mutate(dataset = "wgbs",   weight = 0.45),
  env_06$score_rrbs   %>% mutate(dataset = "rrbs",   weight = 0.225),
  env_06$score_cfrrbs %>% mutate(dataset = "cfrrbs", weight = 0.225)
) %>%
  mutate(
    seq_depth = as.character(seq_depth),
    combination_id = paste(dataset, tumor_type, seq_depth, sep = "_")
  ) %>%
  select(combination_id, dataset, tumor_type, seq_depth, deconv_tool, Score, weight)

n_tools <- n_distinct(combo_scores$deconv_tool)

combo_completeness <- combo_scores %>%
  count(combination_id, dataset, weight, name = "n_tools_present")

cat("Total combinations (dataset x tumor_type x seq_depth):",
    n_distinct(combo_scores$combination_id), "\n")
cat("Total tools:", n_tools, "\n")
cat("Combinations with all", n_tools, "tools present (complete blocks):",
    sum(combo_completeness$n_tools_present == n_tools), "/",
    nrow(combo_completeness), "\n")

head(combo_scores)
```

### 2b. Raw per-metric, per-case table (used for the case-based scheme)

Raw (unscaled) RMSE, SCC, AUC and LoD per case: the values 06 computes
just before min-max scaling, recomputed from 06's own inputs because 06
does not keep them.

```{r}
rmse_case <- function(actual, predicted) sqrt(mean((actual - predicted)^2))

compute_raw_metrics <- function(in_silico_df, aucroc_df) {
  raw_auc <- aucroc_df %>%
    group_by(deconv_tool, seq_depth, tumor_type, expected_tf) %>%
    summarise(AUC_per_fraction = mean(auc), .groups = "drop") %>%
    group_by(deconv_tool, seq_depth, tumor_type) %>%
    summarise(AUC = mean(AUC_per_fraction), .groups = "drop")

  raw_scc <- in_silico_df %>%
    group_by(deconv_tool, seq_depth, tumor_type) %>%
    summarise(SCC = cor(expected_tf, predicted_tf, method = "spearman"), .groups = "drop")

  raw_rmse <- in_silico_df %>%
    filter(expected_tf != 0) %>%
    group_by(deconv_tool, seq_depth, tumor_type, expected_tf) %>%
    summarise(RMSE_per_fraction = rmse_case(expected_tf, predicted_tf), .groups = "drop") %>%
    group_by(deconv_tool, seq_depth, tumor_type) %>%
    summarise(RMSE = mean(RMSE_per_fraction), .groups = "drop")

  raw_auc %>%
    inner_join(raw_scc, by = c("deconv_tool", "seq_depth", "tumor_type")) %>%
    inner_join(raw_rmse, by = c("deconv_tool", "seq_depth", "tumor_type"))
}

raw_metrics_wgbs   <- compute_raw_metrics(env_06$in_silico_wgbs,   env_06$aucroc_data_wgbs)
raw_metrics_rrbs   <- compute_raw_metrics(env_06$in_silico_rrbs,   env_06$aucroc_data_rrbs)
raw_metrics_cfrrbs <- compute_raw_metrics(env_06$in_silico_cfrrbs, env_06$aucroc_data_cfrrbs)

# Raw LoD per case: average 06's `lod.stats` over dmr_tool.
raw_lod_case <- env_06$lod.stats %>%
  mutate(lod = as.numeric(lod), seq_depth = as.character(seq_depth)) %>%
  group_by(deconv_tool, tumor_type, seq_depth) %>%
  summarise(LoD = mean(lod, na.rm = TRUE), .groups = "drop")

raw_metrics_long <- bind_rows(
  raw_metrics_wgbs   %>% mutate(dataset = "wgbs",   weight = 0.45),
  raw_metrics_rrbs   %>% mutate(dataset = "rrbs",   weight = 0.225),
  raw_metrics_cfrrbs %>% mutate(dataset = "cfrrbs", weight = 0.225)
) %>%
  mutate(seq_depth = as.character(seq_depth)) %>%
  left_join(raw_lod_case, by = c("deconv_tool", "tumor_type", "seq_depth")) %>%
  mutate(combination_id = paste(dataset, tumor_type, seq_depth, sep = "_")) %>%
  select(combination_id, dataset, tumor_type, seq_depth, deconv_tool, weight, RMSE, SCC, AUC, LoD)

head(raw_metrics_long)
```

## Step 3: rank-then-aggregate (case-based) scheme

Within each case, tools are ranked per metric (higher is better for SCC
and AUC, lower is better for RMSE and LoD). The per-metric ranks are
averaged into a per-case rank, which is then aggregated across cases
with a weighted mean (default) or weighted median.

```{r}
per_case_consensus_rank <- function(metrics_long) {
  # Missing metrics stay NA (not ranked worst); case_rank averages the rest.
  metrics_long %>%
    group_by(combination_id) %>%
    mutate(
      rank_RMSE = rank(RMSE,  ties.method = "average", na.last = "keep"),  # lower RMSE = better
      rank_SCC  = rank(-SCC,  ties.method = "average", na.last = "keep"),  # higher SCC = better
      rank_AUC  = rank(-AUC,  ties.method = "average", na.last = "keep"),  # higher AUC = better
      rank_LoD  = rank(LoD,   ties.method = "average", na.last = "keep")   # lower LoD = better
    ) %>%
    ungroup() %>%
    rowwise() %>%
    mutate(case_rank = mean(c(rank_RMSE, rank_SCC, rank_AUC, rank_LoD), na.rm = TRUE)) %>%
    ungroup()
}

weighted_median <- function(x, w) {
  ord <- order(x)
  x <- x[ord]; w <- w[ord]
  cum_w <- cumsum(w) / sum(w)
  x[which(cum_w >= 0.5)[1]]
}

rank_then_aggregate <- function(metrics_long, agg_fun = mean) {
  ranked <- per_case_consensus_rank(metrics_long)

  ranked %>%
    group_by(deconv_tool) %>%
    summarise(
      Agg_Rank = if (identical(agg_fun, mean)) {
        sum(case_rank * weight) / sum(weight)
      } else if (identical(agg_fun, median)) {
        weighted_median(case_rank, weight)
      } else {
        agg_fun(case_rank, weight)
      },
      .groups = "drop"
    ) %>%
    arrange(Agg_Rank) %>%
    mutate(Rank_rta = rank(Agg_Rank, ties.method = "average"))
}

rta_ranking_mean <- rank_then_aggregate(raw_metrics_long, agg_fun = mean)
rta_ranking_mean
```

Sensitivity check with the median:

```{r}
rta_ranking_median <- rank_then_aggregate(raw_metrics_long, agg_fun = median)
rta_ranking_median
```

## Step 4: compare the two schemes

```{r}
scheme_comparison <- reference_ranking %>%
  select(deconv_tool, Rank_reference) %>%
  inner_join(rta_ranking_mean %>% select(deconv_tool, Rank_rta), by = "deconv_tool") %>%
  inner_join(
    rta_ranking_median %>% select(deconv_tool, Rank_rta_median = Rank_rta),
    by = "deconv_tool"
  ) %>%
  arrange(Rank_reference)

scheme_comparison
```

Kendall's tau between the two schemes:

```{r}
kendall_tau <- cor(
  scheme_comparison$Rank_reference,
  scheme_comparison$Rank_rta,
  method = "kendall"
)

kendall_test <- cor.test(
  scheme_comparison$Rank_reference,
  scheme_comparison$Rank_rta,
  method = "kendall"
)

cat(sprintf("Kendall's tau (aggregate-then-rank vs. rank-then-aggregate, mean): %.3f (p = %.4g)\n",
            kendall_tau, kendall_test$p.value))

kendall_tau_median <- cor(
  scheme_comparison$Rank_reference,
  scheme_comparison$Rank_rta_median,
  method = "kendall"
)
cat(sprintf("Kendall's tau (aggregate-then-rank vs. rank-then-aggregate, median): %.3f\n",
            kendall_tau_median))
```

## Step 5: bootstrap over combinations

Combinations are resampled with replacement (1000 replicates) and both
schemes are recomputed each time. Duplicate draws get a unique suffix
so they count as separate blocks.

This treats every combination as an independent case. Step 6 resamples
at the level of the biological source instead.

```{r}
combination_ids <- unique(combo_scores$combination_id)
stopifnot(setequal(combination_ids, unique(raw_metrics_long$combination_id)))
n_boot <- 1000

aggregate_then_rank_once <- function(scores_df) {
  # Weighted mean of Score across combinations, then rank.
  scores_df %>%
    group_by(deconv_tool) %>%
    summarise(Overall_Score = sum(Score * weight) / sum(weight), .groups = "drop") %>%
    arrange(desc(Overall_Score)) %>%
    mutate(Rank_reference = rank(-Overall_Score, ties.method = "average"))
}

rank_both_schemes <- function(draw_key, i) {
  # Rank both schemes on the same resample. draw_key maps combination_id to
  # a unique block id (draw_id), so duplicate draws count as separate blocks.
  boot_scores <- combo_scores %>%
    inner_join(draw_key, by = "combination_id", relationship = "many-to-many") %>%
    mutate(combination_id = draw_id) %>%
    select(-draw_id)

  boot_metrics <- raw_metrics_long %>%
    inner_join(draw_key, by = "combination_id", relationship = "many-to-many") %>%
    mutate(combination_id = draw_id) %>%
    select(-draw_id)

  ref_rank <- aggregate_then_rank_once(boot_scores) %>%
    transmute(deconv_tool, replicate = i, scheme = "aggregate_then_rank", Rank = Rank_reference)
  rta_rank <- rank_then_aggregate(boot_metrics, agg_fun = mean) %>%
    transmute(deconv_tool, replicate = i, scheme = "rank_then_aggregate", Rank = Rank_rta)

  bind_rows(ref_rank, rta_rank)
}

run_bootstrap_replicate <- function(i) {
  drawn <- sample(combination_ids, length(combination_ids), replace = TRUE)
  draw_key <- tibble(
    combination_id = drawn,
    draw_id = paste0(drawn, "__b", seq_along(drawn))
  )
  rank_both_schemes(draw_key, i)
}

boot_results <- purrr::map_dfr(seq_len(n_boot), run_bootstrap_replicate)
```

```{r}
rank_distribution <- boot_results %>%
  count(scheme, deconv_tool, Rank, name = "freq") %>%
  group_by(scheme, deconv_tool) %>%
  mutate(pct = 100 * freq / sum(freq)) %>%
  ungroup()

pct_rank1 <- boot_results %>%
  filter(Rank == 1) %>%
  count(scheme, deconv_tool, name = "n_rank1") %>%
  mutate(pct_rank1 = 100 * n_rank1 / n_boot) %>%
  select(-n_rank1)

rank1_summary <- reference_ranking %>%
  select(deconv_tool, Rank_reference) %>%
  left_join(pct_rank1 %>% filter(scheme == "aggregate_then_rank") %>%
              select(deconv_tool, pct_rank1_aggregate_then_rank = pct_rank1),
            by = "deconv_tool") %>%
  left_join(pct_rank1 %>% filter(scheme == "rank_then_aggregate") %>%
              select(deconv_tool, pct_rank1_rank_then_aggregate = pct_rank1),
            by = "deconv_tool") %>%
  mutate(across(starts_with("pct_rank1"), ~replace_na(.x, 0))) %>%
  arrange(Rank_reference)

rank1_summary
```

## Step 6: hierarchy-aware resampling (cluster bootstrap + leave-one-out)

The sequencing depths of a tumor type are subsamples of the same
material, so they are nested within one biological source rather than
independent cases. Here the resampling unit is the **biological
source** (tumor type), together with all of its sequencing depths. Each
tumor type occurs in only one dataset.

1.  **Cluster bootstrap** (1000x), stratified by dataset: sources are
    drawn with replacement within each dataset, so every replicate
    keeps 4 WGBS, 3 RRBS and 1 cfRRBS source (NBL is therefore always
    included).
2.  **Leave-one-tumor-type-out (LOTO)**: drop one source at a time.
3.  **Leave-one-dataset-out (LODO)**: drop one dataset at a time
    (dropping cfRRBS is the same as dropping NBL).

This tests whether the top ranking depends on a single source or
dataset. The ranking functions are those of Step 5, so results are
directly comparable. With only 8 sources the cluster bootstrap is
coarse, so the leave-one-out results are the most informative.

```{r}
cluster_key <- combo_scores %>%
  distinct(combination_id, dataset, tumor_type, seq_depth) %>%
  mutate(cluster_id = paste(dataset, tumor_type, sep = "_"))

# Each tumor type belongs to exactly one dataset.
stopifnot(all(count(distinct(cluster_key, dataset, tumor_type), tumor_type)$n == 1))

clusters <- cluster_key %>%
  count(dataset, tumor_type, cluster_id, name = "n_seq_depths")
clusters
```

```{r}
set.seed(42)

# Resample sources within each dataset.
draw_clusters <- function() {
  unlist(lapply(split(clusters$cluster_id, clusters$dataset),
                function(ids) ids[sample.int(length(ids), length(ids), replace = TRUE)]),
         use.names = FALSE)
}

run_cluster_bootstrap_replicate <- function(i) {
  drawn <- draw_clusters()
  # Expand each drawn source into all of its sequencing depths.
  draw_key <- tibble(cluster_id = drawn, draw_idx = seq_along(drawn)) %>%
    inner_join(cluster_key %>% select(cluster_id, combination_id),
               by = "cluster_id", relationship = "many-to-many") %>%
    transmute(combination_id, draw_id = paste0(combination_id, "__c", draw_idx))
  rank_both_schemes(draw_key, i)
}

cluster_boot_results <- purrr::map_dfr(seq_len(n_boot), run_cluster_bootstrap_replicate) %>%
  mutate(resampling = "cluster")
```

Median rank, 95% rank interval, and % ranked 1st for the two
bootstraps. If Step 5 overstated precision, the cluster intervals would
be wider. They can also be narrower: keeping all depths of a source
together averages out depth-level noise, and stratifying fixes the
balance between datasets.

```{r}
all_boot_results <- bind_rows(
  boot_results %>% mutate(resampling = "combination"),
  cluster_boot_results
) %>%
  mutate(resampling = factor(resampling,
                             levels = c("combination", "cluster")))

boot_comparison <- all_boot_results %>%
  group_by(resampling, scheme, deconv_tool) %>%
  summarise(
    median_rank = median(Rank),
    rank_lo_95 = quantile(Rank, 0.025, type = 1),
    rank_hi_95 = quantile(Rank, 0.975, type = 1),
    pct_rank1 = 100 * mean(Rank == 1),
    pct_top3 = 100 * mean(Rank <= 3),
    .groups = "drop"
  ) %>%
  left_join(reference_ranking %>% select(deconv_tool, Rank_reference), by = "deconv_tool") %>%
  arrange(scheme, Rank_reference, resampling)

boot_comparison
```

% ranked 1st per resampling strategy:

```{r}
pct_rank1_by_resampling <- boot_comparison %>%
  select(scheme, deconv_tool, Rank_reference, resampling, pct_rank1) %>%
  pivot_wider(names_from = resampling, values_from = pct_rank1, names_prefix = "pct_rank1_") %>%
  arrange(scheme, Rank_reference)

pct_rank1_by_resampling
```

### Leave-one-out (tumor type and dataset)

For each left-out unit: the top tool, whether it matches the full-data
top tool, Kendall's tau with the full-data ranking, and the largest
rank change of any tool.

```{r}
identity_key <- function(ids) tibble(combination_id = ids, draw_id = ids)

full_data_ranks <- rank_both_schemes(identity_key(combination_ids), "full") %>%
  select(scheme, deconv_tool, Rank_full = Rank)

run_leave_one_out <- function(level) {
  units <- sort(unique(cluster_key[[level]]))
  purrr::map_dfr(units, function(u) {
    kept <- cluster_key$combination_id[cluster_key[[level]] != u]
    rank_both_schemes(identity_key(kept), u) %>%
      rename(left_out = replicate) %>%
      mutate(level = level)
  })
}

loo_results <- bind_rows(
  run_leave_one_out("tumor_type"),
  run_leave_one_out("dataset")
) %>%
  left_join(full_data_ranks, by = c("scheme", "deconv_tool")) %>%
  mutate(level = dplyr::recode(level,
                               tumor_type = "Leave-one-tumor-type-out",
                               dataset = "Leave-one-dataset-out"))

full_top_tool <- full_data_ranks %>%
  filter(Rank_full == min(Rank_full), .by = scheme) %>%
  summarise(full_top = paste(sort(deconv_tool), collapse = " / "), .by = scheme)

loo_summary <- loo_results %>%
  group_by(level, left_out, scheme) %>%
  summarise(
    top_tool = paste(sort(deconv_tool[Rank == min(Rank)]), collapse = " / "),
    kendall_tau_vs_full = cor(Rank, Rank_full, method = "kendall"),
    max_abs_rank_shift = max(abs(Rank - Rank_full)),
    .groups = "drop"
  ) %>%
  left_join(full_top_tool, by = "scheme") %>%
  mutate(top_tool_unchanged = top_tool == full_top) %>%
  arrange(scheme, level, left_out)

full_top_tool
loo_summary
```

```{r}
loo_overview <- loo_summary %>%
  group_by(scheme, level) %>%
  summarise(
    n_left_out_units = n(),
    n_top_tool_unchanged = sum(top_tool_unchanged),
    min_kendall_tau = min(kendall_tau_vs_full),
    max_abs_rank_shift = max(max_abs_rank_shift),
    .groups = "drop"
  )
loo_overview
```

## Step 7: sensitivity to the dataset weights

06 weights the datasets unequally (WGBS 0.45, RRBS 0.225, cfRRBS 0.225,
scalability 0.1). We recompute both rankings with equal dataset
weights, applied as in 06: combinations are first averaged within each
dataset, then the datasets are combined with a weighted mean. Two
equal-weight variants are tested: equal weights for the three
benchmark datasets (scalability kept at 0.1), and equal weights for all
four components.

For rank-then-aggregate, each dataset contributes the mean per-case rank
of its combinations, and scalability contributes the rank of 06's
scalability score. Under the original weights this version therefore
differs slightly from Step 3, which weights individual combinations and
leaves out scalability.

```{r}
weight_schemes <- tribble(
  ~weighting,                     ~wgbs, ~rrbs, ~cfrrbs, ~scalability,
  "Original (06)",                0.45,  0.225, 0.225,   0.1,
  "Equal datasets",               0.3,   0.3,   0.3,     0.1,
  "Equal datasets + scalability", 0.25,  0.25,  0.25,    0.25
) %>%
  pivot_longer(-weighting, names_to = "component", values_to = "weight") %>%
  mutate(weighting = factor(weighting, levels = unique(weighting)))

# Per-dataset scores from 06 (higher = better).
atr_components <- bind_rows(
  env_06$overall_wgbs_score   %>% mutate(component = "wgbs"),
  env_06$overall_rrbs_score   %>% mutate(component = "rrbs"),
  env_06$overall_cfrrbs_score %>% mutate(component = "cfrrbs"),
  env_06$aggr_scalab_score    %>% mutate(component = "scalability")
) %>%
  select(deconv_tool, component, value = Score)

# Per-dataset mean case rank, and scalability rank (lower = better).
rta_components <- per_case_consensus_rank(raw_metrics_long) %>%
  group_by(deconv_tool, component = dataset) %>%
  summarise(value = mean(case_rank), .groups = "drop") %>%
  bind_rows(env_06$aggr_scalab_score %>%
              transmute(deconv_tool, component = "scalability",
                        value = rank(-Score, ties.method = "average")))

stopifnot(all(count(atr_components, component)$n == n_tools),
          all(count(rta_components, component)$n == n_tools))

rank_under_weights <- function(components, higher_is_better) {
  components %>%
    inner_join(weight_schemes, by = "component", relationship = "many-to-many") %>%
    group_by(weighting, deconv_tool) %>%
    summarise(Agg = sum(value * weight) / sum(weight), .groups = "drop") %>%
    group_by(weighting) %>%
    mutate(Rank = rank(if (higher_is_better) -Agg else Agg, ties.method = "average")) %>%
    ungroup()
}

weighting_results <- bind_rows(
  rank_under_weights(atr_components, higher_is_better = TRUE) %>%
    mutate(scheme = "aggregate_then_rank"),
  rank_under_weights(rta_components, higher_is_better = FALSE) %>%
    mutate(scheme = "rank_then_aggregate")
)

# The original weights must reproduce 06's overall score exactly.
check_06 <- weighting_results %>%
  filter(scheme == "aggregate_then_rank", weighting == "Original (06)") %>%
  inner_join(env_06$overall_score, by = "deconv_tool")
stopifnot(isTRUE(all.equal(check_06$Agg, check_06$Overall_Score)))

weighting_ranks_wide <- weighting_results %>%
  select(scheme, weighting, deconv_tool, Rank) %>%
  pivot_wider(names_from = weighting, values_from = Rank) %>%
  left_join(reference_ranking %>% select(deconv_tool, Rank_reference), by = "deconv_tool") %>%
  arrange(scheme, Rank_reference) %>%
  select(-Rank_reference)

weighting_ranks_wide
```

Top tool and Kendall's tau with the original weighting, per scheme:

```{r}
weighting_summary <- weighting_results %>%
  inner_join(weighting_results %>%
               filter(weighting == "Original (06)") %>%
               select(scheme, deconv_tool, Rank_original = Rank),
             by = c("scheme", "deconv_tool")) %>%
  group_by(scheme, weighting) %>%
  summarise(
    top_tool = paste(sort(deconv_tool[Rank == min(Rank)]), collapse = " / "),
    kendall_tau_vs_original = cor(Rank, Rank_original, method = "kendall"),
    max_abs_rank_shift = max(abs(Rank - Rank_original)),
    .groups = "drop"
  )

weighting_summary
```

## Step 8: Friedman test + Nemenyi post-hoc

Friedman test (tools = treatments, combinations = blocks) with Nemenyi
post-hoc, run separately on `Score` and on the per-case rank.

```{r}
build_block_matrix <- function(long_df, value_col) {
  block_matrix_df <- long_df %>%
    select(combination_id, deconv_tool, all_of(value_col)) %>%
    pivot_wider(names_from = deconv_tool, values_from = all_of(value_col))

  complete_rows <- complete.cases(block_matrix_df %>% select(-combination_id))
  block_matrix_complete <- block_matrix_df[complete_rows, ]
  block_matrix <- as.matrix(block_matrix_complete %>% select(-combination_id))
  rownames(block_matrix) <- block_matrix_complete$combination_id

  list(
    matrix = block_matrix,
    n_total = nrow(block_matrix_df),
    n_complete = sum(complete_rows)
  )
}

run_friedman_nemenyi <- function(block_info, label) {
  cat(sprintf("[%s] Combinations available: %d. Complete blocks (all %d tools present): %d.\n",
              label, block_info$n_total, n_tools, block_info$n_complete))

  if (block_info$n_complete < 2) {
    cat(sprintf("[%s] Fewer than 2 complete blocks available -- Friedman/Nemenyi skipped.\n", label))
    return(invisible(NULL))
  }

  friedman_result <- friedman.test(block_info$matrix)
  nemenyi_result <- PMCMRplus::frdAllPairsNemenyiTest(block_info$matrix)
  print(friedman_result)
  print(nemenyi_result)
  invisible(list(friedman = friedman_result, nemenyi = nemenyi_result))
}

# Nemenyi p-value matrix (lower triangle) -> symmetric long table.
nemenyi_long <- function(nemenyi_result) {
  lower <- as.data.frame(as.table(nemenyi_result$p.value), stringsAsFactors = FALSE) %>%
    filter(!is.na(Freq))
  bind_rows(
    tibble(tool_1 = lower$Var1, tool_2 = lower$Var2, p_adj = lower$Freq),
    tibble(tool_1 = lower$Var2, tool_2 = lower$Var1, p_adj = lower$Freq)
  )
}

output_dir <- "../plots/11_ranking_robustness"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
```

### 8a. Aggregate-then-rank (reference, min-max-scaled `Score`)

```{r}
block_info_score <- build_block_matrix(combo_scores, "Score")
n_total_combinations <- block_info_score$n_total
n_complete_combinations <- block_info_score$n_complete
fn_score <- run_friedman_nemenyi(block_info_score, "aggregate-then-rank (Score)")
```

### 8b. Rank-then-aggregate (case-based, raw per-case consensus rank)

```{r}
case_ranks_wide <- per_case_consensus_rank(raw_metrics_long) %>%
  select(combination_id, deconv_tool, case_rank)

block_info_case_rank <- build_block_matrix(case_ranks_wide, "case_rank")
fn_case_rank <- run_friedman_nemenyi(block_info_case_rank, "rank-then-aggregate (case_rank)")
```

### 8c. Sensitivity: tumor types as blocks

The 36 combination blocks are not independent: the sequencing depths of
a tumor type come from the same material (see Step 6). As a sensitivity
check, we average each tool's value over the sequencing depths of each
tumor type and rerun both tests with the 8 tumor types as blocks. This
respects the hierarchy but has less power.

```{r}
to_tumor_blocks <- function(long_df, value_col) {
  long_df %>%
    group_by(combination_id = paste(dataset, tumor_type, sep = "_"), deconv_tool) %>%
    summarise(value = mean(.data[[value_col]]), .groups = "drop")
}

block_info_score_tt <- build_block_matrix(to_tumor_blocks(combo_scores, "Score"), "value")
fn_score_tt <- run_friedman_nemenyi(block_info_score_tt, "aggregate-then-rank (Score), tumor-type blocks")

block_info_case_rank_tt <- build_block_matrix(
  to_tumor_blocks(per_case_consensus_rank(raw_metrics_long), "case_rank"), "value")
fn_case_rank_tt <- run_friedman_nemenyi(block_info_case_rank_tt, "rank-then-aggregate (case_rank), tumor-type blocks")
```

### 8d. Summary tables

Friedman results, and Nemenyi comparisons of the top-ranked tool
against every other tool, for both schemes and both blockings. The full
Nemenyi matrices are exported as CSV.

```{r}
fn_all <- list(
  list(scheme = "aggregate_then_rank", blocks = "combinations (36)", res = fn_score),
  list(scheme = "rank_then_aggregate", blocks = "combinations (36)", res = fn_case_rank),
  list(scheme = "aggregate_then_rank", blocks = "tumor types (8)",   res = fn_score_tt),
  list(scheme = "rank_then_aggregate", blocks = "tumor types (8)",   res = fn_case_rank_tt)
)

friedman_table <- purrr::map_dfr(fn_all, function(x) tibble(
  scheme = x$scheme,
  blocks = x$blocks,
  chi_squared = unname(x$res$friedman$statistic),
  df = unname(x$res$friedman$parameter),
  p_value = x$res$friedman$p.value
))
friedman_table

nemenyi_all <- purrr::map_dfr(fn_all, function(x) {
  nemenyi_long(x$res$nemenyi) %>% mutate(scheme = x$scheme, blocks = x$blocks)
})

top_tool <- reference_ranking$deconv_tool[which.min(reference_ranking$Rank_reference)]

top_tool_vs_others <- nemenyi_all %>%
  filter(tool_1 == top_tool) %>%
  transmute(comparison = paste(tool_1, "vs.", tool_2),
            tool_2,
            column = paste(scheme, blocks, sep = " | "),
            p_adj) %>%
  pivot_wider(names_from = column, values_from = p_adj) %>%
  left_join(reference_ranking %>% select(tool_2 = deconv_tool, Rank_reference), by = "tool_2") %>%
  arrange(Rank_reference) %>%
  select(-tool_2, -Rank_reference)

top_tool_vs_others %>%
  mutate(across(-comparison, ~signif(.x, 2)))

write.csv(friedman_table, file.path(output_dir, "friedman_tests.csv"), row.names = FALSE)
write.csv(top_tool_vs_others, file.path(output_dir, "nemenyi_top_tool_vs_others.csv"), row.names = FALSE)
write.csv(nemenyi_all %>% select(scheme, blocks, tool_1, tool_2, p_adj),
          file.path(output_dir, "nemenyi_all_pairs.csv"), row.names = FALSE)
```

## Step 9: plots

### Bootstrap rank-distribution blob plot

Dot size = % of replicates in which a tool got a given rank (style of
Fig 2B).

```{r fig.width=9, fig.height=6}
tool_order <- reference_ranking %>% arrange(Rank_reference) %>% pull(deconv_tool)

rank_distribution_plot_df <- rank_distribution %>%
  mutate(
    deconv_tool = factor(deconv_tool, levels = rev(tool_order)),
    scheme = dplyr::recode(scheme,
                     aggregate_then_rank = "Aggregate-then-rank (reference)",
                     rank_then_aggregate = "Rank-then-aggregate (mean rank)")
  )

p_blob <- ggplot(rank_distribution_plot_df, aes(x = Rank, y = deconv_tool, size = pct)) +
  geom_point(alpha = 0.75, color = "#2c7fb8") +
  scale_size_continuous(name = "% of\nresamples", range = c(0.5, 12)) +
  scale_x_continuous(breaks = seq_len(n_tools)) +
  facet_wrap(~scheme) +
  labs(
    title = "Bootstrap rank-distribution across combinations (1000 resamples)",
    x = "Rank (1 = best)",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())

p_blob
```

```{r}
ggsave(file.path(output_dir, "bootstrap_rank_distribution_blobplot.svg"),
       p_blob, width = 12, height = 6)
```

### Scheme-comparison plot

Points on the diagonal = same rank in both schemes.

```{r fig.width=7, fig.height=6}
p_scheme <- ggplot(scheme_comparison, aes(x = Rank_reference, y = Rank_rta, label = deconv_tool)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(size = 3, color = "#e34a33") +
  ggrepel::geom_text_repel(size = 3.5, seed = 42) +
  scale_x_continuous(breaks = seq_len(n_tools)) +
  scale_y_continuous(breaks = seq_len(n_tools)) +
  labs(
    title = sprintf("Aggregate-then-rank vs. rank-then-aggregate (Kendall's tau = %.3f)", kendall_tau),
    x = "Rank -- aggregate-then-rank (reference)",
    y = "Rank -- rank-then-aggregate (mean rank)"
  ) +
  coord_equal() +
  theme_minimal(base_size = 12)

p_scheme
```

```{r}
ggsave(file.path(output_dir, "scheme_comparison.svg"), p_scheme, width = 8, height = 6)
```

### Cluster bootstrap rank-distribution blob plot

Same as above, for the cluster bootstrap of Step 6.

```{r fig.width=9, fig.height=6}
scheme_labels <- c(
  aggregate_then_rank = "Aggregate-then-rank (reference)",
  rank_then_aggregate = "Rank-then-aggregate (mean rank)"
)

cluster_rank_distribution_plot_df <- cluster_boot_results %>%
  count(scheme, deconv_tool, Rank, name = "freq") %>%
  group_by(scheme, deconv_tool) %>%
  mutate(pct = 100 * freq / sum(freq)) %>%
  ungroup() %>%
  mutate(
    deconv_tool = factor(deconv_tool, levels = rev(tool_order)),
    scheme = scheme_labels[scheme]
  )

p_blob_cluster <- ggplot(cluster_rank_distribution_plot_df, aes(x = Rank, y = deconv_tool, size = pct)) +
  geom_point(alpha = 0.75, color = "#2c7fb8") +
  scale_size_continuous(name = "% of\nresamples", range = c(0.5, 12)) +
  scale_x_continuous(breaks = seq_len(n_tools)) +
  facet_wrap(~scheme) +
  labs(
    title = sprintf("Bootstrap rank-distribution across biological sources (%d resamples)", n_boot),
    x = "Rank (1 = best)",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())

p_blob_cluster
```

```{r}
ggsave(file.path(output_dir, "cluster_bootstrap_rank_distribution_blobplot.svg"),
       p_blob_cluster, width = 12, height = 6)
```

### Dataset-weight sensitivity bump chart

Rank of each tool under each weighting (Step 7).

```{r fig.width=11, fig.height=6}
weighting_levels <- levels(weighting_results$weighting)
weighting_axis_labels <- c(
  "Original (06)" = "Original\n0.45/0.225/0.225/0.1",
  "Equal datasets" = "Equal datasets\n0.3/0.3/0.3/0.1",
  "Equal datasets + scalability" = "Equal incl. scalability\n0.25/0.25/0.25/0.25"
)

weighting_plot_df <- weighting_results %>%
  mutate(
    x = as.integer(weighting),
    scheme = scheme_labels[scheme],
    highlight = deconv_tool == "CelFiE"
  )

p_weighting <- ggplot(weighting_plot_df, aes(x = x, y = Rank, group = deconv_tool)) +
  geom_line(aes(color = highlight, linewidth = highlight)) +
  geom_point(aes(color = highlight), size = 2.5) +
  geom_text(data = filter(weighting_plot_df, x == 1),
            aes(label = deconv_tool), hjust = 1, nudge_x = -0.08, size = 3.3, color = "grey20") +
  geom_text(data = filter(weighting_plot_df, x == length(weighting_levels)),
            aes(label = deconv_tool), hjust = 0, nudge_x = 0.08, size = 3.3, color = "grey20") +
  scale_color_manual(values = c(`FALSE` = "grey65", `TRUE` = "#2c7fb8"), guide = "none") +
  scale_linewidth_manual(values = c(`FALSE` = 0.6, `TRUE` = 1.4), guide = "none") +
  scale_x_continuous(breaks = seq_along(weighting_levels),
                     labels = weighting_axis_labels[weighting_levels],
                     expand = expansion(add = c(1, 1))) +
  scale_y_reverse(breaks = seq_len(n_tools)) +
  facet_wrap(~scheme) +
  labs(
    title = "Sensitivity of the ranking to the dataset weights",
    subtitle = "Weights: WGBS / RRBS / cfRRBS / scalability",
    x = NULL,
    y = "Rank (1 = best)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank())

p_weighting
```

```{r}
ggsave(file.path(output_dir, "dataset_weight_sensitivity.svg"),
       p_weighting, width = 13, height = 6)
```

## Summary

```{r}
cat(sprintf(
  "Reference ranking (06, aggregate-then-rank) vs. rank-then-aggregate (mean): Kendall's tau = %.3f\n",
  kendall_tau
))
cat(sprintf(
  "Reference ranking vs. rank-then-aggregate (median, sensitivity check): Kendall's tau = %.3f\n",
  kendall_tau_median
))
cat(sprintf(
  "Bootstrap: %d replicates over %d combinations (dataset x tumor_type x seq_depth).\n",
  n_boot, length(combination_ids)
))
cat(sprintf(
  "Cluster bootstrap: %d replicates over %d biological sources (dataset x tumor_type), stratified by dataset.\n",
  n_boot, nrow(clusters)
))
for (s in names(scheme_labels)) {
  top <- full_top_tool$full_top[full_top_tool$scheme == s]
  pct <- pct_rank1_by_resampling %>% filter(scheme == s, deconv_tool == top)
  lo <- loo_overview %>% filter(scheme == s)
  cat(sprintf(
    "[%s] full-data top tool: %s | %% ranked 1st: combination = %.1f, cluster = %.1f | top tool unchanged in %d/%d LOTO and %d/%d LODO\n",
    scheme_labels[[s]], top,
    pct$pct_rank1_combination, pct$pct_rank1_cluster,
    lo$n_top_tool_unchanged[lo$level == "Leave-one-tumor-type-out"],
    lo$n_left_out_units[lo$level == "Leave-one-tumor-type-out"],
    lo$n_top_tool_unchanged[lo$level == "Leave-one-dataset-out"],
    lo$n_left_out_units[lo$level == "Leave-one-dataset-out"]
  ))
}
weighting_summary %>%
  filter(weighting != "Original (06)") %>%
  with(cat(sprintf("[%s | %s weights] top tool: %s | Kendall's tau vs. original weights = %.3f\n",
                   scheme_labels[scheme], weighting, top_tool, kendall_tau_vs_original), sep = ""))
cat(sprintf(
  "Friedman/Nemenyi: %d / %d combinations used as complete blocks.\n",
  n_complete_combinations, n_total_combinations
))
```
